\chapter*{Changelog}
\addcontentsline{toc}{chapter}{Changelog}

This changelog records which chapters and sections were substantially revised
in each version of the document. Minor corrections (typos, formatting) are not
listed.

Each entry follows the format: \textbf{What} changed, \textit{why} it changed
(new evidence, error correction, reviewer feedback, etc.), and which
\textbf{key sources} motivated the change where applicable.

\section*{Version 7 (2026-03-26 --)}

\subsection*{New Content}

\begin{itemize}

  \item \textbf{Vascular pathology integration (2026-04-10): new content across Chapters 6, 7, 10, 14, 18}
    based on systematic research into endothelial glycocalyx, microvascular dysfunction, RBC pathology,
    respiratory physiology, and endothelial senescence in ME/CFS and Long COVID.
    \begin{itemize}

      \item \textbf{Chapter 7 (Immune Dysfunction) --- new section: Virus-Induced Endothelial Senescence}:
        Added full mechanistic section on SASP-producing senescent endothelial cells, NK/T cell clearance
        failure creating a self-sustaining bidirectional loop, tissue-specific manifestations, and senolytic
        therapeutic implications. Formal \texttt{hypothesis} environment added with falsifiable predictions.
        Motivated by: Nunes et al.\ Cell Death \& Disease 2026.
        \cite{Nunes2026endothelialSenescence}

      \item \textbf{Chapter 10 (Cardiovascular) --- endothelial senescence + failed angiogenesis}:
        New subsection on endothelial senescence mechanism (cross-referencing Chapter~7) and Flaskamp et al.\
        finding that ME/CFS serum specifically loses angiogenic compensation present in post-COVID without ME/CFS.
        Motivated by: Nunes et al.\ 2026; Flaskamp et al.\ Cells 2022.
        \cite{Nunes2026endothelialSenescence, Flaskamp2022endothelial}

      \item \textbf{Chapter 10 --- coagulation and fibrin sections expanded}:
        Platelet hyperactivation and microclot data quantified with exact figures from Nunes et al.\ 2022.
        Fibrin section expanded with microclot proteomics (Kruger et al.\ 2022: 27 immunoglobulin variants,
        VWF 2.6$\times$, PF4 3.5$\times$). Proteomic signature added (Nunes et al.\ 2024: THBS1, PF4, P-selectin elevated,
        Protein S reduced). Cardiovascular haematology review (Nunes et al.\ 2023) added.
        Motivated by: Nunes et al.\ Pharmaceuticals 2022; Kruger et al.\ Cardiovasc Diabetol 2022;
        Nunes et al.\ Cardiovasc Diabetol 2024; Nunes et al.\ Blood Reviews 2023.
        \cite{NunesPlatelets2022mecfs, KrugerMicroclot2022longcovid, NunesProteomics2024mecfs, NunesBloodReviews2023mecfs}

      \item \textbf{Chapter 10 --- RBC deformability section substantially expanded}:
        Added Stanford microfluidic data (Guo et al.\ 2025: impaired hypoxia-sensing in ME/CFS RBCs at
        capillary level), persistent RBC morphological abnormalities in Long COVID at 42 weeks (Grau et al.\
        2024: 20.8\% abnormal vs.\ 2.9\% controls; r=0.51 fatigue correlation), and modelling of oxygen
        delivery deficit magnitude (Rogers et al.\ 2024: 135\% cardiac output increase needed to compensate).
        Motivated by: Guo et al.\ Blood RCI 2025; Grau et al.\ Biology 2024; Rogers et al.\ Front Physiol 2024.
        \cite{GuoRBC2025microfluidic, GrauRBC2024longcovid, RogersRBC2024covidoxygen}

      \item \textbf{Chapter 10 --- CBF mechanisms expanded with orthostatic hypocapnia amplifier}:
        Added van Campen et al.\ 2023 quantitative data (PETCO$_2$ 26--30 mmHg during tilt vs.\ 36 mmHg controls;
        n=535) as a documented amplifier mechanism in the cerebral hypoperfusion mechanisms list. CO$_2$ reactivity
        intact but hypocapnia large, explaining why CBF falls exceed systemic haemodynamic changes.
        Motivated by: van Campen et al.\ Physiol Rep 2023.
        \cite{VanCampen2023co2cbf}

      \item \textbf{Chapter 10 --- two new formal hypotheses added}:
        (1) Vascular pathology as the primary initiating event in ME/CFS (certainty 0.45; falsifiable
        predictions listed). (2) Failed angiogenic compensation as the ME/CFS transition mechanism
        (certainty 0.30; falsifiable predictions listed).
        Motivated by: Scheibenbogen \& Wirth J Cachexia Sarcopenia Muscle 2024; Flaskamp et al.\ 2022.
        \cite{ScheibenbogenWirth2024sarcopenia, Flaskamp2022endothelial}

      \item \textbf{Chapter 6 (Energy Metabolism) --- vascular-primary causal model added}:
        Expanded the ``Substrate Delivery Failure'' section with the Scheibenbogen-Wirth temporal model
        (early PCS vascular-only, later ME/CFS mitochondrial damage), three independent lines of supporting
        evidence (structural, cell-autonomous RBC, interventional), and the AIMM framework implications.
        Motivated by: Scheibenbogen \& Wirth 2024; Guo et al.\ 2025; Joseph et al.\ Chest 2022.
        \cite{ScheibenbogenWirth2024sarcopenia, GuoRBC2025microfluidic, Joseph2022pyridostigmine}

      \item \textbf{Chapter 18 (Emerging Therapies) --- two new sections added}:
        (1) Inspiratory Muscle Training (IMT): pilot RCT evidence (Edgell et al.\ 2025, n=12 ME/CFS,
        safe PEM-avoiding protocol, improved vascular function and COMPASS-31 autonomic score).
        (2) Carotid Body Sensitisation and P2X3 Antagonism: Long COVID evidence (El-Medany et al.\ 2024,
        HVR 145\% elevated), ACE2 mechanism, gefapixant as therapeutic target; speculation environment
        with falsifiable predictions for ME/CFS.
        Motivated by: Edgell et al.\ Respir Physiol Neurobiol 2025; El-Medany et al.\ Commun Med 2024.
        \cite{Edgell2025imt, ElMedany2024carotid}

      \item \textbf{Chapter 14k (new subfile: ch14k-vascular-brainstorm.typ)}:
        Creative brainstorm section with: four novel mechanistic hypotheses (glycocalyx as pacing device,
        orthostatic hypocapnia--microclot loop, SA-EV propagation of senescence, RBC phosphatidylserine
        as microclot nucleation signal); three novel treatment proposals (senolytic D+Q pilot trial,
        glycocalyx prevention trial, P2X3+volume combination); formal ODE sketch of vascular-mitochondrial
        bistable loop (V, M, Na, Ca state variables); research priority matrix.
        Motivated by: synthesis of 2022--2026 vascular literature.

    \end{itemize}

  \item \textbf{Eight new treatment sections added to Chapter~18 (Emerging Therapies)}
    based on coverage gap analysis against the Spooniverse treatment database (75 entries):
    \begin{itemize}
      \item \textbf{Rapamycin (Sirolimus) --- upgraded}: Upgraded from a 4-line placeholder to a
        full subsection with mechanism, clinical evidence, and safety warning.
        Motivated by: Ruan et al.\ J Transl Med 2025 (PMID 41121328, $n=86$ ME/CFS, 74.3\% response
        in fatigue and PEM; first mechanistic trial linking autophagy impairment to PEM).
        \cite{Ruan2025rapamycin, BarTana2025mTOR}
      \item \textbf{Therapeutic Plasma Exchange --- updated}: Added negative sham-controlled RCT result
        (España-Cueto et al.\ Nat Commun 2025, $n=50$, no improvement vs sham) as important
        counterpoint to prior anecdotal reports.
      \item \textbf{Stellate Ganglion Block (SGB)}: New section. Sympathetic chain injection for
        autonomic reboot in Long COVID and ME/CFS. Three retrospective series cited (Liu \& Duricka
        2022/2024, Pearson et al.\ 2023; up to 86\% symptom reduction); no controlled trial exists.
        \cite{Liu2022SGB, Pearson2023SGB, Duricka2024SGB}
      \item \textbf{Sulodexide}: New section. Glycosaminoglycan restoring the endothelial glycocalyx.
        TUN-EndCOV RCT (Charfeddine et al.\ 2022, $n=290$): significant improvement in endothelial
        function, chest pain, and palpitations vs untreated controls. \cite{Charfeddine2022sulodexide}
      \item \textbf{Maraviroc + Pravastatin}: New section. CCR5 antagonist + statin targeting
        non-classical monocyte activation and viral persistence. Patterson et al.\ Front Med 2023
        ($n=18$ case series); Phase III RCT ongoing. \cite{Patterson2023maraviroc}
      \item \textbf{Pembrolizumab}: New section. Anti-PD-1 immune checkpoint inhibitor targeting
        T-cell exhaustion. NINDS Phase 1 trial (PD1-PASC I, $n=15$) underway; serious irAE risk
        prominently flagged.
      \item \textbf{Psilocybin}: New section. 5-HT2A agonist; anecdotal evidence only (Chopra et al.\
        2024 case report); PEM risk during psychedelic sessions highlighted.
        \cite{Chopra2024psilocybin}
      \item \textbf{Guanfacine + NAC}: New section. $\alpha_2$A-agonist + antioxidant combination
        targeting prefrontal noradrenergic deficits and brain fog. No published trial; mechanistically
        grounded.
      \item \textbf{Nicotine Transdermal Patch}: New section. nAChR activation via the cholinergic
        anti-inflammatory pathway. Leitzke 2023 case series; cardiovascular caution flagged.
    \end{itemize}

  \item \textbf{Chapter~18 adversarial review pass --- pharmacology, logic, and rigor corrections}
    (Typst source updated; all 31 items from a comprehensive review pass):
    Motivated by: full chapter review (xref, citation, production-readiness, pharmacology, scientific rigor,
    adversarial persona, sophist, devil's advocate, style, safety audits; build verification).
    \begin{itemize}
      \item \textbf{Rapamycin}: Added ITT denominator ($\sim$34\% vs 74.3\% completers); replaced
        mechanism bullet list with \texttt{speculation} environment noting circular use of Ruan~2025
        as both mechanism evidence and treatment evidence; corrected bioavailability data; stated off-label status.
      \item \textbf{Stellate Ganglion Block}: Replaced ``HRV studies consistently show'' (unsupported
        aggregative claim) with qualified prose acknowledging study heterogeneity; replaced
        mechanism chain prose with \texttt{speculation} environment critiquing the ``reset'' hysteresis
        assumption; added vocal cord paresis to procedural risk list; quantified placebo rate (30--50\%).
      \item \textbf{Sulodexide}: Converted glycocalyx cascade bullets to \texttt{speculation} environment
        noting that trial outcomes may reflect anticoagulant/anti-inflammatory activity rather than
        glycocalyx-specific restoration; added therapeutic anticoagulation contraindication.
      \item \textbf{Maraviroc + Pravastatin}: Added LFT monitoring protocol (baseline ALT/AST/bilirubin;
        discontinuation criteria for hepatotoxicity); added pravastatin P-glycoprotein interaction with
        maraviroc; statin myopathy risk in ME/CFS muscle context; pravastatin NF-$\kappa$B confound in
        Patterson case series; off-label indication stated.
      \item \textbf{Pembrolizumab}: Added counter-argument that PD-1 exhaustion may be protective rather
        than pathological in ME/CFS (uncontrolled autoreactivity risk); added autoimmune comorbidity
        elevated irAE risk; clarified fatality risk framing.
      \item \textbf{Psilocybin}: Corrected ``DMN hyperactivity'' from established fact to contested finding;
        added that 5-HT2A agonism does not directly modulate kynurenine pathway enzymes; upgraded
        SSRI interaction from ``theoretical'' to ``clinically recognised serotonin syndrome risk'';
        added psychosis/bipolar contraindication.
      \item \textbf{Guanfacine + NAC}: Rewrote mechanism (prefrontal $\alpha_2$A vs brainstem vasomotor
        centre distinction); removed inline community doses from body text; moved dose information
        to warning box with explicit ``not validated by trial'' label; added slow taper guidance
        ($\sim$25\%/1--2 weeks); noted pravastatin-monotherapy confound.
      \item \textbf{Nicotine patch}: Corrected pharmacological error --- CYP1A2 induction is produced
        by polycyclic aromatic hydrocarbons in tobacco smoke, not by nicotine itself; transdermal
        nicotine does not meaningfully induce CYP1A2. Reframed as de-induction concern for
        smoking-cessation transition.
      \item \textbf{Tocilizumab}: Replaced ``consistently elevated'' with Hornig et al.\ heterogeneity
        (duration-dependent); added RA flare confound in the case report; added black-box warnings
        (TB reactivation, GI perforation); noted CRP suppression and IV-only administration.
      \item \textbf{Efgartigimod}: Corrected ``all circulating immunoglobulins'' to accurate IgG-selective
        description; renamed limitation box; added explicit falsifiability condition.
      \item \textbf{Temelimab}: Renamed limitation box; added epiphenomenon interpretation (HERV-W as
        non-specific immune activation marker) with equal weight to the positive interpretations;
        added what would substantially weaken the causal hypothesis.
      \item \textbf{Leronlimab}: Downgraded evidence grade from C to D (RCT negative primary endpoint;
        post-hoc subgroup signal has high false-positive risk); added explicit false-positive
        risk discussion for post-hoc subgroup analysis.
      \item \textbf{GLP-1 agonists}: Added black-box warnings (medullary thyroid carcinoma/MEN2
        contraindication, pancreatitis risk); added gastroparesis worsening risk in dysautonomic
        ME/CFS; added narrow-therapeutic-window drug absorption delay; removed unverified 89\%
        community-survey statistic; wrapped mechanisms in \texttt{speculation} environment
        (certainty~0.20); corrected ``by mechanistic extension'' non-sequitur --- positive Long
        COVID trial result would not license GLP-1 use in non-Long-COVID ME/CFS without a
        separate trial.
      \item \textbf{LOLA}: Wrapped ammonia hypothesis in \texttt{speculation} environment (certainty~0.10);
        added hepatic urea cycle spare-capacity counter-argument; added renal impairment
        contraindication; evidence grade updated.
      \item \textbf{Pemivibart}: Replaced ``will be updated'' maintenance note with substantive
        interpretive point (spike persistence as immune-failure consequence vs.\ causal driver).
      \item \textbf{Bezisterim}: Corrected ERK/NF-$\kappa$B MOA inversion --- NE3107 preserves
        ERK-mediated neurotrophic signalling while inhibiting the inflammatory NF-$\kappa$B
        branch; selectivity claim attributed to developer studies pending independent validation;
        replaced false-dichotomy negative-result framing with multi-interpretation analysis.
      \item \textbf{All 17 new sections}: Heading levels changed from \texttt{=} (chapter-level) to
        \texttt{==} (section within Chapter~18) to ensure they render within the experimental
        therapies framing rather than as apparent independent chapters.
    \end{itemize}

  \item \textbf{Nine further treatment sections added to Chapter~18 (Emerging Therapies)} --- completing the full Spooniverse coverage gap analysis (all 18 identified gaps now addressed):
    \begin{itemize}
      \item \textbf{Tocilizumab}: New section. IL-6R blockade; single case report (Visvabharathy 2022);
        mechanism plausible in IL-6-elevated subgroup; no RCT. \cite{Visvabharathy2022tocilizumab}
      \item \textbf{Efgartigimod}: New section. FcRn blocker for IgG reduction. Phase~2 RCT
        (NCT05633407) negative on primary endpoint; peer-reviewed results pending.
        Limitation box explains why a negative FcRn trial does not disprove autoantibody hypothesis.
      \item \textbf{Temelimab}: New section. Anti-HERV-W-Env monoclonal antibody. Phase~2 RCT
        ($n=203$) negative; HERV-W biology review from Charvet et al.\ 2023 and Giménez-Orenga et al.\
        2022 retained as valid scientific foundation for biomarker-enriched future trials.
        \cite{Charvet2023HERV, GimenezOrenga2022HERV}
      \item \textbf{Leronlimab}: New section. CCR5-blocking antibody. RCT ($n=54$) negative on
        primary endpoint; post-hoc responder subgroup signal noted. Contextualised against maraviroc
        Phase~III. \cite{Gaylis2022leronlimab}
      \item \textbf{GLP-1 Receptor Agonists (Semaglutide, Tirzepatide)}: New section.
        Anti-inflammatory, mast cell modulation, mitochondrial mechanisms. Tirzepatide Long COVID
        RCT ($n\approx1000$) enrolled December 2025; results pending. Limitation box flags absence of
        peer-reviewed controlled data.
      \item \textbf{LOLA (Ornithine-Aspartate)}: New section. Ammonia detoxification; computational
        modelling only (Xiao et al.\ 2025); no clinical trial.
      \item \textbf{Pemivibart}: New section. Anti-spike mAb; UCSF RCT ($n=36$, results unpublished).
        Section frames what a positive result would establish for the spike-persistence hypothesis.
      \item \textbf{Sipavibart}: New section. Anti-spike mAb for COVID prophylaxis in
        immunocompromised; \emph{not} a Long COVID treatment. Included to clarify misclassification
        in some secondary databases.
      \item \textbf{Bezisterim (NE3107)}: New section. ERK/NF-$\kappa$B neuroinflammation inhibitor.
        ADDRESS-LC Phase~2/3 RCT (NCT06847191, \$13.1M DoD CDMRP) enrolling; no efficacy data yet.
        Section frames what a positive or negative result would mean for the neuroinflammatory
        hypothesis.
    \end{itemize}

  \item \textbf{Sodium-calcium overload cascade: molecular precision upgrade and AIMM concept}
    (Chapter~6, Section ``The Sodium-Calcium Overload Cascade''; Typst mirror updated):
    The sodium-calcium cascade description has been substantially upgraded with
    molecular-level precision. Previous text described ``NCX reversal'' and
    ``calcium accumulation'' in generic terms; the revised section identifies the
    specific transporters by gene name (NCX1/SLC8A1, NCLX/SLC8B1, NHE1/SLC9A1),
    derives the NCX1 reversal potential equation ($E_\text{NCX} = 3E_\text{Na} - 2E_\text{Ca}$)
    and explains its thermodynamic implications, and distinguishes two independent
    calcium-trapping mechanisms: NCX1 reverse-mode influx and NCLX efflux block
    due to elevated cytoplasmic sodium---a mechanism absent from the previous version.
    Petter et al.~(2022) $^{23}$Na-MRI data added as first-in-human quantitative
    evidence (elevated baseline sodium, $p=0.0034$; inverse correlation with
    grip strength, $p=0.0319$). The AIMM (acquired ischemic mitochondrial myopathy)
    concept from Scheibenbogen and Wirth (2025) is introduced, framing the ionic
    cascade as a self-reinforcing acquired pathology rather than a primary genetic
    disease---directly relevant to pharmacological reversibility. The multi-level
    vicious cycle hypothesis updated to incorporate AIMM and cross-reference
    the MDC002 therapeutic entry.
    Motivated by: MDC002 mechanism of action review identified that the existing
    cascade description was molecularly imprecise and did not account for the
    NCLX-mediated mitochondrial calcium trap, which is a distinct pharmacological
    target from NCX1.
    \cite{WirthScheibenbogen2020,WirthScheibenbogen2021muscle,Petter2022sodium,Scheibenbogen2025muscle}.

  \item \textbf{Ion homeostasis therapeutics: MDC002 (Mitodicure)}
    (Typst Chapter~18, new Section ``Ion Homeostasis Therapeutics: MDC002 (Mitodicure)'',
    label \texttt{sec:mdc002}):
    New section covering MDC002, an oral small-molecule preclinical candidate
    developed by Mitodicure GmbH (Frankfurt) targeting the ionic cascade of
    the AIMM model. Content covers: mechanism of action (dual Na$^+$/K$^+$-ATPase
    stimulation and NCLX activation, addressing both sides of the mitochondrial
    calcium trap); patent basis (three applications specifying NHE1 inhibition,
    PDE7 inhibition, and benzoxazole/benzofuran scaffolds); the mechanistic
    significance of PDE7 inhibition as a downstream bypass for impaired
    beta-2 adrenergic receptor signalling; development status (preclinical;
    BfArM sign-off for first-in-human; Phase~I projected autumn 2025);
    a \texttt{limitation} environment explicitly stating that no human efficacy
    data exist; and an open question on which ME/CFS subtype benefits most
    (AIMM-predominant vs.\ autoimmune subtype), proposing stratified trial
    design using $^{23}$Na-MRI, GPCR autoantibody titres, and clinical subtype criteria.
    Motivated by: MDC002 is the most mechanistically targeted emerging candidate
    in the ionic dysfunction space, directly derived from the AIMM model whose
    supporting evidence is peer-reviewed; absence from the treatment chapter
    was a gap given the detailed coverage of immunoadsorption, daratumumab, and BC007.
    \cite{Scheibenbogen2025muscle,WirthScheibenbogen2021muscle,Petter2022sodium,WirthScheibenbogen2020}.

  \item \textbf{Attractor emergence analytical proof: bistability derived from first principles}
    (Chapter~30, new Section ``Attractors as an Emergent Consequence of the
    Biochemistry'', label \texttt{sec:attractors-emergent}):
    New analytical subsection proving that the model's bistability is a
    \emph{prediction} of the biochemistry rather than a modelling premise. A
    reduced 2D system (ATP concentration $A$ and activated immune cell count $I$)
    is derived from the full ODE system by quasi-steady-state reduction. Nullcline
    analysis (Equations~\ref{eq:i-nullcline}--\ref{eq:a-nullcline}) shows that the
    $I$-nullcline is sigmoidal (Hill kinetics) and the $A$-nullcline is non-monotone
    (ETC saturation); their triple intersection under ME/CFS parameters---satisfying
    condition~\ref{eq:bistability-inequality}---proves bistability analytically. Jacobian
    eigenvalue analysis classifies all three fixed points: the healthy-adjacent fixed
    point and the disease fixed point are both stable nodes; the intermediate saddle
    lies on the separatrix. The healthy regime yields a single stable fixed point;
    the ME/CFS regime yields two. Three structural conclusions established:
    (1)~ME/CFS parameters preserve the healthy attractor but shrink its basin;
    (2)~the disease attractor does not exist under healthy parameters;
    (3)~bistability requires only Hill kinetics and a concave production function---both
    standard biochemistry. A \texttt{limitation} environment notes that quantitative
    claims (exact bifurcation thresholds, basin sizes) require numerical continuation
    software and have not been computed.
    Motivated by: the bifurcation narrative in Chapter~30 asserted bistability as a
    qualitative claim without proof; the analytical derivation is necessary to support
    the claim that bistability is an emergent property of the biochemistry rather than
    an assumed feature, and to establish the epistemological basis for model-derived
    predictions about treatment.

\end{itemize}

\subsection*{Document Infrastructure}

\begin{itemize}

  \item \textbf{Part~V Typst migration} (Chapters~27--33 and Appendix~E):
    Part~V chapters migrated from LaTeX stub files to full Typst source.
    Hypothesis registry converted from LaTeX comment syntax (\texttt{\%}) to Typst
    (\texttt{//}); landscape tables added.
    Motivated by: progressive migration of the document from LaTeX to Typst.

  \item \textbf{Front matter layout fixes}:
    Roman numeral page numbering restored for front matter pages (title page,
    abstract, changelog, table of contents). Title page footer suppressed.
    Appendix separator page added with letter-based appendix numbering (A, B, C\ldots).
    Motivated by: layout regression identified after Typst migration.

\end{itemize}

\subsection*{Corrections}

\begin{itemize}

  \item \textbf{PEM definition corrected from ``worsening'' to ``clearance failure'' framing}
    (Ch.~2 core symptoms; Ch.~15 metabolic danger signals section; Ch.~16
    criterion~2; Ch.~27 exertion crisis section; Ch.~14 treatment PEM management
    section; appendix-a-terminology; patient-faq):
    All definitions of PEM have been revised. The previous formulation---``a
    disproportionate worsening of symptoms following exertion''---implied that
    exertion itself triggers a pathological cascade, which contradicts the
    framework's own analysis (Ch.~16 classifies PEM's deferred-cost architecture
    as \emph{normal physiology exploited by broken restoration machinery}).
    Definitions now read: ``the persistence and accumulation of post-exertion
    byproducts --- metabolic waste, inflammatory mediators, oxidative damage ---
    that healthy physiology clears within hours but ME/CFS physiology cannot.''
    The Ch.~15 section introduction, which described PEM as a ``pathological
    response to exertion,'' has been corrected to attribute pathology to the
    failure of clearance mechanisms rather than to the exertion response itself.
    Motivated by: internal consistency review identifying contradiction between
    the PEM definitional language used throughout the document and the mechanistic
    classification in Ch.~16.

\end{itemize}

\section*{Version 6 (2026-03-08 -- 2026-03-23)}

\subsection*{New Content}

\begin{itemize}

  \item \textbf{Causal hierarchy analysis: root causes, amplifiers, and consequences}
    (New Chapter~16 in Part~II; new Chapter~33 in Part~V; hypothesis registry
    updated with 6 new entries):
    Two sister chapters synthesizing all pathophysiology mechanisms from Chapters~6--15
    and all formal models from Chapters~27--32 into a causal hierarchy. Chapter~16
    introduces a three-tier classification framework (trigger-capable root causes,
    amplifiers, downstream consequences) with four discriminating criteria, identifies
    four trigger-capable mechanisms (CNS energy crisis, metabolic safe mode lock,
    GPCR autoantibody cascade, TRPM3 channelopathy), six amplifiers, and five
    downstream consequences. Proposes a ``multiple entry points, single final common
    pathway'' hypothesis and distinguishes load-bearing from secondary locks.
    Critically argues that root cause $\neq$ treatment priority due to tractability
    constraints. Chapter~33 formalizes the hierarchy using the 64-variable ODE model,
    introduces three new model modules (GPCR autoantibody dynamics, TRPM3/calcium
    signaling, bistable metabolic safe mode switch) extending the model to 67
    variables, and performs trigger sufficiency analysis, parameter sensitivity
    analysis, lock removal analysis, and minimum intervention set enumeration.
    Derives a formal ``Minimum Lock Release Theorem'' predicting that recovery from
    the severe/locked attractor requires simultaneous restoration of 3--4 parameters.
    Motivated by: the observation that the document catalogues 24+ mechanisms without
    distinguishing which are causally sufficient to generate ME/CFS versus which are
    downstream consequences; the need to connect qualitative pathophysiology reasoning
    to quantitative model predictions for research prioritization.
    \cite{Walitt2024NIH,Nakatomi2014neuroinflammation,Loebel2016,Sotzny2021,%
    Stein2024immunoadsorption,Fluge2025daratumumab,Sasso2026trpm3}.

  \item \textbf{Transcranial magnetic stimulation (TMS) for ME/CFS: expanded evidence review}
    (Chapter~18, Section~\ref{subsubsec:tms-mecfs}; Appendix~H, new annotated
    bibliography subsection; hypothesis registry updated with 1 new entry):
    Expanded the TMS stub into a comprehensive section covering neurophysiological
    rationale (motor cortex hypoexcitability demonstrated by diagnostic TMS studies),
    all four ME/CFS therapeutic studies (Kakuda 2016, Yang 2020, Miwa 2023, Miwa 2025;
    total $n \approx 72$, all open-label), indirect sham-controlled evidence from
    fibromyalgia meta-analyses (11--18 RCTs, $n = 303$--643), and practical safety
    considerations.
    Motivated by: Telegraph article on TMS for ME/CFS (March 2026) and MEA response
    noting absence of research evidence; literature review revealed more evidence than
    initially apparent, warranting balanced coverage rather than dismissal.
    \cite{Kakuda2016rTMS,Yang2020rTMS,Miwa2023rTMS,Miwa2025rTMS,Samii1996TMS,Starr2000TMS,Davey2003TMS,Lefaucheur2020NIBS,Toh2022fibrorTMS,Su2021fibrorTMS}.

  \item \textbf{Meningitis--ME/CFS intersection: infection susceptibility, neuroinflammatory overlap, and vaccination}
    (Chapter~7, new Sections~7: ``Infection Susceptibility in ME/CFS,''
    ``Vaccination in ME/CFS: Efficacy and Safety,'' expanded complement
    section with exercise-induced C4a/PEM data;
    Chapter~8, new Section~8: ``Meningitis as a Model for Acute
    BBB-Neuroinflammatory Injury'' with glymphatic system and dural mast cell
    content, ``Double Vulnerability'' speculation;
    Chapter~14d, new paragraphs on post-meningitis fatigue syndrome and
    Q fever as bacterial trigger paradigm;
    Appendix~H, 7 new annotated bibliography entries;
    hypothesis registry updated with 2 new entries):
    Comprehensive integration of meningitis--ME/CFS research across five
    convergent mechanistic domains: (1)~microglial priming/second-hit hypothesis,
    (2)~oxidative/nitrosative stress overlap (peroxynitrite pathway),
    (3)~non-ischaemic mitochondrial dysfunction, (4)~kynurenine pathway
    activation, (5)~endothelial dysfunction and microclots. New content on
    exercise-induced complement activation (C4a) as PEM biomarker with
    4-study replication chain (Sorensen~2003, Nijs~2010, Polli~2019,
    Glass/Hanson~2025). Vaccination safety section establishes that infection,
    not antigen exposure, triggers CFS/ME (Magnus~2015: HR~2.04 for infection
    vs.\ HR~0.97 for vaccine). Glymphatic section introduces the 2025 Cell
    paper on dural mast cells regulating CSF dynamics at ACE points---a novel
    mechanistic bridge between MCAS and glymphatic waste clearance.
    Motivated by: 2026 UK meningitis outbreak prompting analysis of ME/CFS
    patients' infection vulnerability; identification of meningitis as an
    underexplored post-infectious ME/CFS trigger with unique CNS-direct
    inflammatory insult.
    \cite{Hotopf1996meningitis,Schwitter2024meningitis,Ungureanu2021meningitis,%
    Magnus2009meningococcal,Magnus2015influenza,Prinsen2012vaccine,%
    Sorensen2003complement,Nijs2010pem,Polli2019complement,Glass2025ev,%
    iu2024tcell_exhaustion,Spronk2023qfever,Pinas2022glymphatic,%
    Christodoulides2024meningitis,Caligiuri1987nk,Scheld1999oxidative,%
    Schwerk2015meningitis,VanderLinden2023microclots,Perrin2025glymphatic,%
    Midtbo2014meningitis,Gil2023cd8,Klein2024longcovid,Bested2001bbb}

  \item \textbf{Food-triggered mast cell activation in ME/CFS}
    (Chapter~14d, new Section~14d.5 ``Food-Triggered Mast Cell Activation: The
    Silent Pathway''; Chapter~19, expanded Section~19.2.2 ``Mast Cell Activation
    Syndrome''; Chapter~20, new Section~20.5.4 ``Mast Cell Mediator Biomarkers'';
    hypothesis registry updated): Comprehensive new content on non-IgE pathways
    by which food triggers mast cell degranulation without positive allergy tests
    (MRGPRX2, complement C3a/C5a, lectin--IgE glycan cross-linking, food additive
    oxidative pathways). Mediator-to-symptom mapping for fatigue, brain fog, POTS,
    and GI distress. Practical elimination-rechallenge protocol adapted for
    energy-limited patients. Urinary paired mediator testing (NMH, LTE4,
    PGD$_2$ metabolite) as diagnostic approach. Two new hypothesis/speculation
    environments with falsifiable predictions. MCAS overdiagnosis warning with
    strict diagnostic criteria.
    Motivated by: clinical question whether silent (non-IgE) food-triggered mast
    cell degranulation contributes to ME/CFS symptoms in patients with negative
    allergy testing.
    \cite{Rohrhofer2025mecfsmast,Kohno2021potsmast,Roy2021mrgprx2,%
    Thapaliya2022mrgprx2neuro,EliehAliKomi2020complement,%
    ComasBaste2020histamine,Folkerts2018fiber,Voelker2025biomarkersmast,%
    Voelker2024urinemast,Castells2024mcasreview,Gulen2024mcascriteria,%
    Maeda2017pgdm,Rohrhofer2024stratification,Tsilioni2022exosome,%
    Hatziagelaki2018hypothalamus,Pepper2020additives,Weinstock2023MCASneuro}

  \item \textbf{Cognitive dysfunction quantification and post-COVID parallels}
    (Chapter~8, new Section~8.7.2 ``Quantifying Cognitive Impairment:
    Population-Scale Evidence''): Expanded the cognitive dysfunction section with
    quantitative meta-analytic data (effect sizes $g = -0.55$ to $-0.82$ across
    domains), multi-site objective assessment results (MCAM speed--accuracy
    dissociation), and a comprehensive subsection on post-COVID cognitive decline
    including population-scale IQ-equivalent losses ($\sim$3/6/9 points by
    severity), objective cognitive slowing (3~SD), structural neuroimaging
    correlates (grey matter loss), and direct ME/CFS vs.\ Long COVID comparison
    (sustained attention 83\% vs.\ 56\% impaired). Cross-disease cognitive
    comparison added to Chapter~14d (Long COVID paragraph expanded with
    quantitative evidence). Post-COVID cognitive population impact added to
    Chapter~23 (Post-COVID Surge subsection expanded).
    \textit{Motivated by:} Cognitive impairment descriptions lacked quantitative
    data despite available meta-analytic evidence; post-COVID cognitive research
    provides convergent evidence and population-scale
    context~\cite{Hampshire2024cognition,Sebaiti2022cognitive,Lange2024cognitive,Zhao2024cognitive,Douaud2022brain,Azcue2022brainfog}.

  \item \textbf{Stimulation intolerance integrated model} (Chapter~15, new
    Section~15.19): Unified explanation of why ME/CFS patients cannot tolerate
    physical, sensory, cognitive, and emotional stimulation. Synthesizes energy
    bottleneck, central sensitization, microglial amplification, inhibitory brake
    failure, cross-modality convergence, interoceptive amplification, and clinical
    implications into a single coherent model.
    \textit{Motivated by:} Mechanistic explanations were present but scattered
    across ch02, ch03, ch08, ch14h, and ch15; no unified narrative existed linking
    energy depletion to sensory
    amplification~\cite{RenzPolster2022,Nijs2017neuroinflammation,Baraniuk2025NociplasticMECFS}.

  \item \textbf{Nerve sheath pain mechanisms} (Chapter~15,
    Section~15.11 subsection): New subsection explaining why nerve sheaths become
    painful via four converging mechanisms: Schwann cell TRPA1 activation by
    ROS/CGRP, endoneurial hypoxia from vasa nervorum dysfunction, autoimmune
    targeting of nerve sheath antigens, and perineurial mast cell degranulation.
    Includes convergent vulnerability
    hypothesis~\cite{heng2025mecfs,Oaklander2016autoimmuneSFN,Oaklander2022SFN}.
    \textit{Motivated by:} Patients report nerve trunk pain distinct from SFN or
    central sensitization; no existing section explained the mechanism.

  \item \textbf{Periarticular and muscular pain mechanisms} (Chapter~15,
    Section~15.11 subsection): New subsection explaining why pain localizes around
    joints and in muscles despite absence of inflammatory arthritis or
    exercise-induced injury. Covers periarticular mast cell density, muscle
    metabolic nociception (ASIC3 combinatorial gating, P2X3, TRPV1), fascial
    innervation, microvascular ischemia--reperfusion, and interaction with central
    sensitization~\cite{Lien2019lactate,Jammes2021muscle,Nijs2012painCFS}.
    \textit{Motivated by:} Common patient complaint without a unified mechanistic
    explanation in the document.

  \item \textbf{Four-compartment nociceptive model} (Chapter~29, new
    subsections): Mathematical models for nerve sheath inflammation dynamics
    ($\mathcal{N}_s$), periarticular nociceptive input ($P_{\text{peri}}$),
    muscular metabolic nociception ($P_{\text{musc}}$ with ASIC3 upregulation
    kinetics), and a unified four-compartment total nociceptive input equation
    replacing the prior single $P_{\text{noci}}$. Each compartment has distinct
    temporal dynamics and therapeutic targets. Full Model Application Guide entry
    with worked example, measurements, and falsification criteria.
    \textit{Motivated by:} Existing central sensitization model could not predict
    pain distribution or temporal profile; tissue-specific generators required
    compartmental decomposition.

  \item \textbf{Healthy exercise response baseline model} (Chapter~27, new
    Section~27.3; Chapter~31, expanded Section~31.4.1): Formal model of the
    healthy exertion--recovery--adaptation cycle using the same ODE framework as
    the ME/CFS models, providing face validation of the model and identifying the
    precise branch point where physiology becomes pathology. Introduces metabolic
    headroom ratio $R_{\text{headroom}}$ (Eq.~27.8), supercompensation integral
    $\Delta M_h^{\text{net}}$ (Eq.~27.9), iterated training adaptation
    (Eq.~27.10), DOMS inflammatory timeline model (Eq.~27.11), and critical
    branch-point condition (Eq.~27.12). New hypothesis: Supercompensation Failure
    as Mechanistic Basis of Exercise Intolerance (certainty 0.50) with four
    testable predictions. Full Model Application Guide entry with healthy vs.\
    ME/CFS worked example. Hypothesis registry updated.
    \textit{Motivated by:} The model's equations were only validated against
    ME/CFS data; reproducing well-characterized healthy exercise physiology
    (DOMS timing, supercompensation window, training adaptation) with the same
    equations at healthy parameters constitutes a first face validation and
    clarifies why GET harms
    patients~\cite{Hood2009biogenesis,Clarkson2002DOMS,Peake2017muscle,Geraghty2019GET}.

  \item \textbf{BC007 (rovunaptabin) section expanded} (Chapter~18,
    Section~18.2.1): Expanded from a single paragraph to a full subsubsection
    covering mechanism of action (selective aptamer-based GPCR autoantibody
    neutralisation vs.\ non-selective immunoadsorption), clinical evidence
    (Hohberger 2021 case report, Phase~IIa $n$=29, Phase~II RCT $n$=114
    NCT05911009), comparison table with immunoadsorption across seven
    dimensions, limitations and open questions (post-COVID-only cohorts,
    unknown duration of effect, upstream production not addressed), energy
    profile (Category~A--B), and cross-reference to the GPCR-endothelial
    cascade hypothesis.
    \textit{Motivated by:} BC007 evidence was documented only in appendices
    (Appendix~H annotated bibliography, Appendix~I research registry) but
    absent from the main treatment chapter; immunoadsorption and daratumumab
    had substantially more developed coverage in
    ch18~\cite{Hohberger2021bc007}.

  \item \textbf{Evolutionary deferred-cost hypothesis for PEM timing}
    (Chapter~14e, new Section~14e.8 ``The Evolutionary Deferred-Cost Hypothesis
    for PEM Timing'', label \texttt{spec:evolutionary-deferred-cost}):
    New speculation integrating evolutionary medicine, sickness behaviour biology,
    and ME/CFS adrenergic physiology into a unified account of why PEM has a
    characteristic 12--72 hour delay. Central claim: the deferred-cost
    architecture (act now, pay later) was selected for over evolutionary time
    because organisms that could override fatigue and pain during acute threats
    survived; in ME/CFS the emergency override mechanism is intact but the
    restoration phase is broken. Includes: evolutionary rationale with references
    to endurance running evolution and stress-induced analgesia; evidence for
    preserved adrenergic override (meta-analysis of adrenergic dysfunction in
    ME/CFS); evidence for broken restoration (post-exertional cytokine cascade);
    relationship to the ``stuck sickness behaviour'' hypothesis in ch14a;
    the ``you seemed fine'' social misunderstanding; severity gradient prediction;
    and four testable predictions. Registry updated with new Domain~10
    (Evolutionary Medicine). Six new citations added
    (\texttt{Hart1988SickBehavior}, \texttt{Dantzer2007SicknessBehavior},
    \texttt{Bramble2004EnduranceRunning}, \texttt{Marino2022FatigueEvolution},
    \texttt{Parikh2011StressAnalgesia}, \texttt{Hendrix2025AdrenergicDysfunction});
    post-exertional cytokine evidence uses pre-existing \texttt{Moneghetti2018}.
    \textit{Motivated by:} Author observation that the PEM delay is unexplained
    by current mechanistic accounts; evolutionary framing provides a principled
    reason why the delay is normal biology rather than a disease
    anomaly~\cite{Hart1988SickBehavior,Dantzer2007SicknessBehavior,%
    Bramble2004EnduranceRunning,Marino2022FatigueEvolution,%
    Parikh2011StressAnalgesia,Che2025,Moneghetti2018,%
    Hendrix2025AdrenergicDysfunction}.

  \item \textbf{MCAS antihistamine up-dosing section}
    (Chapter~15, new Section ``Antihistamine Up-Dosing in MCAS''): Added
    comprehensive guidance on higher-than-standard antihistamine dosing for MCAS,
    covering four H1 antihistamines (cetirizine, fexofenadine, loratadine,
    rupatadine) and two H2 antihistamines (famotidine, cimetidine). Includes
    drug-specific dosing tables with standard vs.\ MCAS ranges, evidence grades
    and responder rates, rupatadine-specific escalation protocol, H2 dose
    escalation guidance, and a practical 4-step up-dosing protocol with
    ME/CFS-specific cautions. Cross-references added in Chapter~19 MCAS
    treatment section. Six annotated bibliography entries added to Appendix~H.
    Nine new references added to bibliography.

    \textit{Motivated by:} Document gap identified during clinical dosing review
    --- existing MCAS treatment sections used standard allergy doses without
    reflecting the fourfold up-dosing recommendation from the
    EAACI/GA\textsuperscript{2}LEN urticaria guideline or MCAS specialist
    protocols~\cite{Zuberbier2022urticaria,Afrin2013MCAS,%
    Molderings2016pharmacological,Podder2023updosing,%
    VandenElzen2017updosing,Salvucci2023antihistamineLongCOVID,%
    Nurmatov2015H1systematic,Magerl2015rupatadine,Izquierdo2024rupatadine}.

  \item \textbf{IgG immune complex--mediated endothelial and mitochondrial injury}
    (Chapter~6, new Section~\ref{sec:igg-mito-disruption}; Chapter~7, expanded
    immunoadsorption paragraph; Chapter~10, new
    Section~\ref{sec:igg-endothelial}; hypothesis registry updated with 1 new
    entry):
    Integration of Liu et al.\ (2026) demonstrating that IgG immune complexes
    purified from post-infectious ME/CFS patient sera ($n = 39$--40) enter
    endothelial cells, induce mitochondrial fragmentation, and trigger
    IL-1$\beta$ secretion. Mass spectrometry proteomics revealed
    subtype-specific signatures: classic ME/CFS IgG enriched in ECM
    reorganisation pathways, post-COVID ME/CFS IgG enriched in hemostasis and
    clot formation pathways. Chapter~6 covers the mitochondrial effects
    (fragmentation without bioenergetic collapse, glycolytic compensation,
    female-predominant effect). Chapter~7 links the finding to immunoadsorption
    efficacy, providing a mechanistic rationale for IgG removal. Chapter~10
    positions immune complex deposition as a candidate mechanism for the
    endothelial dysfunction documented by FMD, proteomic, and capillary
    remodelling evidence. New speculation on immune complex--endothelial injury
    as a central vascular mechanism (certainty: 0.35) with three falsifiable
    predictions (immunoadsorption improving FMD, biopsy IgG deposition, in
    vitro fragmentation correlating with EndoPAT scores).
    \textit{Motivated by:} The document documented autoantibodies and
    immunoadsorption efficacy but lacked a direct mechanistic link between
    circulating IgG and end-organ (endothelial, mitochondrial)
    injury~\cite{Liu2026IgGcomplexes}.

\end{itemize}

\subsection*{Literature Integration}

\begin{itemize}

  \item \textbf{ACHTSAM study protocol integrated} (Fricke et al.\ 2026,
    \textit{BMJ Open}): First systematic investigation of home-based diagnostic
    tolerability in severe/very severe ME/CFS (Bell score $\leq$30). Cited in
    ch05-severity-levels (invisible population), ch10-cardiovascular (EndoPAT
    endothelial function), ch21-clinical-trials (severe patient exclusion gap),
    appendix-c (home-based assessment battery), and appendix-h (annotated
    bibliography).
    \textit{Motivated by:} Addresses critical research gap in which severe
    patients are systematically excluded from clinical
    studies~\cite{Fricke2026achtsam}.

  \item \textbf{NETosis and immune dysregulation batch integration} (2026-03-12):
    Integrated 11 new peer-reviewed and preprint papers on neutrophil extracellular
    traps (NETs), autoantibodies, viral persistence, and immunological biomarkers.
    Chapter 7 (Immune System Dysfunction) expanded with: Krinsky2023 achievement
    documenting persistent NETosis induction in long COVID as potential pathogenic
    driver; Son2023 observation identifying anti-nuclear autoantibodies at 3 months
    post-infection as predictors of persistent symptoms; Schonrich2016, Monsalve2025,
    Veras2023, Hetland2022 inline citations establishing viral NETosis triggers,
    NLRP3-driven autoimmunity cascade, DNase-I therapeutic efficacy, and calprotectin
    as NET burden biomarker; Gao2025LongCovidBiomarkers achievement describing
    distinct proteomic signatures (CCL3/CD40/IKBKG/IL-18/IRAK1) identifying Long COVID
    breathlessness endotype; Peluso2024ViralPersistence achievement documenting viral
    RNA persistence in gut tissue up to 676 days with constitutive innate immune
    sensor activation; Frustrated NETosis hypothesis (certainty: 0.35) linking
    mitochondrial energy failure to incomplete NET formation and autoantigen exposure;
    Calprotectin-PEM open question (certainty: testable) examining whether serial
    calprotectin measurement post-exertion could objectify post-exertional malaise
    through NET dynamics; NET-autoantibody synthesis paragraph positioning NETosis as
    upstream driver and autoantibodies as downstream effectors of unified post-viral
    autoimmune cascade. Chapter 8 (Neurological Dysfunction) expanded with:
    GuedesDeSa2024 hypothesis on autoantibody-driven neurological dysfunction with
    passive transfer causal evidence; VanElzakker2024NeuroinflammationProject context
    reference to ongoing PET-MRI neuroinflammation study. Chapter 10 (Cardiovascular
    Dysfunction) expanded with Weckbach2019 inline citation on cardiac NET-mediated
    fibrosis and ventricular dysfunction. Chapter 6 (Energy Metabolism) expanded with
    Cullen2026ATP speculation on stress-responsive mitochondrial failure phenotype
    paralleling post-exertional energy collapse; NETosis-as-metabolic-sink speculation
    (certainty: 0.30) proposing that energy-expensive NETosis acts as metabolic drain
    competing with normal cellular demands, creating energy depletion cycle.
    \textit{Motivated by:} Literature-integrator phase identified 11 relevant papers
    on NETosis dysregulation, autoimmunity, and viral persistence mechanisms directly
    applicable to ME/CFS pathophysiology, with novel cross-paper synthesis revealing
    frustrated NETosis hypothesis linking energy depletion to incomplete autoantigen
    generation and calprotectin dynamics potentially mediating
    PEM~\cite{Krinsky2023,Son2023,Schonrich2016,Monsalve2025,Veras2023,Hetland2022,Weckbach2019,Gao2025LongCovidBiomarkers,GuedesDeSa2024,Peluso2024ViralPersistence,Cullen2026ATP,VanElzakker2024NeuroinflammationProject}.

  \item \textbf{Viral persistence, biomarkers, and neuroinflammation batch integration} (2026-03-12):
    Integrated 6 new peer-reviewed papers on viral reservoir targeting, blood biomarkers,
    spike protein neuroinflammatory priming, IL-18-mediated CD8 suppression, circulating
    neuroinflammation markers, and pediatric long COVID prevention.
    Chapter~7 (Immune System Dysfunction) expanded with: Beentjes2025Biomarkers achievement
    documenting hundreds of blood-based traits distinguishing ME/CFS independent of inactivity
    (UK Biobank $n=131{,}758$, replicated in All-of-Us, certainty: 0.75); Proal2025Reservoir
    achievement establishing international consensus framework for viral reservoir targeting
    in long COVID via Lancet Infectious Diseases position paper synthesizing persistent
    SARS-CoV-2 evidence and trial design principles (certainty: 0.75); Autaa2025CD8Aging
    inline citation on IL-18-mediated CD8 suppression with aging, linking immunosenescence
    to impaired viral reservoir clearance in older post-COVID ME/CFS patients (certainty: 0.70).
    Chapter~8 (Neurological Dysfunction) expanded with: Frank2024S1Priming hypothesis
    documenting S1 spike subunit sensitization of brain innate immunity in rats with
    sustained neuroinflammatory priming (MhcII$\alpha$, Nlrp3, Tlr4 upregulation persisting
    7 days), proposing spike-driven TLR4 disinhibition as mechanism for neuroinflammatory
    threshold lowering and symptom relapse post-infection (certainty: 0.55, registered in
    hypothesis-registry); Omdal2026LongCovidNeuroinflammation observation documenting null
    circulating neuroinflammation markers (NfL, GFAP) in long-COVID case-control study
    ($n=96$, median 69 weeks post-infection) as balancing evidence that blood biomarkers
    may not detect localized glial activation (certainty: 0.55).
    Chapter~23 (Epidemiology) expanded with: Wu2024LongCovidChildren achievement documenting
    BNT162b2 effectiveness against long COVID in children and adolescents ($n>385{,}000$,
    95.4\% effectiveness in adolescents during Delta), with causal mediation analysis
    showing protection operates via infection prevention rather than post-infection immune
    modification (certainty: 0.80), confirming that preventing the acute infection event
    is fundamentally more effective than modifying post-infection immune trajectory.
    New paragraph ``Vaccination as Prevention Strategy in Children and Adolescents'' added
    to Post-COVID Surge subsection.
    \textit{Motivated by:} Six high-impact papers from 2024--2025 providing landmark
    evidence on viral persistence as therapeutic target, activity-independent blood biomarker
    signatures in ME/CFS with replication, spike protein-driven neuroinflammatory priming
    as mechanism linking viral antigen to prolonged neurological sensitization, and pediatric
    long COVID as preventable syndrome providing population-level intervention evidence and
    mechanistic clarity on infection-driven vs.\ immune-modulation approaches~\cite{Beentjes2025Biomarkers,Proal2025Reservoir,Frank2024S1Priming,Autaa2025CD8Aging,Omdal2026LongCovidNeuroinflammation,Wu2024LongCovidChildren}.

  \item \textbf{Chapter~9 (Endocrine Dysfunction) --- post-viral pituitary injury} (2026-03-13):
    New subsubsection \textit{Post-Viral Pituitary Injury and Secondary Adrenal Insufficiency}
    (\S\ref{sec:postviral-pituitary}) added to the HPA Axis section, covering a fourth
    mechanistic pathway distinct from the three functional mechanisms already described.
    Content includes: achievement environment for Leow et al.\ 2005 SARS-1 data (39.3\%
    central hypocortisolism in 61 survivors); integration of Carosi et al.\ 2024 COVID-19
    pituitary review; Ruiz-Pablos et al.\ 2024 autoimmune hypophysitis model; new
    \texttt{fhypothesis} environment (\texttt{hyp:postviral-pituitary}, certainty 0.50) with
    four testable predictions; \texttt{warning} environment on the inadequacy of morning
    cortisol for detecting secondary AI; \texttt{limitation} environment on SARS-1
    generalisability. Existing HPA \texttt{limitation} block updated to reference Leow 2005 as
    closest-to-prospective evidence. Future Research Directions updated to include prospective
    pituitary evaluation as a priority. Hypothesis registry updated (Domain~4, new entry).
    Appendix~C updated with matching warning block cross-referencing ch09. Chapter~29 model
    falsification criteria amended with subgroup qualifier distinguishing structural from
    functional HPA phenotypes.
    \textit{Motivated by:} LinkedIn post by Manuel Ruiz-Pablos (2026-03-09) identifying
    structural pituitary injury as a gap in coverage; SARS-1 evidence (Leow 2005) and
    COVID-19 pituitary literature providing strong foundational evidence for post-viral
    secondary AI as a distinct and under-recognised ME/CFS mechanism.
    Key sources: \cite{Leow2005sars,Ruiz2024hpa,Carosi2024hypopituitarism}.

\end{itemize}

\section*{Version 5 (2026-02-18 -- 2026-03-07)}

\subsection*{Document Infrastructure}

\begin{itemize}

  \item \textbf{Revision bar system removed}: The \texttt{\textbackslash rev\{N\}}/\texttt{\textbackslash revend}
    paragraph-marking system (changebar, marginnote packages) has been fully removed and
    replaced with no-op stubs. All markers stripped from 24~chapter files. Improves
    build performance and reduces source noise; the changelog remains the sole
    record of changes.

  \item \textbf{Epistemic stratification --- \texttt{limitation} environment}: New
    \texttt{limitation} and \texttt{limitation*} environments deployed across
    40+ locations in Parts~I--IV to explicitly mark what the data do \emph{not}
    allow concluding. Adversarial review round corrected 15~findings across the
    initial deployment.
    \textit{Motivated by:} Reviewer feedback requesting a fourth epistemic layer
    (``what not to conclude'') beyond \texttt{hypothesis}/\texttt{speculation}/\texttt{warning}.

  \item \textbf{Model-unique insight environment}: New \texttt{modelunique}
    \texttt{tcolorbox} environment for conclusions that emerge only from formal
    mathematical modeling, visually distinct from empirical observations.

  \item \textbf{Methodology section}: New front-matter section describing
    the document's methodology, evidence assessment framework, and epistemic
    conventions.

  \item \textbf{Build system}: Added \texttt{.latexmkrc} configuration and
    TeX memory settings (\texttt{TEXMFCNF}) for large-document builds.

  \item \textbf{Biomedical review system}: Added 13~specialized auditor agents
    (pathway, immunology, biochemistry, neuro, microbiome, biomarker,
    epidemiology, comorbidity, safety, pharmacology, patient-safety,
    falsifiability, bibliography) and the \texttt{review-biomedical} skill for
    comprehensive 20-category scientific review.

  \item \textbf{Changelog convention}: V5 entries now include ``why'' (motivation)
    and key sources for each change, per reviewer recommendation.

\end{itemize}

\subsection*{Part I --- Clinical Overview}

\begin{itemize}

  \item \textbf{Chapter 1 --- Introduction to ME/CFS}: Expanded coverage of
    ME/CFS terminology, disease classification, and epidemiological context.

  \item \textbf{Chapter 5 --- Disease Course and Subgroups}: Enhanced endotype
    table with reproducibility status (R/P), evidence levels (A/B/C), and
    biomarker accessibility tiers. Added \texttt{limitation} environment for
    provisional classification status.
    \textit{Motivated by:} Reviewer request for integrated endotype table with
    evidence grades.

  \item \textbf{Patient FAQ}: New chapter in front matter with 12 questions
    addressing common patient concerns (What is ME/CFS?, PEM, exercise, diagnosis,
    treatments, supplements, Long COVID overlap, promising research).
    \textit{Motivated by:} Reviewer request for accessible patient-facing content.

\end{itemize}

\subsection*{Part II --- Pathophysiology}

\begin{itemize}

  \item \textbf{Chapter 6 --- Energy Metabolism}: Major expansion covering new
    research on mitochondrial dysfunction, metabolic trap hypothesis, and the
    relationship between immune activation and energy production. Three-pass
    scientific rigor review of heterogeneity and energy sections.
    \textit{Motivated by:} Wang et al.~2023 (WASF3--Complex~IV mechanism)~\cite{wang2023wasf3};
    Keller et al.~2024 (two-day CPET)~\cite{keller2024cpet};
    Cheema et al.~2023 (miRNA convergence)~\cite{Cheema2023mirna}.

  \item \textbf{Chapter 7 --- Immune System Dysfunction}: Integrated new findings
    on the gut-immune-brain axis, including the role of vagal signalling in
    immune dysregulation and the cholinergic anti-inflammatory pathway. Resolved
    22~citation violations. Substantially expanded Neutrophil Extracellular Traps
    (NETs) section with evidence on NET dysregulation in COVID-19 and long COVID,
    including impaired NET degradation via DNase1L3 deficiency, thrombo-inflammatory
    mechanisms (D-dimer elevation, microthrombi), and persistent multi-organ
    dysfunction. Proposes hypothesis that ME/CFS may involve defective NET
    clearance despite suppressed baseline neutrophil activation.
    \textit{Motivated by:} Heng et al.~2025 (T cell effector memory profiling)~\cite{heng2025mecfs};
    scientific rigor audit identifying uncited claims; Garcia et al.~2024 (NET
    degradation dysregulation in COVID-19 severity)~\cite{GarciaLabroucheColomer2024};
    Romano et al.~2022 (NETosis-driven thrombo-inflammation and coagulopathy)~\cite{Romano2022};
    Ewing et al.~2024 (persistent multi-organ damage in long COVID)~\cite{Ewing2024}.

  \item \textbf{Chapter 8 --- Neurological Dysfunction}: Added the
    enterochromaffin-vagal pathway hypothesis (Wirth--Scheibenbogen 2025),
    explaining how gut dysbiosis drives autonomic dysfunction via serotonin and
    vagal mechanisms.
    \textit{Motivated by:} Barton et al.~2023, 2025 (enterochromaffin serotonin--vagal
    signaling)~\cite{Barton2023,Barton2025}.

  \item \textbf{Chapter 10 --- Cardiovascular Dysfunction}: Expanded
    gut-mediated vagal impairment section with Wirth--Scheibenbogen 2025
    mechanistic support and new cross-references. Added replication status
    warnings.
    \textit{Motivated by:} van~Campen et al.~2024 (cardiac output--cerebral blood flow
    study, $n=534$)~\cite{VanCampenEtAl2024}.

  \item \textbf{Chapter 11 --- Gut Microbiome}: Expanded clinical implications of
    microbiome alterations and their connection to systemic symptoms. Added
    replication status warnings.

  \item \textbf{Chapter 13 --- Integrative Models}: Revised Heng 2025 multi-omics
    description; added replication warnings; expanded integrative content.

  \item \textbf{Chapter 14d --- Cross-Disease Comparisons}: Substantially
    expanded cross-condition analysis with new disease parallels and
    shared-mechanism evidence.

  \item \textbf{Chapter 14j --- Selective Energy Dysfunction Hypothesis}: Thorough
    scientific rigor review; strengthened internal consistency, corrected
    mathematical formulations, and expanded related-hypotheses discussion
    across 10 blocks.

  \item \textbf{Chapter 15 --- Symptom-Producing Mechanisms}: Updated MCAS
    prevalence citations and mast cell--microglia reference keys.

\end{itemize}

\subsection*{Part III --- Treatment}

\begin{itemize}

  \item \textbf{Chapter 14 --- Symptom Management}: Major expansion with
    new sections on pacing and energy envelope theory, PEM medications,
    sleep management, pain management, cognitive dysfunction management,
    and autonomic symptom management. Added \texttt{limitation} environments
    flagging sections with pending citations and evidence quality caveats.

  \item \textbf{Chapter 14b --- Action Plans for Mild to Moderate Cases}: Integrated
    a new acute-onset protocol for patients in early disease stages; expanded
    severity-stratified care pathways.

  \item \textbf{Chapter 14c --- Speculative model language}: Revised text
    presenting the time-dependent reversibility model from asserting it
    ``explains'' outcomes to stating it ``hypothesizes'' outcomes, with
    ``speculative'' qualifier added.
    \textit{Motivated by:} A8 audit findings (Issues~28--29, HIGH PRIORITY).

  \item \textbf{Chapter 15 --- Medications Targeting Underlying Mechanisms}:
    Revised with review corrections; removed unsupported claims.

  \item \textbf{Chapter 16 --- Supplements and Nutraceuticals}: Substantially
    revised; removed unsupported claims identified during adversarial review.

  \item \textbf{Chapter 17 --- Lifestyle Interventions}: Major expansion with
    comprehensive sleep optimization guidance (bedroom environment, schedule
    consistency, light exposure, temperature regulation, evening routines),
    expanded dietary considerations, and revised post-prandial pacing
    hypothesis language.

  \item \textbf{Chapter 18 --- Experimental and Emerging Therapies}: Corrected
    suramin text to explicitly state the cited pilot study~\cite{Naviaux2018suraminpilot}
    was conducted in autism spectrum disorder, not ME/CFS. Added energy profile
    for immunoadsorption; corrected nicotinamide riboside trial data and citation.
    \textit{Motivated by:} Assertion--source audit finding (DOES NOT SUPPORT).

  \item \textbf{Chapter 19 --- Integrative and Personalised Approaches}: Revised
    with review corrections; tightened evidence claims.

  \item \textbf{Environment type correction (Chapter 15)}: Converted
    \texttt{observation} to \texttt{hypothesis} for mechanistic treatment
    sequencing rationale (certainty: 0.25), fixing epistemic mismatch.
    \textit{Motivated by:} A8 audit finding (Issue~13, HIGH PRIORITY).

  \item \textbf{A8 audit completed}: Full Part~III hypothesis vs.\ recommendation
    audit identified 31 findings (3 critical fixed, 14 medium, 14 low priority).

\end{itemize}

\subsection*{Part IV --- Research}

\begin{itemize}

  \item \textbf{Chapter 21 --- Clinical Trials}: Major expansion with detailed
    coverage of rituximab trials (pilot through RituxME phase~III), low-dose
    naltrexone evidence, and other immune modulators. New section documenting
    9~key negative or null-result trials (rituximab phase~III, PACE trial, IVIG,
    acyclovir, valganciclovir, rintatolimod, galantamine, fluoxetine,
    hydrocortisone) with summary table and cross-cutting lessons.
    \textit{Motivated by:} Reviewer request for explicit coverage of negative
    evidence to combat confirmation bias.

  \item \textbf{Chapter 22 --- Mechanistic Studies}: Substantially expanded with
    detailed analysis of NIH deep phenotyping study (Walitt et al.\ 2024),
    WASF3 mitochondrial findings (Wang et al.\ 2023), and methodological
    critique. Corrected attribution (Hwang~$\to$~Wang).

  \item \textbf{Chapter 23 --- Epidemiology and Outcomes}: Major expansion
    with updated prevalence data, risk factor analysis, and outcome
    measurement discussion.

  \item \textbf{Chapter 24 --- Controversies}: Comprehensive rewrite covering
    nomenclature debate (ME vs.\ CFS vs.\ SEID), diagnostic criteria
    controversies (Oxford through IOM), and detailed PACE trial controversy
    analysis (methodology criticisms, reanalysis results, impact on GET/CBT
    recommendations).

  \item \textbf{Chapter 25 --- Translational Findings}: Revised shared-mechanism
    tables with added citations; corrected treatment dosing and trial data.

  \item \textbf{Chapter 25a --- Research Infrastructure}: Expanded with new
    content on research funding gaps and infrastructure needs.

  \item \textbf{Chapter 25b --- Proposed Research Studies}: New chapter describing
    fourteen original research protocols targeting key open questions in ME/CFS
    pathophysiology and treatment.

  \item \textbf{Hypothesis Registry}: Major expansion consolidating 40+ key
    hypotheses, speculations, predictions, and open questions from a systematic
    harvest of 305~environments across the document. Organized by mechanism
    domain with testable predictions and available tests.
    \textit{Motivated by:} Reviewer request for hypothesis--prediction--test mapping.

\end{itemize}

\subsection*{Part V --- Computational Modeling}

\begin{itemize}

  \item \textbf{Chapter 26 --- Modeling Foundations}: New section surveying
    existing mathematical and computational models of ME/CFS (Phair's IDO
    metabolic trap, Broderick's cytokine network analysis, Morris~\& Maes's
    neuro-immune framework, Li et al.'s genome-scale metabolic modeling,
    Xiong/Oh's BioMapAI), positioning Part~V within that landscape. Honest
    assessment of contributions and limitations (no computational implementation,
    parameter uncertainty, validation gap).
    \textit{Motivated by:} Need for transparent novelty assessment; prior work
    survey absent from original Part~V
    draft.~\cite{broderick2010cytokine,morris2013neuroimmune,phair2019ido,li2025systems,xiong2025biomap}.

  \item \textbf{Chapter 27 --- Energy Metabolism Models}: Major expansion with
    formal ODE systems, parameter tables, and sensitivity analysis frameworks
    for mitochondrial and metabolic models.

  \item \textbf{Chapter 28 --- Immune System Models}: Major expansion with
    formal immune cell population dynamics, B cell depletion modeling (rituximab
    vs.\ plasma cell persistence), and cytokine network models.

  \item \textbf{Chapter 29 --- Neuroendocrine Models}: Major expansion with
    HPA axis feedback models, autonomic regulation frameworks, and
    neurotransmitter dynamics.

  \item \textbf{Chapter 30 --- Integrated Systems}: Major expansion describing
    multi-system coupling architecture, cross-system feedback loops, and
    emergent properties of the integrated model.

  \item \textbf{Chapter 31 --- Temporal Evolution}: Major expansion with
    disease trajectory modeling, phase transition analysis, and the
    energy ratchet mechanism formalized as a mathematical model.

  \item \textbf{Chapter 32 --- Predictive Applications}: Major expansion
    covering clinical decision support, treatment response prediction,
    and biomarker-guided intervention timing.

  \item \textbf{Chapters 26--31 --- Model Application Guides}: Added structured
    ``Model Application Guide'' sections to each modeling chapter
    (Chs.~27--31), specifying for every model group: measurements required,
    how to use the model, worked examples with realistic parameter values,
    inter-model dependencies, falsification criteria, and clinical
    implications. Added model coverage justification section to Ch.~26
    with Part~II$\to$Part~V mapping table, selection rationale, and explicit
    enumeration of unmodeled domains.
    \textit{Motivated by:} Gap analysis identifying that models lacked
    practical application guidance, falsification criteria, and coverage
    justification.

  \item \textbf{TikZ figures}: Added 10~new TikZ illustrations: baroreflex
    model, clinical decision pipeline, disease trajectories, energy metabolism
    model, energy ratchet, HPA axis model, immune cell dynamics, integrated
    systems architecture, modeling framework overview, and tryptophan branching
    model.

  \item \textbf{Homocysteine-methylation-mitochondrial dysfunction pathway}
    (Chapter~6, Chapter~12, Chapter~14, Chapter~16, Chapter~20): Integrated
    evidence linking MTHFR polymorphisms, elevated CSF homocysteine, and
    mitochondrial ETC inhibition. Added hypothesis in Ch.~6 (hyp:homocysteine-etc-disruption)
    connecting homocysteine-mediated mitochondrial damage to ME/CFS fatigue;
    observation in Ch.~12 documenting MTHFR C677T as a candidate predisposing
    variant; expanded Ch.~20 biomarker section with CSF homocysteine as a
    candidate CNS-compartment biomarker; added speculation in Ch.~14
    (spec:methyl-assault-onset) proposing post-viral methyl-group depletion as
    an ME/CFS onset mechanism; cautious treatment evidence integration in Ch.~16
    with warning environment on B12/folate RCT absence; comprehensive annotated
    bibliography section in Appendix~H with 6~papers.
    \textit{Motivated by:} Mechanistic chain (viral exposure $\to$ methylation
    depletion $\to$ homocysteine elevation $\to$ ETC inhibition $\to$ fatigue) is
    plausible and supported across multiple literature domains, but lacks direct
    ME/CFS patient validation. Integration using appropriate evidence hierarchy
    (hypothesis for unconfirmed mechanisms, observation for small clinical
    studies, speculation for post-viral hypothesis) maintains precision while
    opening integration pathway for future confirmatory studies~\cite{Kaplan2020,Regland1997,Regland2015,vanCampen2019,Zarembska2023,McCaddon2021,Liao2021,Maksoud2023}.

\end{itemize}

\subsection*{Appendices and Bibliography}

\begin{itemize}

  \item \textbf{Appendix A --- Glossary}: Expanded from stub to comprehensive
    alphabetical glossary of medical, biochemical, immunological, and
    statistical terms used throughout the document.

  \item \textbf{Appendix C --- Diagnostic Tools}: Substantially expanded with
    detailed descriptions of diagnostic instruments and assessment tools.

  \item \textbf{Appendix E --- Mathematical Model Details}: Expanded from stub
    to complete mathematical specifications for Part~V models, including full
    ODE systems, parameter tables, derivations, numerical methods, and
    sensitivity analysis procedures.

  \item \textbf{Appendix F --- Supplement Guide}: Substantially expanded with
    updated supplement protocols and dosage information.

  \item \textbf{Appendix H --- Annotated Bibliography}: Major expansion providing
    comprehensive annotated bibliography of ME/CFS literature organized by
    research domain.

  \item \textbf{Appendix I --- Research Registry}: New appendix cataloguing
    ongoing and planned ME/CFS studies.

  \item \textbf{Bibliography}: Added approximately 200~new references covering
    recent ME/CFS research across pathophysiology, treatment, and modeling.

\end{itemize}

% Changelog synchronized up to commit 704a55c (2026-03-07).
% Next changelog update should cover commits after 704a55c.

\section*{Version 4 (2026-02-18)}

Fourth publicly released version of the document. Established the full structure
across five parts: clinical overview, pathophysiology, treatment, research, and
mathematical modelling.

% --- Template for future versions ---
% \section*{Version N (YYYY-MM-DD)}
% \begin{itemize}
%   \item \textbf{Chapter X — Title}: What changed.
%     \textit{Motivated by:} Why (new evidence / error correction / reviewer feedback).
%     Key sources: \cite{key1,key2}.
% \end{itemize}

