
\subsection{Références primaires citées}

\begin{enumerate}
\item \textbf{Bateman L et al. (2021)} -- ME/CFS: Essentials of Diagnosis and Management. Mayo Clinic Proceedings. Recommandations traitement US ME/CFS Clinician Coalition. [DOI: 10.1016/j.mayocp.2021.07.004]

\item \textbf{Taub PR et al. (2021)} -- Randomized Trial of Iva-bradine in Patients With Hyper-adrenergic POTS. Journal of the American College of Cardiology (JACC). [DOI: 10.1016/j.jacc.2020.12.029]

\item \textbf{Raj SR et al. (2009)} -- Propranolol Decreases Tachycardia and Improves Symptoms in the Postural Tachy\-cardia Syndrome: Less Is More. \textit{Circulation}. [\href{https://doi.org/10.1161/CIRCULATIONAHA.108.846501}{DOI}]

\item \textbf{Raj SR et al. (2005)} -- Renin-aldosterone paradox and perturbed blood volume regulation underlying POTS. Circulation. [DOI: 10.1161/01.CIR.0000160356.97313.5d]

\item \textbf{Freitas J et al. (2000)} -- Clinical improvement in patients with orthostatic intolerance after treatment with bisoprolol and fludrocortisone. Clinical Autonomic Research. [PMID: 11198485]

\item \textbf{Polo O et al. (2019)} -- Low-dose naltrexone in the treatment of ME/CFS. Fatigue: Biomedicine, Health \& Behavior. [DOI: 10.1080/21641846.2019.1692770]

\item \textbf{Bolton MJ et al. (2020)} -- Low-Dose Naltrexone as a Treatment for Chronic Fatigue Syndrome. BMJ Case Reports.

\item \textbf{Cabanas H et al. (2021)} -- LDN restores TRPM3 ion channel function in natural killer cells. Frontiers in Immunology. [DOI: 10.3389/fimmu.2021.687806]

\item \textbf{Hurwitz BE et al. (2010)} -- Chronic fatigue syndrome: illness severity, sedentary lifestyle, blood volume and evidence of diminished cardiac function. Clinical Science.

\item \textbf{Stock JM et al. (2022)} -- Dietary sodium and health: how much is too much for those with orthostatic disorders? Autonomic Neuroscience.

\item \textbf{Lerner AM et al. (2002)} -- Valacyclovir Treatment in Epstein-Barr Virus Subset of Chronic Fatigue Syndrome: Thirty-Six Months Follow-up. \textit{In Vivo}. [PMID: 11878721]

\item \textbf{Lerner AM et al. (2007)} -- A Small, Randomized, Placebo-Controlled Trial of the Use of Antiviral Therapy for Patients with Chronic Fatigue Syndrome. \textit{Clinical Infectious Diseases}. [DOI: 10.1086/513033]

\item \textbf{Lerner AM et al. (2010)} -- Treatment of Chronic Fatigue Syndrome (CFS) with Antivirals: Evidence Supporting a Viral Etiology for CFS. \textit{Journal of Chronic Fatigue Syndrome}. [DOI: 10.1300/J092v09n04\_03]

\item \textbf{Montoya JG et al. (2013)} -- Randomized Clinical Trial to Evaluate the Efficacy and Safety of Valganciclovir in a Subset of Patients with Chronic Fatigue Syndrome. \textit{Journal of Medical Virology}. [DOI: 10.1002/jmv.23713]

\item \textbf{Goldstein JA (1986)} -- Cimetidine, ranitidine, and Epstein-Barr virus infection. \textit{Annals of Internal Medicine}. [DOI: 10.7326/0003-4819-105-1-139\_2] [PMID: 3013060]

\item \textbf{Stuijt R et al. (2026)} -- Use of cimetidine to enhance systemic acyclovir concentrations in patients with ineffective suppressive therapy for recurring herpes simplex virus infections: A novel purpose for an old drug. \textit{British Journal of Clinical Pharmacology}. [DOI: 10.1002/bcp.70313]

\item \textbf{Simons FER et al. (2019)} -- Immunomodulatory properties of cimetidine: Its therapeutic potentials for treatment of immune-related diseases. \textit{International Immunopharmacology}. [DOI: 10.1016/j.intimp.2018.12.061] [PMID: 30802678]
\end{enumerate}

\subsection{Revues systématiques référencées}

\begin{enumerate}
\setcounter{enumi}{10}
\item \textbf{Oral medications for POTS: systematic review (2024)} -- Frontiers in Neurology. [DOI: 10.3389/fneur.2024.1515486]

\item \textbf{Systematic literature review: treatment of POTS (2025)} -- Clinical Autonomic Research. [DOI: 10.1007/s10286-025-01172-2]

\item \textbf{Evidence for treatments for POTS: systematic review of randomized trials (2025)} -- American Heart Journal Plus. [DOI: 10.1016/j.ahjo.2025.100933]
\end{enumerate}
