% FILE: Pediatric ME/CFS treatment — school-attending and ambulatory pediatric cases, early intervention, educational accommodations
\chapter{Pediatric ME/CFS: School-Attending and Ambulatory Cases}
\label{ch:pediatric-ambulatory}

Children and adolescents with mild to moderate ME/CFS who retain capacity for school attendance---even partial attendance---represent a critical population for early intervention. The substantially better prognosis in pediatric ME/CFS (54--94\% improvement or recovery versus $\leq$22\% in adults) creates both opportunity and urgency: appropriate early management may preserve the window for recovery, while inappropriate management (particularly graded exercise therapy) risks progression to severe, potentially irreversible disease~\cite{Rowe2017pediatric, CDC2024pediatric}.

This chapter addresses ambulatory pediatric patients who can attend school at least part-time. For children who are housebound or bedbound, see Chapter~\ref{ch:pediatric-severe}. For adult mild-moderate management, see Chapter~\ref{ch:action-mild-moderate}.

\section{The Early Intervention Imperative}
\label{sec:ped-early-intervention}

Early intervention in pediatric ME/CFS is not merely beneficial---it may be critical for preserving recovery potential. The window of opportunity that distinguishes pediatric from adult prognosis appears to narrow with illness duration and repeated severe crashes.

\subsection{Why Early Intervention is More Critical in Pediatrics}

Several factors make early intervention uniquely important in pediatric ME/CFS:

\begin{enumerate}
    \item \textbf{Better prognosis creates higher stakes}: The 54--94\% improvement/recovery rate in pediatric ME/CFS (versus $\leq$22\% in adults) means there is more to preserve. Inappropriate management that converts a child from the ``likely to recover'' group to the ``chronic severe'' group represents a catastrophic outcome that might have been prevented.

    \item \textbf{Critical developmental window}: Lost school years during childhood and adolescence have consequences beyond immediate disability. Missed education, interrupted social development, and disrupted identity formation during formative years create cumulative deficits that persist even after physical recovery.

    \item \textbf{Preventing progression to severe disease}: Many severely affected adults report that their ME/CFS began as mild-moderate disease that worsened due to continued overexertion, often encouraged by well-meaning but misinformed healthcare providers, teachers, and parents. Early recognition and appropriate pacing may prevent this trajectory.

    \item \textbf{Evidence that early intervention improves outcomes}: Studies consistently show that shorter diagnostic delay correlates with better prognosis. Patients diagnosed and managed appropriately within the first year have better long-term outcomes than those who struggle for years before receiving correct diagnosis and treatment~\cite{Rowe2019pediatric}.

    \item \textbf{Developing nervous system plasticity}: The pediatric nervous system and immune system retain plasticity that may facilitate recovery---but this plasticity may be damaged by repeated severe crashes or inappropriate treatments. Preserving this biological capacity requires appropriate management from the earliest stages.
\end{enumerate}

\subsection{The Cost of Delayed or Inappropriate Intervention}

Conversely, delayed or inappropriate intervention carries substantial risks:

\begin{itemize}
    \item \textbf{Progression to severe disease}: Children pushed to maintain normal activity levels despite symptoms frequently progress from mild to moderate to severe ME/CFS over months to years
    \item \textbf{Cumulative crash damage}: Each severe crash may cause partially irreversible damage; the ``crash limit rule'' from patient communities suggests tolerance for severe crashes is limited. The crash dose-response framework (Chapter~\ref{ch:action-mild-moderate}, \S\ref{subsubsec:crash-dose-response}) explains why large crashes cause disproportionate, irreversible harm through ATP depletion thresholds, mitochondrial turnover limits, inflammatory cascade intensity, and epigenetic locking mechanisms.
    \item \textbf{Lost recovery potential}: The window for pediatric recovery may close with prolonged illness duration, converting a recoverable case into a chronic condition. The vicious cycle recruitment cascade (Chapter~\ref{ch:core-symptoms}, \S\ref{sec:pem}) shows how initially single-cycle dysfunction progressively recruits additional cycles, reducing escapability over time.
    \item \textbf{Educational derailment}: Each semester of academic struggle or failure compounds into long-term educational deficits
    \item \textbf{Psychological harm}: Repeated experiences of pushing through symptoms, being disbelieved, and watching function decline create lasting psychological trauma
\end{itemize}

\begin{keypoint}[The Urgency of Correct Early Management]
Every month of delayed diagnosis, every crash caused by inappropriate pressure to exercise, and every semester of forced school attendance without accommodations potentially reduces the likelihood of full recovery. The pediatric advantage is not automatic---it must be preserved through appropriate management from the earliest stages of illness.
\end{keypoint}

\section{Subtype Assessment for Ambulatory Pediatric Cases}
\label{sec:ped-ambulatory-subtype}

Early identification of which systems are primarily affected helps guide treatment prioritization and educational/activity accommodations. Ambulatory pediatric patients (unlike severe cases) retain the potential for recovery across all subtypes, making early appropriate intervention particularly critical.

\begin{recommendation}[Pediatric Subtype Assessment and Early Intervention Prioritization]
\label{rec:ped-ambulatory-subtype}

\textbf{Rationale:} The selective energy dysfunction hypothesis (Chapter~\ref{sec:selective-dysfunction}) proposes that ME/CFS involves different primary compartment dysfunction across subtypes. Identifying your child's dominant subtype guides which interventions to prioritize.

\textbf{Parent/caregiver assessment:}

\begin{enumerate}
    \item \textbf{What improved LEAST with rest?}
    \begin{itemize}
        \item Forgetfulness, poor concentration, difficulty learning new things → \textbf{CNS-Primary}
        \item Dizziness, tachycardia, feeling faint when standing → \textbf{Autonomic-Primary}
        \item Muscle weakness, fatigue, widespread pain → \textbf{Peripheral-Primary}
        \item Multiple symptoms equally severe → \textbf{Global}
    \end{itemize}

    \item \textbf{How does your child perform across different demands?}
    \begin{itemize}
        \item Can walk/play physically but struggles academically → \textbf{Suggests CNS-Primary}
        \item Academic performance okay but can't tolerate standing/activity → \textbf{Suggests Autonomic-Primary}
        \item Both physical and cognitive demands limited by fatigue/pain → \textbf{Suggests Peripheral-Primary}
    \end{itemize}

    \item \textbf{School accommodations needed?}
    \begin{itemize}
        \item Difficulty concentrating, reading, learning new material → \textbf{Suggests CNS-Primary}
        \item Can't attend full days, needs frequent rest periods → \textbf{Suggests multiple systems, likely Autonomic or Global}
        \item Fatigue, weakness, exercise intolerance → \textbf{Suggests Peripheral-Primary or Global}
    \end{itemize}
\end{enumerate}

\textbf{Early intervention prioritization by subtype:}

\begin{description}
    \item[Subtype A (CNS-Primary):] Cognitive dysfunction dominates
    \begin{itemize}
        \item \textbf{Priority 1}: Cognitive support through education accommodations; learning aids; shortened school days to maximize cognitive capacity during peak hours
        \item \textbf{Priority 2}: Sleep optimization (crucial for pediatric brain development and cognitive function recovery)
        \item \textbf{Priority 3}: Gentle nutritional support; omega-3 supplementation for brain health
        \item \textbf{School accommodation}: Extended time on tests, preferential seating near teacher (reduces cognitive load of filtering other conversations), written instructions rather than verbal (reduces processing demand)
    \end{itemize}

    \item[Subtype B (Autonomic-Primary):] Orthostatic intolerance dominates
    \begin{itemize}
        \item \textbf{Priority 1}: Hydration protocol (2.5--3 liters/day for teens); salt loading if medically appropriate
        \item \textbf{Priority 2}: Compression garments (age-appropriate sizing); frequent position changes during school day
        \item \textbf{Priority 3}: Medications if non-pharmacological measures insufficient (midodrine is pediatric-approved for POTS)
        \item \textbf{School accommodation}: Preferential seating (lying down or semi-reclined if possible), permission to move around class, frequent water breaks, cooler classroom if heat-triggered symptoms
    \end{itemize}

    \item[Subtype C (Peripheral-Primary):] Muscle weakness/fatigue dominates
    \begin{itemize}
        \item \textbf{Priority 1}: Activity pacing; strict adherence to energy envelope
        \item \textbf{Priority 2}: Nutritional optimization (adequate protein for muscle recovery, calories for growth)
        \item \textbf{Priority 3}: Gentle mitochondrial support (CoQ10, carnitine—pediatric doses)
        \item \textbf{School accommodation}: Reduced class load, exemption from PE, opportunity for rest periods between subjects
    \end{itemize}

    \item[Subtype D (Global):] Multi-system involvement
    \begin{itemize}
        \item \textbf{Approach}: Comprehensive accommodations addressing all domains
        \item \textbf{Strategy}: Start with hydration + sleep + pacing (foundational for all subtypes), then add domain-specific treatments
        \item \textbf{School accommodation}: Combination of measures—half-day or part-time attendance, reduced academic load, frequent rest periods, accommodations from all subtype categories as needed
        \item \textbf{Critical}: Do NOT push your child to maintain full schedule; early appropriate accommodation prevents progression to severe disease
    \end{itemize}
\end{description}

\textbf{Evidence level}: Plausible (subtype framework from Chapter~\ref{sec:selective-dysfunction}, adapted for pediatrics with school considerations)

\textbf{Key principle}: Subtype assessment informs prioritization but should NOT delay implementation of foundational interventions (pacing, sleep, hydration) while optimizing.

\end{recommendation}

\section{Diagnosis and Assessment}
\label{sec:ped-diagnosis}

Accurate diagnosis is the foundation of appropriate treatment. Pediatric ME/CFS diagnosis requires modifications to adult criteria to account for developmental context.

\subsection{Pediatric Diagnostic Criteria Modifications}
\label{subsec:ped-diagnostic-criteria}

The core diagnostic criteria for ME/CFS (substantial reduction in activity, post-exertional malaise, unrefreshing sleep, plus cognitive dysfunction or orthostatic intolerance) apply to pediatrics with the following modifications~\cite{Rowe2017pediatric}:

\begin{itemize}
    \item \textbf{Duration threshold}: Standard criteria require 6 months of symptoms. For pediatric patients with severe presentation, 3 months may be sufficient for provisional diagnosis and treatment initiation. Waiting 6 months while a child deteriorates is not clinically justified.

    \item \textbf{Activity reduction assessment}: Adults are assessed by work capacity; children should be assessed by school attendance, extracurricular participation, social activities, and self-care---all compared to pre-illness baseline, not to peers. A child who previously played sports, maintained friendships, and earned good grades but now struggles with half-day school attendance has experienced substantial activity reduction regardless of how this compares to ``average.''

    \item \textbf{Symptom reporting}: Younger children may have difficulty articulating symptoms like ``brain fog'' or ``post-exertional malaise.'' Parent observation and age-appropriate questioning are essential. Ask about specific situations: ``Can you play as long as you used to?'' ``How do you feel the day after a busy day?'' ``Do you get tired from things that didn't used to tire you?''

    \item \textbf{Post-exertional malaise recognition}: PEM in children may manifest as increased irritability, tearfulness, pain, or flu-like symptoms 12--72 hours after exertion, rather than the adult description of ``crashing.'' The pattern of delayed worsening after activity is diagnostic, regardless of the specific symptom expression.

    \item \textbf{Orthostatic intolerance prevalence}: OI affects 70--90\% of pediatric ME/CFS patients---higher than adults~\cite{Rowe2017pediatric}. Standing tests (10-minute stand test or tilt table testing) should be part of evaluation. Many children with unexplained fatigue have unrecognized POTS or orthostatic hypotension.
\end{itemize}

\subsection{Distinguishing ME/CFS from Differential Diagnoses}
\label{subsec:ped-differentials}

Several conditions can mimic or overlap with pediatric ME/CFS. Accurate diagnosis requires distinguishing ME/CFS from these alternatives while recognizing that some conditions (particularly POTS) frequently co-occur.

\subsubsection{School Avoidance/School Refusal}

Perhaps the most common misdiagnosis for pediatric ME/CFS is ``school avoidance'' or ``school refusal''---the assumption that the child's symptoms represent psychological resistance to school rather than physical illness.

\textbf{Key distinguishing features:}

\begin{itemize}
    \item \textbf{PEM pattern}: Children with ME/CFS crash AFTER exertion, not before. A child who attends school Monday, crashes Tuesday-Wednesday, and improves by Friday demonstrates PEM. A child who refuses school Monday morning but is active in the afternoon may have school avoidance.

    \item \textbf{Weekend/vacation pattern}: ME/CFS does not resolve on weekends or vacations. Children with true ME/CFS remain symptomatic during breaks from school. School avoidance typically improves when school is not in session.

    \item \textbf{Desire to participate}: Children with ME/CFS typically WANT to attend school and participate in activities---they are limited by symptoms. Children with school avoidance may express reluctance or resistance beyond what symptoms would explain.

    \item \textbf{Physical findings}: Objective findings (positive tilt table test, documented tachycardia, measurable orthostatic hypotension) support ME/CFS diagnosis. School avoidance does not produce these findings.

    \item \textbf{Response to activity}: If a child can sustain hours of enjoyable activity (gaming, socializing) but cannot tolerate equivalent school time, this suggests different mechanisms may be at play. However, note that cognitive and physical exertion differ, and school may be more demanding than leisure activities even at similar durations.
\end{itemize}

\begin{practicalwarning}[The Danger of Misdiagnosis as School Avoidance]
Misdiagnosing ME/CFS as school avoidance is not merely inconvenient---it is potentially dangerous. Children labeled as school-avoidant are often subjected to forced school attendance, psychological interventions aimed at overcoming ``avoidance,'' and graded exposure programs that push increasing activity. For children with actual ME/CFS, these interventions cause harm: physical deterioration, progression to severe disease, and psychological trauma from being disbelieved. Always take reported symptoms seriously and look for objective evidence of ME/CFS before concluding that symptoms are psychologically mediated.
\end{practicalwarning}

\subsubsection{Post-Infectious Mononucleosis}

Epstein-Barr virus (EBV) mononucleosis is a common trigger for ME/CFS in adolescents. Post-infectious fatigue lasting weeks to months is normal after mono; ME/CFS is diagnosed when symptoms persist beyond expected recovery and include characteristic features (PEM, OI, cognitive dysfunction).

\begin{itemize}
    \item \textbf{Timeline}: Most adolescents recover from mono within 2--4 months. Fatigue persisting beyond 6 months with PEM suggests ME/CFS.
    \item \textbf{Management during early post-mono period}: Appropriate rest and pacing during the post-mono period may reduce risk of developing chronic ME/CFS. Pushing early return to full activity after mono is inadvisable.
    \item \textbf{EBV reactivation}: Some ME/CFS patients have evidence of chronic EBV reactivation; this represents ongoing viral contribution to ME/CFS rather than simple post-infectious fatigue.
\end{itemize}

\subsubsection{Growth-Related Fatigue}

Puberty is associated with increased sleep need and may cause temporary fatigue. Menstruation can cause cyclic fatigue from blood loss and iron deficiency.

\begin{itemize}
    \item \textbf{Growth fatigue}: Responds to adequate sleep; does not cause PEM; does not worsen progressively
    \item \textbf{Iron deficiency}: Check ferritin in all adolescent females with fatigue; supplement if ferritin $<$30 ng/mL (some experts recommend $<$50)
    \item \textbf{Cyclic pattern}: Fatigue clearly worse around menstruation suggests hormonal or iron-related cause; ME/CFS fatigue is present throughout cycle (though may worsen perimenstrually)
\end{itemize}

\subsubsection{POTS and Orthostatic Intolerance}

POTS (Postural Orthostatic Tachycardia Syndrome) and ME/CFS frequently co-occur. Some patients have ``pure'' POTS without ME/CFS; others have ME/CFS with prominent POTS; many have overlapping presentations.

\begin{itemize}
    \item \textbf{POTS without ME/CFS}: Symptoms primarily orthostatic (worse with standing, improved with lying); no significant PEM; exercise may actually improve symptoms (cardiovascular reconditioning)
    \item \textbf{ME/CFS with POTS}: Both orthostatic symptoms AND PEM; exercise worsens overall condition even if brief standing tolerance improves with treatment
    \item \textbf{Management differs}: Pure POTS may benefit from graded cardiovascular reconditioning; ME/CFS with POTS requires pacing and will worsen with exercise programs. Accurate diagnosis is essential.
\end{itemize}

\subsubsection{Depression and Anxiety}

Depression and anxiety can cause fatigue and are common comorbidities in ME/CFS. They do not cause PEM.

\begin{itemize}
    \item \textbf{PEM distinguishes}: Depression causes persistent fatigue that may improve with pleasurable activity. ME/CFS causes fatigue that WORSENS after any sustained activity, pleasurable or not.
    \item \textbf{Comorbidity is common}: Many children with ME/CFS develop secondary depression from the experience of chronic illness, loss of function, and being disbelieved. Treating comorbid depression is appropriate but does not address ME/CFS.
    \item \textbf{Antidepressants for ME/CFS}: May help comorbid depression and certain symptoms (sleep, pain) but do not treat the underlying ME/CFS and do not cure the illness.
\end{itemize}

\section{School Accommodation Framework}
\label{sec:ped-school-accommodations}

Educational accommodations are essential for children with ME/CFS to maintain academic progress while managing their illness. In the United States, two primary frameworks exist: IEP (Individualized Education Program) under IDEA, and 504 Plans under Section 504 of the Rehabilitation Act.

\subsection{IEP vs.\ 504 Plan Decision Tree}
\label{subsec:iep-vs-504}

Understanding the distinction between IEP and 504 Plans is essential for securing appropriate accommodations.

\subsubsection{Section 504 Plans}

\textbf{Legal basis}: Section 504 of the Rehabilitation Act of 1973 prohibits discrimination against individuals with disabilities in programs receiving federal funding, including public schools.

\textbf{Eligibility}: Any student with a physical or mental impairment that substantially limits one or more major life activities. ME/CFS clearly qualifies---it limits learning, concentrating, thinking, and physical activity.

\textbf{What 504 provides}:
\begin{itemize}
    \item Accommodations to ensure equal access to education
    \item Modifications to the regular education program
    \item Does NOT provide specialized instruction or related services
\end{itemize}

\textbf{Best for}: Students who can access the general curriculum with accommodations (modified schedule, testing modifications, etc.) but do not need fundamentally different instruction.

\textbf{Process}:
\begin{enumerate}
    \item Parent requests 504 evaluation in writing
    \item School evaluates (may use existing medical documentation)
    \item Team (including parents) develops 504 Plan
    \item Plan reviewed annually (or as needed)
\end{enumerate}

\subsubsection{IEP (Individualized Education Program)}

\textbf{Legal basis}: Individuals with Disabilities Education Act (IDEA) entitles children with qualifying disabilities to Free Appropriate Public Education (FAPE).

\textbf{Eligibility}: Student must have a disability in one of 13 categories AND require special education services. ME/CFS may qualify under ``Other Health Impairment'' (OHI)---a chronic or acute health problem that adversely affects educational performance.

\textbf{What IEP provides}:
\begin{itemize}
    \item Specialized instruction tailored to student's needs
    \item Related services (speech therapy, occupational therapy, counseling)
    \item Measurable annual goals
    \item Transition planning (for older students)
    \item More procedural protections than 504
\end{itemize}

\textbf{Best for}: Students who need specialized instruction, not just accommodations. May be appropriate for severely affected ME/CFS students who need fundamental modifications to curriculum or instruction method.

\textbf{Process}:
\begin{enumerate}
    \item Parent requests evaluation in writing
    \item School conducts comprehensive evaluation (60-day timeline in most states)
    \item Eligibility determination meeting
    \item If eligible, IEP team develops plan
    \item Annual IEP review; reevaluation every 3 years
\end{enumerate}

\subsubsection{Decision Framework}

\begin{protocol*}[504 vs.\ IEP Decision Framework]

\textbf{Start with 504 if}:
\begin{itemize}
    \item Student can participate in general education curriculum
    \item Primary needs are accommodations (schedule, testing, attendance flexibility)
    \item Student does not require specialized instruction
    \item Faster implementation needed (504 process is typically quicker)
\end{itemize}

\textbf{Pursue IEP if}:
\begin{itemize}
    \item Student cannot access general curriculum even with accommodations
    \item Specialized instruction is needed (modified curriculum, alternative teaching methods)
    \item Related services are needed (counseling, OT for cognitive strategies)
    \item Student is failing despite 504 accommodations
    \item Stronger legal protections are desired
\end{itemize}

\textbf{Consider both}:
\begin{itemize}
    \item Start with 504 for immediate accommodations
    \item Pursue IEP evaluation simultaneously if specialized services may be needed
    \item IEP supersedes 504 (you don't need both)
\end{itemize}
\end{protocol*}

\subsection{Specific Accommodations by Symptom}
\label{subsec:accommodations-by-symptom}

Accommodations should be tailored to the specific symptoms limiting each student's educational participation.

\subsubsection{Accommodations for Cognitive Dysfunction}

Brain fog, difficulty concentrating, and memory problems are nearly universal in ME/CFS and significantly impact academic performance.

\begin{itemize}
    \item \textbf{Extended test time}: 50--100\% additional time; may need testing over multiple sessions
    \item \textbf{Reduced homework}: Homework should not exceed what can be completed in 50\% of standard expected time; quality over quantity
    \item \textbf{Note-taking assistance}: Peer note-taker, teacher-provided notes, or audio recording of lectures
    \item \textbf{Alternative assignment formats}: Oral presentations instead of written papers; multiple-choice instead of essays; projects over time instead of timed tests
    \item \textbf{Preferential seating}: Front of room reduces distraction; near door allows quiet exit if overwhelmed
    \item \textbf{Reduced course load}: Fewer classes per semester; prioritize core requirements
    \item \textbf{Calculator/reference sheet use}: Reduce cognitive load of memorization
    \item \textbf{Quiet testing location}: Separate room reduces sensory demands during high-stakes testing
\end{itemize}

\subsubsection{Accommodations for Orthostatic Intolerance}

OI symptoms (dizziness, lightheadedness, difficulty standing) require specific environmental modifications.

\begin{itemize}
    \item \textbf{Unlimited water and snacks}: Student should be permitted water bottle and salty snacks at all times; bathroom access without restriction
    \item \textbf{Sitting privileges}: Permission to sit during assemblies, pledge, or other standing activities
    \item \textbf{Elevator access}: If school has multiple floors; stair climbing is particularly triggering for OI
    \item \textbf{Climate control}: Access to cooler environments; heat worsens OI
    \item \textbf{Rest breaks}: Permission to lie down in nurse's office when symptomatic
    \item \textbf{Compression garment permission}: Student may need to wear compression stockings or shorts; should not be restricted by dress code
    \item \textbf{Late arrival}: If mornings are worst (common), late start may improve function
\end{itemize}

\subsubsection{Accommodations for Post-Exertional Malaise}

PEM requires flexibility in attendance and assignment deadlines.

\begin{itemize}
    \item \textbf{Excused absences for crashes}: PEM-related absences should not count against attendance requirements; documentation of ME/CFS diagnosis should suffice without requiring repeated physician notes
    \item \textbf{Flexible deadlines}: Extensions on assignments following crashes
    \item \textbf{Rest breaks during day}: Permission to leave class and rest when approaching energy limits
    \item \textbf{Reduced physical education}: Exemption from standard PE; may substitute with pacing-appropriate activity (gentle stretching, walking to tolerance) or academic alternative
    \item \textbf{Split attendance}: Part-day at school, part-day home instruction; maintains school enrollment and peer connection while reducing daily demands
    \item \textbf{Modified schedule}: Attend every other day; attend mornings only; attend for testing only
\end{itemize}

\subsubsection{Accommodations for Sensory Sensitivity}

Light, sound, and other sensory inputs may be overwhelming.

\begin{itemize}
    \item \textbf{Sunglasses indoors}: If fluorescent lights are triggering
    \item \textbf{Noise-canceling headphones}: During independent work or when overwhelmed
    \item \textbf{Quiet testing location}: Separate room for tests
    \item \textbf{Reduced sensory environment}: Front-row seating away from windows; permission to step out of noisy environments
    \item \textbf{Digital textbooks}: Adjustable font size, screen brightness, text-to-speech options
\end{itemize}

\subsection{Communication Templates}
\label{subsec:school-communication}

Effective advocacy requires clear communication with school personnel.

\subsubsection{Letter Requesting Evaluation}

A written request triggers the school's legal obligation to evaluate.

\begin{quote}
\textit{Dear [Principal/Special Education Director],}

\textit{I am writing to formally request an evaluation of my child, [Name], for [504 Plan/special education services under IDEA]. [Name] has been diagnosed with Myalgic Encephalomyelitis/Chronic Fatigue Syndrome (ME/CFS), a chronic neuroimmune illness that significantly affects [his/her] ability to participate in school.}

\textit{Specifically, [Name] experiences: [list 3--4 key symptoms affecting school---e.g., severe fatigue, difficulty concentrating, post-exertional malaise that causes crashes after school days, orthostatic intolerance].}

\textit{These symptoms have resulted in: [describe educational impact---e.g., missing X days this semester, declining grades, inability to complete homework, difficulty participating in class].}

\textit{I am enclosing medical documentation from [Name]'s physician confirming the ME/CFS diagnosis. Please contact me to schedule an evaluation meeting. I understand the school has [30/60 days depending on state] to complete the evaluation.}

\textit{Thank you for your attention to this request.}

\textit{Sincerely, [Parent Name]}
\end{quote}

\subsubsection{School Nurse Protocol}

Providing the school nurse with clear guidance improves daily symptom management.

\begin{quote}
\textit{[Student Name] has Myalgic Encephalomyelitis/Chronic Fatigue Syndrome (ME/CFS). Please allow the following:}

\textit{1. Rest in nurse's office when symptomatic (lying down with lights dimmed)}

\textit{2. Unlimited water and salty snacks}

\textit{3. Call parent for pickup if: [list specific symptoms warranting dismissal]}

\textit{4. Do NOT encourage ``pushing through'' symptoms---this worsens the condition}

\textit{5. Symptoms often appear 12--72 hours AFTER overexertion (pattern of Monday school attendance, Tuesday-Wednesday crash is typical ME/CFS)}

\textit{Emergency contact: [Parent phone]}

\textit{Physician contact: [Physician name and phone]}
\end{quote}

\subsubsection{Teacher Education Brief}

Teachers who understand ME/CFS provide better support.

\begin{quote}
\textit{About ME/CFS:}

\textit{ME/CFS is a chronic neuroimmune illness---not depression, laziness, or ``just being tired.'' Key features:}

\textit{1. Post-exertional malaise: Activity causes symptom worsening 12--72 hours later. A student who seems fine during class may crash the next day.}

\textit{2. Cognitive dysfunction (``brain fog''): Difficulty concentrating, processing information, and remembering---even when appearing alert.}

\textit{3. Energy limits: Students have a fixed ``energy envelope.'' Exceeding it causes crashes and potentially long-term worsening.}

\textit{What helps: Flexible deadlines, reduced workload, rest breaks, quiet environment, believing the student's reported symptoms.}

\textit{What hurts: Pressure to ``push through,'' skepticism about symptoms, inconsistent accommodation implementation, comparing student to peers.}

\textit{[Student Name]'s specific accommodations: [list from 504/IEP]}
\end{quote}

\section{Pacing for Children and Adolescents}
\label{sec:ped-pacing}

Pacing---staying within the energy envelope to prevent crashes---is the cornerstone of ME/CFS management. For children, pacing must be adapted to developmental stage and implemented with adult support.

\subsection{Age-Appropriate Energy Management}
\label{subsec:ped-energy-management}

\subsubsection{The Energy Envelope Concept for Children}

Explain the energy envelope in age-appropriate terms:

\begin{itemize}
    \item \textbf{For younger children (6--10)}: ``Your body has a battery. When the battery gets low, you feel bad and need to rest. If you use too much battery, you'll feel sick tomorrow. We need to save enough battery each day.''

    \item \textbf{For older children (11--14)}: ``You have a daily energy budget---like a bank account. Every activity costs energy. If you spend more than you have, you go into debt and crash later. We need to figure out your budget and stay within it.''

    \item \textbf{For adolescents (15--18)}: Full energy envelope concept as described in adult chapters---fixed available energy, all activities cost energy, exceeding envelope triggers PEM and potentially long-term worsening.
\end{itemize}

\subsubsection{Activity Tracking for Children}

Monitoring activity and symptoms helps identify the energy envelope:

\begin{itemize}
    \item \textbf{Younger children}: Parent tracks activity and symptoms; simple rating scales (happy face/sad face for symptoms)
    \item \textbf{Older children}: Child participates in tracking with parent support; symptom diary
    \item \textbf{Adolescents}: Self-tracking with apps; parent reviews patterns
\end{itemize}

Track:
\begin{itemize}
    \item Activities each day (school hours, homework time, physical activity, social activity)
    \item Symptom rating (1--10 scale, or simple categories)
    \item PEM episodes (when they occur, what preceded them by 12--72 hours)
\end{itemize}

After 2 weeks, identify: What activity level consistently avoids PEM? That is the energy envelope.

\subsubsection{Heart Rate Monitoring}

Heart rate monitoring provides objective guidance for staying within the aerobic threshold:

\begin{itemize}
    \item \textbf{Calculate threshold}: $(220 - \text{age}) \times 0.55$ to $0.60$ for ME/CFS patients (lower than standard aerobic threshold)
    \item \textbf{Wearable monitors}: Fitness watches or chest straps; many children will engage with wearable technology
    \item \textbf{Gamification}: Some children respond well to ``keeping the number below X'' as a game; apps can provide alerts
    \item \textbf{Ideally}: Cardiopulmonary exercise testing (CPET) provides actual measured anaerobic threshold; standard formulas are approximations
\end{itemize}

\begin{keypoint}[Heart Rate as Training Wheels]
Heart rate monitoring is ``training wheels'' for pacing. Initially, use it to learn what activities and intensities exceed threshold. Over time, children learn to recognize body signals (early fatigue, heart racing, breathlessness) that indicate threshold approach. Eventually, most can pace effectively by feel, using heart rate monitoring to verify during uncertainty.
\end{keypoint}

\subsubsection{Distinguishing Pacing from Laziness}

Parents and teachers may struggle to distinguish appropriate energy conservation from ``laziness.'' Key distinctions:

\begin{itemize}
    \item \textbf{Desire}: Children with ME/CFS typically WANT to do activities but cannot. Lazy children do not want to do activities. If a child is distressed about missing activities, that suggests illness, not laziness.

    \item \textbf{PEM pattern}: Laziness does not cause delayed symptom worsening. If rest is followed by improvement and activity by delayed crashes, that is ME/CFS.

    \item \textbf{Consistency}: ME/CFS limits all sustained activity---even enjoyable activities. Laziness is selective.

    \item \textbf{Physical findings}: Documented tachycardia, orthostatic changes, and other objective findings support illness. Laziness does not produce these.
\end{itemize}

\subsection{Managing the ``Push Through'' Pressure}
\label{subsec:push-through-pressure}

Children with ME/CFS face enormous pressure to exceed their energy envelope---from parents, teachers, peers, healthcare providers, and their own desires. Managing this pressure is essential for preventing progression.

\subsubsection{Academic Expectations}

\begin{itemize}
    \item \textbf{Communicate with school}: Ensure teachers understand that reduced output reflects illness, not lack of effort
    \item \textbf{Prioritize ruthlessly}: If only X hours of schoolwork are possible, focus on core requirements
    \item \textbf{Quality over quantity}: Completing fewer assignments well is better than struggling through everything poorly
    \item \textbf{Long-term perspective}: A year of reduced academics will not ruin a child's future; progression to severe ME/CFS might
    \item \textbf{Alternative paths}: If traditional academic trajectory is impossible, explore alternatives (GED, community college, gap year)
\end{itemize}

\subsubsection{Sports and Physical Activities}

Sports decisions are particularly difficult. Recommendations:

\begin{itemize}
    \item \textbf{Assess realistically}: Can the child participate without triggering PEM? If not, participation is harmful regardless of desire.
    \item \textbf{Modification options}: Can the child participate in reduced capacity (half practices, no games, coaching role)?
    \item \textbf{When to stop}: If any participation triggers crashes, stopping is necessary. This is not failure---it is medical management.
    \item \textbf{Alternative connection}: If team membership is important for social reasons, explore non-physical roles (manager, statistician)
    \item \textbf{Grieving}: Losing sports identity is a significant loss. Allow space for grief while maintaining the boundary.
\end{itemize}

\subsubsection{Peer Pressure and Social Activities}

Adolescents particularly struggle with social pressure to maintain normal activity:

\begin{itemize}
    \item \textbf{Selective participation}: Can attend some events by choosing carefully (one party per month, not every weekend gathering)
    \item \textbf{Modified participation}: Attend early and leave before exhaustion; sit rather than stand; take rest breaks
    \item \textbf{Explain to friends}: Close friends can understand if given accurate information; they can become allies in pacing
    \item \textbf{Alternative socialization}: Low-energy options (movie at home, video calls, texting) maintain connection without physical cost
    \item \textbf{Quality over quantity}: Fewer, shorter, better-paced social interactions preserve relationships and health
\end{itemize}

\subsubsection{Adolescent Identity and Disability Acceptance}

Adolescence is a critical period for identity formation. Developing ME/CFS during this period creates challenges:

\begin{itemize}
    \item \textbf{Disability as identity}: ME/CFS becomes part of identity; this is neither good nor bad, but should be acknowledged
    \item \textbf{Avoid self-blame}: The child did not cause their illness and cannot will themselves better
    \item \textbf{Realistic self-concept}: Learning to assess actual capacity accurately (not overestimating to seem ``normal'' or underestimating to avoid disappointment)
    \item \textbf{Finding meaning}: What activities, relationships, and goals remain possible? Focus on what can be done, not only what is lost
    \item \textbf{Peer support}: Connecting with other adolescents with ME/CFS or chronic illness reduces isolation and provides modeling
\end{itemize}

\section{Medical Management}
\label{sec:ped-ambulatory-medical}

Medical management for ambulatory pediatric ME/CFS addresses the same symptoms as severe disease (Chapter~\ref{ch:pediatric-severe}) but with adjustments for the school-attending context.

\subsection{Early Intervention Window Protocol}
\label{subsec:ped-early-intervention}

Pediatric ME/CFS has exceptional recovery potential---54--94\% of children improve or recover compared to $\leq$22\% of adults. This window of opportunity is not automatic; it must be actively preserved through aggressive early intervention.

\begin{protocol}[Early Intervention Window Preservation]
\label{protocol:ped-early-intervention-window}

\textbf{Aggressive Early Treatment: Don't ``Wait and See''}

The window for pediatric recovery appears to narrow with illness duration and repeated severe crashes. Early intervention is not optional:

\begin{itemize}
    \item \textbf{Start treatment immediately upon diagnosis}: Do not wait months for improvement or to ``see how it goes.'' Appropriate treatment should begin during the first months of illness.
    \item \textbf{Optimize OI treatment first}: This is the most treatable symptom and often improves multiple other symptoms (fatigue, cognitive dysfunction, exercise tolerance). Do not accept ``mild OI'' as a reason to defer treatment.
    \item \textbf{Aggressive sleep optimization}: Sleep dysfunction perpetuates neuroinflammation. Use pharmacological treatment (melatonin, trazodone) immediately if sleep hygiene is insufficient.
    \item \textbf{Infection prevention from day one}: Each infection causes both acute PEM and lasting sensitization increases. Preventing infections from the outset preserves recovery potential.
    \item \textbf{Avoid crashes through strict pacing}: Each crash is not merely a bad day---it is a potential step toward disease progression and lost recovery potential.
\end{itemize}

\textbf{Brain-First Medication Sequence for School-Attending Pediatric Patients}

The goal is improved cognitive function and fatigue tolerance while maintaining school attendance, WITHOUT enabling overexertion:

\begin{enumerate}
    \item \textbf{Orthostatic intolerance treatment} (Step 1): Non-pharmacological measures first; add fludrocortisone or midodrine if needed. Many children experience dramatic cognitive improvement with effective OI management.

    \item \textbf{Sleep optimization} (Step 2): Melatonin first-line; add trazodone or low-dose amitriptyline if melatonin insufficient. Adequate sleep is foundational for all other improvements.

    \item \textbf{Low-dose naltrexone (LDN)} (Step 3, if needed): Consider if cognitive dysfunction or fatigue remain limiting despite OI and sleep optimization. Weight-adjusted dosing (0.1 mg/kg, max 4.5 mg). LDN modulates neuroinflammation without the risks of stimulants.
    \begin{itemize}
        \item Advantages: Good safety profile in pediatrics; addresses underlying neuroinflammatory mechanisms; does not mask symptoms or enable harmful overexertion
        \item Timeline: 2--8 weeks for measurable effect
        \item Monitoring: Track cognitive function and fatigue; look for improvement without stimulant-like side effects
    \end{itemize}

    \item \textbf{Cognitive support alternatives} (Step 4, if LDN insufficient): Before considering stimulants, maximize accommodations (extended time, reduced homework, preferential seating). Many children function adequately academically with supports alone.

    \item \textbf{Stimulants only if necessary} (Step 5, specialist evaluation): If cognitive dysfunction remains severely limiting despite accommodation and above interventions:
    \begin{itemize}
        \item Stimulants (methylphenidate, amphetamine) may improve concentration but carry risks:
        \begin{itemize}
            \item Can mask symptoms and enable overexertion beyond energy envelope
            \item Tachycardia may worsen orthostatic intolerance
            \item Appetite suppression impairs nutrition
            \item Effects wear off, potentially revealing deeper fatigue
        \end{itemize}
        \item If used, must be coupled with strict monitoring to prevent exceeding energy envelope
        \item Alternative: Low-dose modafinil has been used in some adolescents with limited pediatric data
        \item \textbf{Intranasal delivery option}: If cognitive dysfunction persists despite above interventions, discuss intranasal dopamine precursors or other CNS-targeted compounds with a neurologist familiar with ME/CFS. Not standard care but mechanistically rational for BBB dysfunction (Chapter~\ref{sec:selective-dysfunction}, lines 238--257). EXPERIMENTAL.
    \end{itemize}
\end{enumerate}

\textbf{School Accommodation as Treatment}

School accommodations are not mere kindness---they are therapeutic interventions that preserve recovery potential:

\begin{itemize}
    \item \textbf{Reduced hours/flexible schedule}: Part-time attendance is legitimate medical management, not failure to ``push through''
    \item \textbf{Reduced homework}: Homework that exceeds capacity causes crashes and sensitization. Capping homework is medical necessity.
    \item \textbf{Rest breaks}: Permission to leave class and rest when approaching energy limits prevents crashes during the school day
    \item \textbf{Flexible deadlines}: PEM-related absences and the need for deadline extensions are medical accommodations, not academic favoritism
    \item \textbf{Each crash prevented}: Each accommodation that prevents a crash preserves the recovery window
\end{itemize}

\textbf{Infection Prevention Imperative}

In pediatric patients, infection prevention has compounded importance:

\begin{enumerate}
    \item \textbf{Prevents baseline decline}: As in adults, prevents the immediate PEM and symptom worsening from infection
    \item \textbf{Preserves the pediatric recovery window}: Each infection creates lasting sensitization. Preventing infections preserves the biological conditions for recovery.
    \item \textbf{Maintains school attendance continuity}: Post-infection crashes often cause absences that snowball into educational deficits. Infection prevention maintains educational continuity.
    \item \textbf{Protects developmental trajectory}: Each crash during childhood has developmental consequences beyond the immediate functional decline.
\end{enumerate}

Infection prevention strategies: N95/FFP2 masking during viral seasons, prompt antiviral treatment if infection occurs, consideration of prophylaxis during high-exposure periods (exam weeks, high-transmission seasons), and when possible, isolation of child from sick household members.

\end{protocol}

\subsection{Orthostatic Intolerance (First-Line)}
\label{subsec:ped-ambulatory-oi}

OI is the most treatable symptom domain in pediatric ME/CFS. Many children experience significant improvement in fatigue, cognitive function, and overall wellbeing with effective OI management.

\subsubsection{Non-Pharmacological Measures}

Start with non-pharmacological interventions (see Protocol~\ref{protocol:ped-oi-nonpharm} in Chapter~\ref{ch:pediatric-severe} for details):

\begin{itemize}
    \item \textbf{Hydration}: Age-appropriate targets (2--3 liters/day for adolescents)
    \item \textbf{Salt}: 3--8 grams sodium/day depending on weight
    \item \textbf{Compression}: Waist-high compression garments if tolerated
    \item \textbf{Positioning}: Avoid prolonged standing; sit when possible; leg crossing/tensing
\end{itemize}

\textbf{School accommodations for OI}: Ensure the 504/IEP includes water bottle access, bathroom access, sitting privileges, and late arrival if mornings are worst.

\subsubsection{Compression Garments for School-Attending Children with Orthostatic Intolerance}

\begin{recommendation}[Medical-Grade Compression Stockings for Pediatric OI]
\label{rec:compression-ped-ambulatory}

\textbf{Mechanism:} Compression reduces autonomic load by maintaining venous pressure, decreasing sympathetic activation required for orthostatic compensation (Chapter~\ref{sec:selective-dysfunction}, lines 609--626).

\textbf{Pediatric Implementation:}

\begin{enumerate}
    \item \textbf{Compression class and fit}:
    \begin{itemize}
        \item \textbf{Class II (20--30 mmHg)}: Standard for pediatric POTS and OI
        \item \textbf{Waist-high or thigh-high}: More effective for OI than knee-high
        \item \textbf{Proper fitting}: Measure leg circumference; incorrect sizing loses effectiveness
        \item \textbf{Material}: Merino wool or medical-grade synthetic (breathable for school day wear)
    \end{itemize}

    \item \textbf{School-day wearing schedule}:
    \begin{itemize}
        \item \textbf{During school hours}: Wear throughout day (improved alertness, reduced afternoon crashes)
        \item \textbf{Before/after school}: Optional depending on activity level
        \item \textbf{At home}: Remove during recumbent rest or sleep
        \item \textbf{Total daily wear}: 6--10 hours typical
    \end{itemize}

    \item \textbf{Expected benefits}:
    \begin{itemize}
        \item Reduced tachycardia and dizziness during school day
        \item Improved cognitive clarity and focus
        \item Better school attendance and academic performance
        \item Reduced afternoon crashes during school
        \item May reduce need for medications
    \end{itemize}

    \item \textbf{Practical school considerations}:
    \begin{itemize}
        \item \textbf{Acceptance}: Many children prefer stockings to medications; ease into use
        \item \textbf{Fitting at school}: Some children may prefer removing at lunch/bathroom breaks
        \item \textbf{504 accommodation}: Can request "compression garment breaks" if needed
        \item \textbf{Cost}: \$30--60 per pair; insurance may cover with POTS diagnosis
        \item \textbf{Maintenance}: Replace every 3--6 months as compression decreases
    \end{itemize}

    \item \textbf{Integration with other OI treatments}:
    \begin{itemize}
        \item Essential component of first-line non-pharmacological approach
        \item Synergistic with salt loading and hydration
        \item Can be combined with medications if needed
    \end{itemize}
\end{enumerate}

\textbf{Evidence level}: Moderate (20--30 mmHg compression established for POTS; pediatric ME/CFS data limited)

\textbf{Expected outcomes:} 25--50\% reduction in orthostatic symptoms, improved school function, reduced fatigue during day when combined with salt/hydration protocol.

\textbf{Success predictor:} If child shows improvement in tachycardia or cognitive clarity within 1 week, continue indefinitely.

\end{recommendation}

\subsubsection{Pharmacological Management}

If non-pharmacological measures are insufficient, medications may help (see Protocol~\ref{protocol:ped-oi-pharm} in Chapter~\ref{ch:pediatric-severe} for detailed pediatric dosing):

\begin{itemize}
    \item \textbf{Fludrocortisone}: Starting dose 0.05 mg daily; plasma volume expansion
    \item \textbf{Midodrine}: Starting dose 2.5 mg BID-TID; vasoconstriction (78\% response rate in pediatric POTS)~\cite{Ojha2024pediatricPOTS}
    \item \textbf{Pyridostigmine}: Starting dose 15--30 mg TID; autonomic support
\end{itemize}

\textbf{Monitoring response}: Track symptoms before and after OI treatment. Many children notice improved cognitive function, reduced fatigue, and better exercise tolerance within 1--2 weeks of effective OI management.

\subsection{Mast Cell Activation Syndrome (MCAS) Management}
\label{subsec:ped-ambulatory-mcas}

Mast cell activation affects 30--50\% of ME/CFS patients. In school-attending children, MCAS contributes to post-meal fatigue, brain fog fluctuations, and anxiety-like episodes through histamine and mediator release.

\begin{recommendation}[School-Attending Pediatric MCAS Management]
\label{rec:mcas-ped-ambulatory}

\textbf{Screening:} Consider MCAS trial if child reports:
\begin{itemize}
    \item Food sensitivities or intolerances (especially post-meal fatigue)
    \item Flushing, hives, itching
    \item Reactive to fragrances or chemicals
    \item GI symptoms (bloating, nausea, diarrhea)
    \item Unexplained anxiety or panic-like episodes
\end{itemize}

\textbf{Age-Appropriate Treatment (pediatric dosing):}

\begin{enumerate}
    \item \textbf{H1+H2 Antihistamine Combination}:
    \begin{itemize}
        \item H1: Cetirizine 5 mg morning (younger) or 10 mg (adolescent), OR loratadine 5--10 mg daily
        \item H2: Famotidine 10--20 mg BID (younger children) to 20--40 mg BID (adolescents)
        \item Low-histamine diet (avoid aged cheese, fermented foods, cured meats, leftovers >24 hours)
    \end{itemize}

    \item \textbf{Optional: Quercetin supplement}:
    \begin{itemize}
        \item 250--500 mg daily for younger children
        \item 500--1000 mg daily for adolescents
        \item Natural mast cell stabilizer; can combine with antihistamines
    \end{itemize}

    \item \textbf{4-week trial}: If 30--50\% improvement in energy or cognitive clarity, continue indefinitely

    \item \textbf{Prophylactic Intensification for School Events}:
    \begin{itemize}
        \item \textbf{Before predictable triggers} (exams, presentations, school field trips, social events): Start 24 hours prior
        \begin{itemize}
            \item Increase H1 antihistamine to maximum tolerated dose
            \item Increase H2 to full therapeutic dose
            \item Add quercetin if available
            \item Strict low-histamine diet
        \end{itemize}

        \item \textbf{Activity modification}:
        \begin{itemize}
            \item Reduce non-essential activities day-of event
            \item Ensure adequate rest before and after
            \item Avoid additional cognitive or emotional demands
        \end{itemize}

        \item \textbf{Parent tracking}:
        \begin{itemize}
            \item WITHOUT prophylaxis: "School presentation triggered fatigue crash; needed 2 days recovery"
            \item WITH prophylaxis: "Same presentation with prophylaxis caused minimal fatigue; recovered in 1 day"
            \item Adjust future event planning based on prophylaxis efficacy
        \end{itemize}
    \end{itemize}
\end{enumerate}

\textbf{Evidence level}: Moderate (pediatric MCAS management established; ME/CFS school-event crash mitigation outcomes pending)

\textbf{Expected outcomes:} If MCAS component is significant, 25--50\% reduction in event-related fatigue or cognitive impact.

\end{recommendation}

\subsection{Sleep}
\label{subsec:ped-ambulatory-sleep}

Sleep dysfunction impairs school performance and exacerbates all other symptoms.

\subsubsection{Sleep Hygiene for School-Attending Children}

\begin{itemize}
    \item \textbf{Screen time}: No screens 1--2 hours before bed; blue light filtering if screens are used
    \item \textbf{Homework timing}: Complete homework earlier in evening; avoid cognitive work close to bedtime
    \item \textbf{Consistent schedule}: Same sleep and wake times daily---including weekends (shifts disrupt circadian rhythm)
    \item \textbf{Social media}: Particularly disruptive for adolescents; phones should charge outside bedroom
\end{itemize}

\subsubsection{School Start Time Considerations}

Early school start times conflict with adolescent circadian biology and ME/CFS sleep dysfunction:

\begin{itemize}
    \item \textbf{Request late start}: If school offers late start options, request them
    \item \textbf{Advocate for policy change}: Many schools are shifting to later start times based on adolescent sleep research; parents can advocate
    \item \textbf{504/IEP accommodation}: Late arrival can be an accommodation if early arrival is consistently problematic
\end{itemize}

\subsubsection{Sleep Medications}

If sleep hygiene is insufficient (see Protocol~\ref{protocol:ped-sleep} in Chapter~\ref{ch:pediatric-severe}):

\begin{itemize}
    \item \textbf{Melatonin}: 0.5--5 mg, 30--60 minutes before desired sleep time; first-line
    \item \textbf{Trazodone}: 12.5--50 mg at bedtime if melatonin insufficient
    \item \textbf{Low-dose amitriptyline}: 5--25 mg at bedtime; dual benefit for sleep and pain
\end{itemize}

\subsubsection{Circadian Light Therapy for Sleep-Energy Alignment}

\begin{recommendation}[Bright Light Therapy for Circadian Entrainment in School-Attending Children]
\label{rec:circadian-light-ped-ambulatory}

\textbf{Mechanism:} Morning bright light exposure resets the circadian oscillator, improving alignment between energy availability and school schedule. Many school-attending children with delayed sleep phase struggle with early school start times; morning light therapy can advance sleep phase by 1--2 hours.

\textbf{Pediatric School-Context Protocol:}

\begin{enumerate}
    \item \textbf{Equipment}: 10,000 lux light therapy box (\$25--100). Can be used while eating breakfast or doing morning routine.

    \item \textbf{Timing}: 20--30 minutes immediately upon waking (within 30 minutes). Consistency is critical.

    \item \textbf{Position}: 16--24 inches from child's face, 30° downward angle

    \item \textbf{School integration}: If child struggles with early school start, morning light therapy 30--60 minutes before departure improves alertness and may gradually advance sleep phase by 45 min--2 hours over 2--4 weeks

    \item \textbf{Do NOT use after 3pm} (risks delaying sleep further)
\end{enumerate}

\textbf{Evidence level}: Moderate (circadian disruption documented; light therapy established for circadian disorders and adolescent sleep phase; ME/CFS-circadian-energy RCTs pending)

\textbf{Expected outcomes:}
\begin{itemize}
    \item Earlier sleep onset at night (improved morning school attendance)
    \item Better morning alertness
    \item More consistent daytime energy
    \item Timeline: 2--4 weeks for phase shift
\end{itemize}

\end{recommendation}

\subsubsection{Sleep Spindle Enhancement via Acoustic Stimulation (Low Priority, Experimental)}

\begin{recommendation}[Pink/White Noise for Sleep Architecture in School-Attending Children]
\label{rec:sleep-spindle-ped-ambulatory}

\textbf{Mechanism:} Sleep spindles are brief bursts of brain electrical activity during light sleep that contribute to cognitive consolidation and memory protection. Acoustic stimulation during sleep (pink noise or pink noise burst sequences) can enhance spindle density, potentially improving memory encoding and resilience to cognitive fatigue. In school-attending children with ME/CFS, improved sleep architecture may translate to better cognitive performance during school day. See Chapter~\ref{sec:selective-dysfunction}, lines 552--569, for detailed mechanism and neurophysiology.

\textbf{Pediatric School-Context Protocol:}

\begin{itemize}
    \item \textbf{Equipment}: White or pink noise machine (\$20--60) or free app (e.g., myNoise, Noisli). Ensure sound quality is consistent and volume is safe for children (not exceeding 50 dB).

    \item \textbf{How to use for school-attending children}: Play pink or white noise from sleep onset through entire sleep period. Standard setting: 50 dB continuous, starting 15 minutes before desired sleep time. Can be used alongside other sleep interventions (melatonin, light therapy).

    \item \textbf{School integration}: Children who are struggling with test performance and memory retention may see modest benefit within 3--6 weeks. Better sleep architecture supports consolidation of school learning.

    \item \textbf{Tracking}: Monitor sleep quality (child report), morning alertness, and school cognitive performance (test scores, teacher feedback) over 4--6 weeks.
\end{itemize}

\textbf{Evidence level}: Speculative (spindle enhancement demonstrated in sleep neuroscience; ME/CFS sleep architecture and acoustic enhancement RCTs pending)

\textbf{Why low priority for this population:} School-attending ambulatory children should prioritize sleep hygiene, circadian light therapy, and activity pacing before considering experimental sleep architecture enhancement. Useful only if other sleep interventions are optimized and memory/cognitive consolidation remains a limiting factor.

\end{recommendation}

\subsection{Cognitive Symptoms}
\label{subsec:ped-ambulatory-cognitive}

Cognitive dysfunction (``brain fog'') is often the most academically limiting symptom.

\subsubsection{Academic Accommodations First}

Before considering medications for cognitive symptoms, maximize accommodations:

\begin{itemize}
    \item Reduce cognitive load (fewer classes, reduced homework)
    \item Provide cognitive supports (notes, extended time, calculator use)
    \item Schedule demanding tasks during best times of day
    \item Allow rest breaks to prevent cognitive fatigue
\end{itemize}

Many children can function adequately academically with accommodations alone, without cognitive medications.

\subsubsection{Non-Pharmacological Cognitive Support}

\begin{itemize}
    \item \textbf{Organizational systems}: Calendars, apps, written reminders compensate for memory difficulties
    \item \textbf{Study strategies}: Spaced repetition, active recall, chunking information
    \item \textbf{Assistive technology}: Text-to-speech, speech-to-text, audiobooks
    \item \textbf{Cognitive rest}: Brief rest breaks (10 minutes every 30--45 minutes of cognitive work)
\end{itemize}

\subsubsection{Pharmacological Considerations}

If cognitive dysfunction remains severely limiting despite accommodations:

\begin{itemize}
    \item \textbf{OI treatment first}: Cognitive function often improves substantially with effective OI management. Always optimize OI treatment before adding cognitive medications.

    \item \textbf{Stimulants}: Medications like methylphenidate or dextroamphetamine may improve concentration. However:
    \begin{itemize}
        \item Developing brain concerns with chronic stimulant use
        \item Cardiovascular effects (tachycardia) may worsen OI
        \item Appetite suppression can impair nutrition
        \item Does not address underlying cause
        \item Risk of masking symptoms and enabling overexertion
    \end{itemize}
    Stimulants require specialist evaluation and careful monitoring; not first-line for pediatric ME/CFS cognitive symptoms.

    \item \textbf{Low-dose modafinil}: Used in some adolescents for fatigue/cognitive symptoms; limited pediatric data; similar concerns as stimulants regarding masking symptoms
\end{itemize}

\begin{practicalwarning}[Cognitive Medications May Enable Harmful Overexertion]
Medications that improve cognitive function or reduce fatigue perception can be dangerous if they enable students to exceed their energy envelope. A student who takes stimulants to focus through a full school day may crash harder afterward. Cognitive medications should support sustainable activity within the energy envelope, not expand the envelope artificially.
\end{practicalwarning}

\subsubsection{Transcranial Direct Current Stimulation (tDCS) for School-Attending Adolescents}

\begin{recommendation}[Pediatric tDCS for Cognitive Support (Age >12 with Caution)]
\label{rec:tdcs-ped-ambulatory}

\textbf{Mechanism:} tDCS applied to dorsolateral prefrontal cortex (DLPFC) may reduce baseline energy cost of executive function, improving sustainable academic performance (Chapter~\ref{sec:selective-dysfunction}, lines 207--228).

\textbf{Pediatric Considerations:}

\begin{enumerate}
    \item \textbf{Age requirements}:
    \begin{itemize}
        \item Minimum age: 12 years (limited pediatric safety data below this age)
        \item Adolescents 12+: Can consider with parental consent and physician supervision
        \item Requires mature understanding of electrode placement and safety
    \end{itemize}

    \item \textbf{Protocol (pediatric-adapted):}
    \begin{itemize}
        \item \textbf{Intensity}: 1--1.5 mA (lower than adult 2 mA due to developing brain)
        \item \textbf{Duration}: 15--20 minutes daily
        \item \textbf{Montage}: F3-F4 (DLPFC bilateral)
        \item \textbf{Frequency}: 5 days/week
        \item \textbf{Trial duration}: 4--8 weeks
        \item \textbf{Physician oversight}: Initial setup and safety screening required; ongoing monitoring
    \end{itemize}

    \item \textbf{Academic tracking}:
    \begin{itemize}
        \item Rate executive function daily during trial (focus, planning, organization)
        \item Track school performance (grade trends, assignment quality)
        \item Monitor for fatigue or mood changes
        \item Weekly summary by parent/adolescent
    \end{itemize}

    \item \textbf{Safety and cautions}:
    \begin{itemize}
        \item \textbf{Contraindications}: Metal implants in head, seizure history, pregnancy (adolescent females)
        \item \textbf{Mild side effects}: Tingling under electrodes (usually resolves), mild headache
        \item \textbf{Stop if}: Persistent headache, mood changes, anxiety, behavior problems
        \item \textbf{Parental supervision}: Required for electrode placement and safety monitoring
    \end{itemize}

    \item \textbf{Critical caveat}:
    \begin{itemize}
        \item tDCS is EXPERIMENTAL in pediatric ME/CFS; no published trials in this population
        \item May improve executive function without increasing energy envelope
        \item School accommodations remain essential; tDCS augments but does not replace them
        \item Risk: Improved focus might tempt increased academic load; strict energy envelope monitoring essential
    \end{itemize}
\end{enumerate}

\textbf{Evidence level}: Speculative (tDCS cognitive enhancement documented in other populations; pediatric ME/CFS safety and efficacy unknown)

\textbf{Expected outcomes (if effective):} 20--30\% improvement in focus/organization/planning. Not expected to improve fatigue directly.

\textbf{Who should try tDCS}: Adolescents age 12+ with prominent executive dysfunction limiting academics despite accommodations and OI optimization.

\end{recommendation}

\subsection{Infection Prevention as Recovery Window Preservation}
\label{subsec:ped-infection-prevention}

\begin{observation}[Infection Prevention and Pediatric Recovery Window]
\label{obs:ped-infection-recovery-window}

In pediatric patients with ME/CFS, infection prevention has compounded importance beyond the baseline harm prevention seen in adults. While all ME/CFS patients experience worsening from infections (see Chapter~\ref{ch:immune-dysfunction}, \hyperref[hyp:infection-damage-ratchet]{infection damage ratchet hypothesis}), pediatric patients face an additional stake: preserving the window of opportunity for recovery that distinguishes pediatric from adult prognosis.

\textbf{The Triple Impact of Infection in Pediatric ME/CFS:}

\begin{enumerate}
    \item \textbf{Acute PEM and baseline decline}: Each infection causes both immediate symptom exacerbation and lasting increases in baseline symptom severity---the ratchet effect common to all ME/CFS patients.

    \item \textbf{Sensitization cascade}: In pediatric patients, infection-induced sensitization appears to narrow the recovery window. Children who experience multiple infections during their illness trajectory show worse long-term outcomes than those who successfully prevent infections. Each infection may represent a lost opportunity for recovery.

    \item \textbf{Educational disruption with developmental consequences}: Post-infection crashes frequently force absences from school, creating cascading educational deficits. For developing children and adolescents, each semester of missed education has consequences beyond the immediate academic loss---social relationships are disrupted, identity formation around school-based activities is interrupted, and psychological impacts accumulate.
\end{enumerate}

\textbf{Clinical Implications for School-Attending Pediatric Patients:}

Infection prevention is not merely supportive care---it is active preservation of recovery potential. Strategies include:

\begin{itemize}
    \item \textbf{N95/FFP2 masking}: During known viral transmission in community or household. Given risk to recovery window, masking is medically justified even if symptomatic infections are not currently occurring.

    \item \textbf{Prompt antiviral treatment}: If viral infection suspected (respiratory symptoms, fever), initiate antivirals quickly to limit duration and severity. First-generation severity of infection is less damaging than allowing progression.

    \item \textbf{Prophylaxis during high-exposure periods}: Consider antiviral prophylaxis during high-transmission seasons or in households with sick members. The cost-benefit calculation is different for pediatric patients where infection prevention may preserve recovery trajectory.

    \item \textbf{Household isolation when possible}: When child is severely affected or during periods of particular vulnerability, isolation of the child from sick household members may be medically justified to preserve recovery potential.

    \item \textbf{School exposure management}: During high-transmission periods, discuss with school whether temporary remote learning is possible to reduce infection exposure. This is not academic coddling---it is medical management.
\end{itemize}

\textbf{The Pediatric Difference:}

Adult ME/CFS patients benefit from infection prevention to prevent baseline decline and symptom severity. Pediatric ME/CFS patients benefit from infection prevention for this reason AND to preserve the biological and psychological conditions in which recovery becomes possible. From a developmental perspective, each prevented infection is an investment in long-term outcome.

\end{observation}

\section{Activity and Exercise}
\label{sec:ped-activity-exercise}

The relationship between activity, exercise, and ME/CFS is widely misunderstood. This section provides evidence-based guidance.

\subsection{Critical Warning: GET is Contraindicated}
\label{subsec:ped-get-warning}

\begin{practicalwarning}[Graded Exercise Therapy Causes Harm in ME/CFS]
Graded Exercise Therapy (GET)---structured programs of progressively increasing exercise---is contraindicated in ME/CFS. This applies to children as well as adults, with potentially greater harm in developing patients.

\textbf{Evidence of harm}:
\begin{itemize}
    \item Patient surveys consistently report GET as one of the treatments most likely to cause deterioration
    \item The PACE trial, which promoted GET, has been discredited following independent reanalysis revealing methodological flaws and inflated claims~\cite{Rowe2017pediatric}
    \item NICE guidelines (UK) no longer recommend GET for ME/CFS
    \item CDC guidance no longer promotes GET
\end{itemize}

\textbf{Why GET harms ME/CFS patients}:
\begin{itemize}
    \item ME/CFS involves exercise intolerance as a core feature---the abnormal response to exertion IS the disease
    \item Progressive exercise triggers the post-exertional malaise mechanism, causing symptom worsening
    \item Repeated PEM episodes may cause cumulative damage and disease progression
    \item GET assumes deconditioning is the problem; in ME/CFS, the problem is abnormal energy metabolism and exercise response
\end{itemize}

\textbf{Why pediatric harm may be greater}:
\begin{itemize}
    \item Children are less able to advocate against inappropriate treatment
    \item School PE requirements may enforce harmful exercise
    \item Parents and teachers may pressure ``just exercise more'' despite symptoms
    \item Developmental period may be more vulnerable to damage from repeated overexertion
    \item Better baseline prognosis means more to lose from inappropriate treatment
\end{itemize}

\textbf{What to do if GET is recommended}:
\begin{itemize}
    \item Provide the recommending clinician with current NICE guidelines and CDC guidance
    \item Request documentation of the evidence base for GET in pediatric ME/CFS (none exists)
    \item Seek second opinion from ME/CFS-knowledgeable provider
    \item Document in writing your refusal of GET and the reasons
\end{itemize}
\end{practicalwarning}

\subsection{Safe Activity Maintenance}
\label{subsec:ped-safe-activity}

While GET is harmful, complete inactivity is also not ideal. The goal is maintaining baseline function within the energy envelope.

\subsubsection{Establishing Baseline}

\begin{enumerate}
    \item Track current activity and symptom patterns for 2 weeks
    \item Identify maximum sustainable activity level---the level that does NOT trigger PEM
    \item This becomes baseline: the ceiling, not the floor
\end{enumerate}

\subsubsection{Safe Activity Principles}

\begin{itemize}
    \item \textbf{Stay within envelope}: Never intentionally exceed sustainable capacity
    \item \textbf{No progression until stable}: Do not increase activity unless current level has been sustained for weeks without PEM
    \item \textbf{Minimal increases}: If increasing, increase by only 5--10\%, not more
    \item \textbf{Return to baseline after any PEM}: Any crash means activity was too high; return to lower level
    \item \textbf{``Good days'' are not permission}: On good days, do NOT do more---bank the energy for bad days
\end{itemize}

\subsubsection{Signs of Overexertion}

Teach children to recognize early warning signs that they are approaching or exceeding limits:

\begin{itemize}
    \item Heart rate rising above threshold
    \item Feeling ``wired'' or artificially energetic (adrenaline compensation)
    \item Difficulty catching breath
    \item Muscles feeling heavy or weak
    \item Cognitive function declining (losing words, difficulty concentrating)
    \item Onset of pain
\end{itemize}

When these signs appear, STOP and REST---do not ``push through.''

\subsubsection{Modified Physical Education}

Standard PE requirements are often impossible for children with ME/CFS. Options:

\begin{itemize}
    \item \textbf{Complete exemption}: If any physical activity triggers PEM, full PE exemption is appropriate
    \item \textbf{Modified participation}: Walking to tolerance, gentle stretching, yoga (restorative, not vigorous)
    \item \textbf{Academic alternative}: Health education coursework instead of physical activity
    \item \textbf{Manager/helper role}: Participate in PE class as manager, scorekeeper, or equipment manager without physical demands
\end{itemize}

\section{Social Development Considerations}
\label{sec:ped-social-development}

ME/CFS threatens normal social development during critical years. Active intervention is needed to maintain social health.

\subsection{Maintaining Peer Connections}

\begin{itemize}
    \item \textbf{Quality over quantity}: A few close, understanding friends are more valuable than many superficial connections
    \item \textbf{Educate close friends}: Friends who understand ME/CFS can accommodate limitations and become allies
    \item \textbf{Low-energy socialization}: Video calls, texting, gaming together online, watching movies together
    \item \textbf{Brief in-person contacts}: Short visits are better than no visits; friends can visit the child at home
    \item \textbf{School-based support}: Lunchtime with friends even on days with reduced class attendance maintains school social connections
\end{itemize}

\subsection{Online Social Options}

For children too ill for regular in-person socializing, online connection becomes essential:

\begin{itemize}
    \item \textbf{Chronic illness communities}: Online groups for young people with ME/CFS or chronic illness provide peer understanding
    \item \textbf{Gaming}: Multiplayer games provide social interaction with minimal physical demand
    \item \textbf{Interest-based communities}: Forums, Discord servers, or groups focused on the child's interests
    \item \textbf{Monitored appropriately}: Parents should ensure age-appropriate online safety while allowing necessary social connection
\end{itemize}

\subsection{Extracurricular Modification}

Extracurricular activities contribute to development and identity. Options:

\begin{itemize}
    \item \textbf{Reduced participation}: Attend practices or meetings when able; miss when unable
    \item \textbf{Alternative roles}: Manager, mentor, or remote participant rather than active participant
    \item \textbf{Low-demand activities}: Activities that can be done seated, at home, or with flexible participation (art, writing, online clubs)
    \item \textbf{Cessation with grief support}: If no modification enables safe participation, stopping may be necessary. Allow space to grieve the loss.
\end{itemize}

\subsection{Identity Formation}

Adolescents with ME/CFS face identity challenges:

\begin{itemize}
    \item \textbf{Disability identity}: ME/CFS becomes part of identity; this can be integrated healthily
    \item \textbf{Avoiding overidentification}: The child is not ONLY their illness; other aspects of identity matter
    \item \textbf{Future orientation}: What can the child aspire to? Realistic but hopeful goal-setting
    \item \textbf{Role models}: Adults with ME/CFS who have found meaningful lives can provide hope
    \item \textbf{Self-advocacy skills}: Learning to explain, advocate, and set boundaries is a life skill
\end{itemize}

\subsection{Preventing Isolation}

Social isolation increases depression risk and worsens prognosis. Active prevention:

\begin{itemize}
    \item Schedule regular social contact (even brief, even remote)
    \item Maintain at least one activity that connects to peers
    \item Family activities provide social interaction when peer interaction is limited
    \item Professional support (counseling) if isolation is significant
\end{itemize}

\section{Long-Term Management and Monitoring}
\label{sec:ped-longterm}

ME/CFS is typically a chronic condition requiring ongoing management, though pediatric prognosis is substantially better than adult.

\subsection{Trajectory Tracking}

\begin{itemize}
    \item \textbf{Baseline documentation}: Document function at diagnosis (school attendance, activity level, symptom severity)
    \item \textbf{Regular reassessment}: At least annually, document current function
    \item \textbf{Compare to own baseline}: Progress is measured against the child's own history, not peers
    \item \textbf{Identify trends}: Is the child improving, stable, or declining over months to years?
\end{itemize}

\subsection{When to Escalate Treatment Intensity}

Signs that warrant treatment escalation:

\begin{itemize}
    \item Function declining despite current management
    \item Increasing severity of crashes
    \item New symptoms developing
    \item School attendance declining
    \item Current treatment not achieving expected improvement
\end{itemize}

Escalation options:
\begin{itemize}
    \item Add medications not yet tried (OI medications, sleep medications)
    \item Increase accommodations (more reduced schedule, transition to homebound)
    \item Consult ME/CFS specialist if not already involved
    \item Consider aggressive interventions from Chapter~\ref{ch:urgent-action-severe} if appropriate
\end{itemize}

\subsection{Reevaluation Frequency}

\begin{itemize}
    \item \textbf{Initial stabilization}: Frequent contact (every 2--4 weeks) until symptoms stabilize
    \item \textbf{Stable management}: Every 3--6 months
    \item \textbf{Transitions}: More frequent during school transitions (new school year, middle to high school)
    \item \textbf{Puberty}: Hormonal changes may affect symptoms; monitor during pubertal transition
\end{itemize}

\subsection{Transition Planning}

For older adolescents, plan transitions:

\begin{itemize}
    \item \textbf{Transition to adult care}: Begin at age 16--17; identify adult ME/CFS providers
    \item \textbf{College planning}: If college-bound, consider accommodations, reduced course load, online options, schools with strong disability services
    \item \textbf{Workforce planning}: If not college-bound, vocational planning with disability considerations
    \item \textbf{Insurance continuity}: Plan for insurance coverage after aging off parents' plan
    \item \textbf{Self-management skills}: Adolescents should progressively take over managing their own pacing, medications, and appointments
\end{itemize}

\section{Prognosis Communication}
\label{sec:ped-prognosis-communication}

Children and families need accurate prognosis information---hopeful but realistic.

\subsection{The Pediatric Advantage}

Pediatric ME/CFS prognosis is substantially better than adult~\cite{Rowe2019pediatric}:

\begin{itemize}
    \item \textbf{Improvement/recovery rates}: 54--94\% (versus $\leq$22\% in adults)
    \item \textbf{Recovery by 10 years}: 68\%
    \item \textbf{Mean illness duration}: 5 years (range 1--15)
    \item \textbf{Function at follow-up}: Mean 8/10; only 5\% remain very unwell
\end{itemize}

This is genuinely good news. Most children with ME/CFS improve significantly or recover.

\subsection{Defining Recovery Appropriately}

``Recovery'' requires careful definition:

\begin{itemize}
    \item \textbf{Full resolution}: Complete return to pre-illness function with no symptoms (less common)
    \item \textbf{Substantial improvement}: Major increase in function; minimal ongoing symptoms; can live normal or near-normal life (more common)
    \item \textbf{No longer meeting criteria}: Still has some symptoms but does not meet diagnostic threshold (common)
\end{itemize}

Any of these outcomes is success. A child does not need to achieve perfect health to have a good outcome.

\subsection{Realistic Timelines}

\begin{itemize}
    \item \textbf{Months}: Symptom management may improve; functional improvement limited
    \item \textbf{Year 1--2}: Gradual improvement in many; some achieve significant gains
    \item \textbf{Years 2--5}: Continued improvement; many recover during this period
    \item \textbf{Years 5--10}: Most who will recover have done so; remaining patients may have chronic course
\end{itemize}

Improvement is slow and non-linear. Good weeks and bad weeks alternate. The trajectory is positive even if daily experience fluctuates.

\subsection{Hope Maintenance During Difficult Periods}

When progress is slow or absent:

\begin{itemize}
    \item \textbf{Acknowledge difficulty}: It is hard to be ill; it is hard to wait for improvement
    \item \textbf{Reframe success}: Stability is success; not getting worse is success; managing symptoms well is success
    \item \textbf{Focus on present}: What can be enjoyed today, regardless of tomorrow?
    \item \textbf{Connect with others}: Other families facing ME/CFS understand; support groups help
    \item \textbf{Professional support}: Mental health support for adjustment to chronic illness
\end{itemize}

\subsection{Contrast with Adult Prognosis}

For families asking ``what if this continues into adulthood?'':

\begin{itemize}
    \item Adult prognosis is worse, but not hopeless (20--25\% improve)
    \item Many adults with ME/CFS live meaningful lives with accommodations
    \item Adult treatments continue to improve
    \item The child's trajectory is not determined by adult statistics
\end{itemize}

\begin{keypoint}[The Goal of Early Management]
The goal of appropriate early management is to maximize the likelihood of being in the 54--94\% who improve or recover, rather than the small percentage who develop chronic severe disease. Every decision---about pacing, accommodations, avoiding GET, treating symptoms---should be evaluated against this goal. Early intervention is not about curing ME/CFS today; it is about preserving the window for recovery over the coming years.
\end{keypoint}

\section{Summary of Key Recommendations}
\label{sec:ped-ambulatory-summary}

\begin{enumerate}
    \item \textbf{Diagnose accurately}: Use pediatric-modified criteria; distinguish from school avoidance by PEM pattern; evaluate for OI.

    \item \textbf{Secure school accommodations}: Pursue 504 Plan or IEP with specific accommodations for cognitive dysfunction, OI, and PEM. Provide school with clear documentation.

    \item \textbf{Teach and enforce pacing}: Help children identify their energy envelope and stay within it. Resist pressure to ``push through'' from any source.

    \item \textbf{Treat OI aggressively}: Non-pharmacological measures first (hydration, salt, compression); add medications if needed. OI treatment often improves multiple symptoms.

    \item \textbf{Optimize sleep}: Melatonin first-line; address sleep hygiene; consider school start time accommodations.

    \item \textbf{Never recommend GET}: Graded exercise therapy is contraindicated and causes harm. If recommended by another provider, provide current evidence and decline.

    \item \textbf{Maintain social development}: Active effort to preserve peer connections and identity development despite illness limitations.

    \item \textbf{Monitor and adjust}: Regular reassessment; escalate treatment if declining; prepare for transitions.

    \item \textbf{Communicate realistic hope}: Most children improve; recovery is common; timelines are long; success has many definitions.
\end{enumerate}

\begin{continuation}[Related Content]
For children who are housebound or too ill for school attendance, see Chapter~\ref{ch:pediatric-severe}. For adult mild-moderate management, see Chapter~\ref{ch:action-mild-moderate}. For detailed pathophysiology, see Chapter~\ref{ch:cardiovascular} (orthostatic mechanisms), Chapter~\ref{ch:energy-metabolism} (metabolic dysfunction), and Chapter~\ref{ch:immune-dysfunction} (immune abnormalities).
\end{continuation}
