% FILE: Non-pharmaceutical interventions — supplements, nutraceuticals, vitamins, minerals, herbal products, dosage protocols
\chapter{Supplements and Nutraceuticals}
\label{ch:supplements}

\begin{flushright}
\textit{``First, do no harm. Second, be honest about what we don't know.''}
\end{flushright}

\vspace{1em}

This chapter reviews supplements and nutraceuticals commonly used in ME/CFS. We emphasize accessible, over-the-counter options that patients can try without prescription, while being honest about the limited evidence base. Most supplements have not been rigorously tested in ME/CFS-specific trials; recommendations are based on mechanism, related-condition evidence, and patient-reported outcomes.

\begin{observation}[Evidence Limitations]
Very few supplements have been tested in randomized controlled trials specifically for ME/CFS. Most evidence comes from: (1) mechanistic plausibility based on documented ME/CFS abnormalities, (2) trials in related conditions (fibromyalgia, Long COVID, chronic fatigue), (3) small open-label studies, and (4) patient surveys. We note the evidence level for each supplement. ``Widely used'' does not mean ``proven effective.''
\end{observation}


\section{Foundational Supplements: Electrolytes and Hydration}
\label{sec:electrolytes}

For patients with autonomic dysfunction (the majority of ME/CFS patients), electrolyte and fluid management is often the single most impactful intervention.

\subsection{Why Electrolytes Matter in ME/CFS}

Autonomic dysfunction in ME/CFS frequently manifests as:
\begin{itemize}
    \item Reduced blood volume (hypovolemia)
    \item Impaired vasoconstriction
    \item Excessive venous pooling
    \item Orthostatic intolerance
\end{itemize}

Adequate sodium, potassium, and fluid intake help maintain blood volume and support cardiovascular compensation.

\subsection{Sodium}

\paragraph{Rationale.} Sodium increases blood volume by promoting water retention. POTS guidelines recommend 10--12~g of salt daily (versus 2--3~g typical intake).

\paragraph{Evidence.} Moderate for POTS; limited ME/CFS-specific data, but widely reported helpful.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Table salt:} Cheapest; 2.3~g sodium per teaspoon
    \item \textbf{Electrolyte drinks:} LMNT, Liquid IV, Drip Drop, Nuun---convenient but expensive
    \item \textbf{Salt tablets:} Precise dosing; some find easier than drinking salty fluids
    \item \textbf{Oral rehydration salts (ORS):} WHO formula; includes glucose for sodium-glucose cotransport
\end{itemize}

\rev{5}
\paragraph{Practical Protocol.}
\begin{enumerate}
    \item Add 1/4--1/2 teaspoon salt to each liter of water
    \item Drink before standing or activity
    \item Morning loading: 16--24~oz with salt before getting up
    \item Target: 2--3~L fluid plus 8--12~g salt daily (approximately 3--5~g elemental sodium)
\end{enumerate}
\revend

\paragraph{Cautions.}
\begin{itemize}
    \item Check with physician if hypertension, heart failure, or kidney disease
    \item Some patients with hyperadrenergic POTS may worsen with excess sodium
    \item Monitor for edema; some is expected and indicates effectiveness
    \item \textbf{Potential for paradoxical intracellular effects:} ME/CFS patients may have baseline intracellular sodium elevation in muscle tissue~\cite{Petter2022sodium}. Rapid sodium loading could theoretically worsen intracellular sodium overload via impaired Na$^+$/K$^+$-ATPase function (see Section~\ref{subsec:electrolyte-initiation-caution})
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category A--B (context-dependent). Improved blood volume enhances oxygen and nutrient delivery to cells, supporting energy production. However, processing sodium via Na$^+$/K$^+$-ATPase consumes 20--40\% of cellular ATP. In ME/CFS patients with documented intracellular sodium overload, additional sodium loading may paradoxically increase ATPase energy demands. Net effect depends on individual pathophysiology: beneficial for hypovolemia-dominant presentation, potentially harmful if intracellular sodium is already elevated.
\revend

\subsection{Potassium}

\paragraph{Rationale.} Intracellular potassium is essential for nerve and muscle function. Some ME/CFS patients show functional potassium deficiency even with normal serum levels.

\paragraph{Evidence.} Theoretical; no ME/CFS-specific trials.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Potassium chloride:} Most common; Nu-Salt, Morton Lite Salt
    \item \textbf{Potassium citrate:} Better tolerated GI-wise
    \item \textbf{Coconut water:} Natural source ($\sim$600~mg per cup)
\end{itemize}

\paragraph{Dosing.} 2,000--4,700~mg daily (food + supplements); start low.

\paragraph{Cautions.} Excess potassium can cause cardiac arrhythmias. Do not exceed 99~mg per dose in supplement form without medical supervision. Those on ACE inhibitors, ARBs, or with kidney disease should be especially cautious.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Essential cofactor for Na$^+$/K$^+$-ATPase function---without adequate potassium, the pump cannot maintain membrane potential, and cellular energy production collapses. Correcting potassium deficiency directly enables ATP-dependent cellular processes. Minimal processing overhead.
\revend

\subsection{Magnesium}

\paragraph{Rationale.} Magnesium is a cofactor for $>$300 enzymes, including ATP synthesis. Deficiency is common and underdiagnosed (serum magnesium poorly reflects tissue status). Relevant to ME/CFS because:
\begin{itemize}
    \item Required for mitochondrial ATP production
    \item Modulates NMDA receptors (relevant to central sensitization)
    \item Supports autonomic function
    \item Promotes sleep (GABA-A receptor modulation)
\end{itemize}

\paragraph{Evidence.} Low--Moderate for ME/CFS; one small trial showed benefit with IM magnesium sulfate.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Magnesium glycinate:} Well-absorbed; calming; good for sleep; less GI upset
    \item \textbf{Magnesium malate:} Malic acid may support TCA cycle; often recommended for fibromyalgia
    \item \textbf{Magnesium L-threonate:} Crosses blood-brain barrier; may help cognition; expensive
    \item \textbf{Magnesium citrate:} Well-absorbed; can cause loose stools (useful if constipated)
    \item \textbf{Magnesium oxide:} Poorly absorbed; cheap; mainly useful as laxative
    \item \textbf{Magnesium taurate:} Combined with taurine; may benefit cardiovascular system
\end{itemize}

\paragraph{Dosing.} 200--600~mg elemental magnesium daily; split doses. Start low (100--200~mg) and increase gradually. Bowel tolerance is the limiting factor for oral forms.

\paragraph{Practical Tip.} Topical magnesium (Epsom salt baths, magnesium oil) provides modest absorption and may help with muscle symptoms, though evidence is limited.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Required cofactor for over 300 enzymatic reactions, including ATP synthesis itself (ATP exists physiologically as Mg-ATP complex). Deficiency directly impairs energy production at multiple points: glycolysis, TCA cycle, and oxidative phosphorylation. Supplementation restores enzymatic capacity with minimal metabolic processing cost. Among the highest-priority supplements for energy-depleted patients.
\revend

\subsection{Complete Electrolyte Formulas}

Many patients find pre-mixed electrolyte formulas convenient. Key ingredients to look for:
\begin{itemize}
    \item Sodium: 500--1000~mg per serving
    \item Potassium: 200--400~mg per serving
    \item Magnesium: 50--100~mg per serving
    \item Minimal or no sugar (some glucose aids sodium absorption; excessive sugar is counterproductive)
\end{itemize}

\paragraph{DIY Oral Rehydration Solution.}
\begin{quote}
1~L water + 1/2~tsp salt + 1/4~tsp potassium chloride (Nu-Salt) + 2~tbsp sugar or honey + optional: squeeze of citrus
\end{quote}
Cost: pennies per liter versus \$1--3 for commercial products.

\begin{warning}[Histamine Intolerance and Electrolyte Products]
\label{warn:hit-electrolytes}
A substantial subset of ME/CFS patients have histamine intolerance (HIT) or mast cell activation syndrome (MCAS)~\cite{Afrin2017MCAS}, making standard electrolyte products problematic (see Section~\ref{sec:mcas} for MCAS pathophysiology). Common triggers in commercial electrolyte formulas:

\textbf{Ingredients to avoid with HIT/MCAS:}
\begin{itemize}
    \item \textbf{Citric acid}: Ubiquitous in commercial products; can trigger mast cell degranulation
    \item \textbf{Citrus flavorings}: Natural lemon, lime, orange flavors are high-histamine
    \item \textbf{Artificial colors}: Red 40, Yellow 5/6 can trigger mast cells
    \item \textbf{Fermented ingredients}: Some products contain fermented sugars
    \item \textbf{Stevia}: Triggers reactions in some sensitive individuals
\end{itemize}

\textbf{HIT-safe DIY electrolyte recipe:}
\begin{quote}
1~L filtered water + 1/2~tsp sea salt (unrefined) + 1/4~tsp potassium chloride + 1--2~tbsp maple syrup (pure, grade A)
\end{quote}
Omit citrus entirely. Maple syrup provides glucose for sodium-glucose co-transport without histamine issues. Some patients tolerate small amounts of fresh ginger for flavor.

\textbf{Commercial options for HIT/MCAS:}
\begin{itemize}
    \item LMNT unflavored (citric acid-free)
    \item Pure Encapsulations Electrolyte/Energy Formula
    \item DIY remains safest option for highly sensitive patients
\end{itemize}

\textbf{Note:} POTS patients with HIT/MCAS face a double challenge---they need electrolytes for blood volume but react to most products. Trial elimination of problematic ingredients before assuming electrolyte intolerance. Many patients who thought they ``couldn't tolerate electrolytes'' actually couldn't tolerate citric acid or flavorings; plain salt in water may work when commercial products don't.
\end{warning}

\subsection{Caution: Electrolyte Initiation in ME/CFS}
\label{subsec:electrolyte-initiation-caution}

\begin{warning}[Intracellular vs.\ Extracellular Sodium: A Critical Distinction]
Standard POTS/dysautonomia guidelines recommend high sodium intake (10--12~g/day) to expand \textit{blood volume}---addressing \textit{extracellular} sodium. However, emerging evidence suggests ME/CFS patients may already have \textit{intracellular} sodium overload in muscle tissue, creating a potential paradox where aggressive sodium supplementation could worsen underlying pathophysiology even while improving some orthostatic symptoms~\cite{Wirth2021muscle,Petter2022sodium,Scheibenbogen2025muscle}.
\end{warning}

\paragraph{The Evidence.}
\begin{itemize}
    \item \textbf{MRI findings (2022):} Sodium MRI demonstrated elevated baseline intracellular muscle sodium in ME/CFS patients across all five lower leg muscle compartments (12.2~mM vs 9.4~mM in controls for anterior extensors, $p=0.003$). Post-exercise sodium accumulation was also greater (+30\% vs +17\% at 12~min)~\cite{Petter2022sodium}. \textit{Evidence quality: Low (single small study, $n=6$ per group); requires replication.}
    \item \textbf{Functional correlation:} Elevated muscle sodium correlated inversely with hand grip strength ($p=0.03$, $R^2=0.38$), suggesting functional consequences~\cite{Petter2022sodium}.
\end{itemize}

\rev{5}
\begin{speculation}[The NCX Reversal Mechanism]
\label{spec:ncx-reversal}
\textbf{Certainty: 0.25.} Mechanistic hypothesis based on indirect evidence (elevated intracellular sodium on MRI~\cite{Petter2022sodium}, known NCX biophysics); not directly tested in ME/CFS. Wirth and Scheibenbogen propose that $\beta_2$-adrenergic receptor dysfunction impairs Na$^+$/K$^+$-ATPase activity, allowing intracellular sodium accumulation~\cite{Wirth2021muscle,Scheibenbogen2025muscle}. When intracellular sodium exceeds a threshold, the sodium-calcium exchanger (NCX) reverses direction---importing calcium instead of exporting it---causing calcium overload and mitochondrial dysfunction. This model suggests ME/CFS patients operate near the NCX reversal threshold, explaining why even minor exertion triggers post-exertional malaise.
\end{speculation}
\revend

\paragraph{Clinical Implications.}
Dietary sodium primarily affects blood volume, not intracellular muscle sodium directly. However:
\begin{itemize}
    \item \textbf{Blood volume expansion remains beneficial} for most ME/CFS patients with orthostatic intolerance---the question is \textit{rate of initiation}
    \item \textbf{Rapid sodium loading} during energy-depleted states could theoretically stress already-compromised Na$^+$/K$^+$-ATPase function
    \item \textbf{Individual response varies}---some patients tolerate high sodium well; others report worsening with aggressive supplementation
\end{itemize}

\paragraph{``Start Low, Go Slow'' Protocol.}
Given the uncertainty, conservative initiation is prudent:
\begin{enumerate}
    \item \textbf{Week 1:} Baseline---maintain current sodium intake; establish symptom diary
    \item \textbf{Week 2:} Add 500--1000~mg sodium daily (approximately 1/4~tsp salt or one electrolyte packet)
    \item \textbf{Weeks 3--6:} If tolerated, increase by 500--1000~mg weekly
    \item \textbf{Target:} Reach 3--6~g supplemental sodium over 4--6 weeks (in addition to dietary intake)
    \item \textbf{Monitor:} Blood pressure (sitting and standing), energy levels, PEM threshold
\end{enumerate}

\paragraph{Warning Signs---Stop or Reduce If:}
\begin{itemize}
    \item Increased muscle pain, weakness, or fatigue unexplained by activity
    \item New or worsening palpitations or resting tachycardia
    \item Worsening PEM threshold (crashing at lower activity levels)
    \item Sustained blood pressure $>$130/80~mmHg at rest~\cite{Stock2022sodium}
    \item Significant peripheral edema beyond mild ankle swelling
\end{itemize}

\paragraph{Additional Considerations.}
\begin{itemize}
    \item \textbf{TRPM3 connection:} ME/CFS involves TRPM3 ion channel dysfunction affecting calcium handling (Section~\ref{sec:trpm3-dysfunction}). The relationship between TRPM3 dysfunction and NCX-mediated calcium overload warrants investigation
    \item \textbf{Energy state matters:} Supplementation during crashes may be less well-tolerated than during stable periods
    \item \textbf{Electrolyte balance:} Always supplement sodium alongside potassium and magnesium
    \item \textbf{Long-term risks:} High sodium intake carries cardiovascular risks independent of blood pressure~\cite{Stock2022sodium}; regular monitoring advisable
\end{itemize}

\begin{observation}[Research Gap]
The interaction between dietary sodium and intracellular muscle sodium in ME/CFS has not been directly studied. This caution is based on documented intracellular elevation, plausible mechanisms, and clinical reports of sodium intolerance in some patients. Prospective studies comparing titration protocols with sodium MRI endpoints would clarify this relationship.
\end{observation}


\section{Mitochondrial and Energy Support}
\label{sec:mito-support}

Given the evidence for energy metabolism dysfunction in ME/CFS (Chapter~\ref{ch:energy-metabolism}), supplements supporting mitochondrial function are among the most commonly used.

\subsection{Coenzyme Q10 (CoQ10/Ubiquinone/Ubiquinol)}

\paragraph{Rationale.} CoQ10 is a mobile electron carrier in the electron transport chain, shuttling electrons from Complex~I and Complex~II to Complex~III, and is a potent lipid-soluble antioxidant.

\paragraph{Evidence.} Moderate. Multiple small studies show benefit in ME/CFS~\cite{Maes2009CoQ10,CastroMarrero2015} (Castro-Marrero 2015: n=73, CoQ10+NADH combination) and fibromyalgia~\cite{Cordero2013}. A 2022 systematic review and meta-analysis of 13 RCTs (n=1,126) found CoQ10 significantly reduced fatigue across several chronic conditions (cancer fatigue, MS, fibromyalgia, cardiovascular disease; not ME/CFS-specific)~\cite{Tsai2022CoQ10fatigue}.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Ubiquinone:} Oxidized form; must be converted to ubiquinol
    \item \textbf{Ubiquinol:} Reduced (active) form; better absorbed, especially over age 40; more expensive
\end{itemize}

\paragraph{Dosing.}
\begin{itemize}
    \item Typical: 100--300~mg daily
    \item Higher doses in studies: 400--600~mg daily
    \item Take with fat-containing meal for absorption
    \item Split doses if $>$200~mg
\end{itemize}

\paragraph{Response Timeline.} Benefits may take 4--12 weeks to manifest.

\paragraph{Cautions.} Generally well-tolerated. May reduce warfarin effectiveness (see Section~\ref{subsec:mito-combination-guide}, drug interactions). Can cause insomnia if taken late in day (clinical observation).

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Functions as a mobile electron carrier in the ETC (see Rationale above). See Chapter~\ref{ch:medications-mechanisms} for detailed energy profile.
\revend

\subsection{NADH (Reduced Nicotinamide Adenine Dinucleotide)}
\label{subsec:nadh-supplement}

\paragraph{Rationale.} NADH is the reduced form of NAD$^+$ and serves as the primary electron donor to Complex~I (NADH dehydrogenase) of the mitochondrial electron transport chain. Supplementing NADH directly provides electrons for ATP synthesis, complementing CoQ10 which shuttles electrons between complexes. Together, NADH and CoQ10 support electron supply at two distinct points in the proximal electron transport chain.

\rev{5}
\paragraph{Evidence.} Low--Moderate (Low for NADH monotherapy; Moderate for the NADH+CoQ10 combination).
\begin{itemize}
    \item \textbf{Forsyth et al.\ (1999)}~\cite{Forsyth1999NADH}: Randomized, double-blind, placebo-controlled crossover trial (n=26); 10~mg NADH daily for 4 weeks showed 31\% response rate vs.\ 8\% placebo
    \item \textbf{Santaella et al.\ (2004)}~\cite{Santaella2004NADH}: Randomized trial (n=31, no placebo arm) comparing NADH to an active control (nutritional supplements plus psychotherapy) over 24 months; significant improvement in the NADH group in the first three months ($p < 0.001$ vs.\ control), but outcomes later comparable between groups
    \item \textbf{Castro-Marrero et al.\ (2016)}~\cite{CastroMarrero2016}: RCT (n=80); CoQ10 (200~mg) + NADH (20~mg) improved maximum heart rate recovery; some improvement in fatigue perception, though fatigue effects were inconsistent across the broader programme of trials
    \item \textbf{Castro-Marrero et al.\ (2021)}~\cite{CastroMarrero2021fatigue}: Largest RCT to date (n=207, double-blind, placebo-controlled); 200~mg CoQ10 + 20~mg NADH daily for 12 weeks showed significant improvements in cognitive fatigue ($p < 0.001$), overall fatigue ($p = 0.022$), quality of life ($p < 0.05$), and sleep parameters (sleep duration at week~4, $p = 0.018$; sleep efficiency at week~8, $p = 0.038$). Benefits persisted 4 weeks post-treatment. No serious adverse events
\end{itemize}

\rev{5}
The Castro-Marrero 2021 trial provides the strongest evidence for the NADH+CoQ10 combination in ME/CFS. It does not, however, distinguish the individual contribution of NADH from that of CoQ10. See Section~\ref{subsec:mito-combination-guide} for combination rationale.

\rev{5}
\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Enteric-coated tablets:} Protect against gastric acid degradation; most common form
    \item \textbf{Sublingual:} Bypasses gastric degradation; may provide faster absorption
    \item \textbf{Enada:} Brand used in multiple clinical trials; standardized formulation
\end{itemize}

\rev{5}
\paragraph{Dosing.}
\begin{itemize}
    \item Standard: 10--20~mg daily
    \item Clinical trial dose: 20~mg daily (Castro-Marrero protocol)
    \item Must be taken on an empty stomach, 30--60 minutes before breakfast (gastric acid degrades unprotected NADH)
\end{itemize}
\revend

\paragraph{Response Timeline.} 4--8 weeks for monotherapy; combination with CoQ10 may show benefits within 4 weeks for sleep parameters.

\paragraph{Cautions.} Generally well-tolerated. Unprotected oral NADH is degraded by stomach acid---use enteric-coated or sublingual forms only. Some patients report mild nausea if taken with food. Cost is moderate (\$20--40/month at therapeutic doses).

\paragraph{NADH vs.\ NAD$^+$ Precursors.} NADH and NAD$^+$ precursors (NR, NMN) are complementary, not interchangeable---they target different axes of cellular energy metabolism. See Section~\ref{subsec:mito-combination-guide} for the full two-axis framework and combination recommendations.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Direct electron donor to Complex~I; each NADH molecule reaching the mitochondrial matrix yields approximately 2.5 ATP via oxidative phosphorylation. Oral bioavailability is a limiting factor (see Forms above).
\revend

\subsection{NAD$^+$ Precursors: Nicotinamide Riboside (NR) and NMN}

\paragraph{Rationale.} NAD$^+$ is essential for mitochondrial function, DNA repair, and cellular signaling. The Heng 2025 study~\cite{heng2025mecfs} documented NAD$^+$ metabolism abnormalities in ME/CFS. While direct oral NAD$^+$ supplementation has poor bioavailability (gastric degradation), sublingual NAD$^+$ (100--300~mg/day) offers a direct route, and precursors efficiently raise intracellular levels. NAD$^+$ precursors replenish the cellular NAD$^+$ pool for sirtuins and PARP-mediated DNA repair---a complementary mechanism to the direct electron donation provided by NADH (Section~\ref{subsec:nadh-supplement}). For combination guidance, see Section~\ref{subsec:mito-combination-guide}.

\paragraph{Evidence.} Preliminary. A 2025 RCT in Long COVID (n=58; 20-week intervention) showed NR 2000~mg/day increased whole-blood NAD$^+$ levels 2.6--3.1 fold (measured at interim timepoints of 5 and 10 weeks); cognitive benefits were variable, with overall group effects limited but some individuals showing improvement after $\geq$10 weeks~\cite{Wu2025NRLongCOVID}.

\rev{5}
\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Nicotinamide Riboside (NR):} Tru Niagen is the most studied brand
    \item \textbf{Nicotinamide Mononucleotide (NMN):} One step closer to NAD$^+$; theoretically more direct but less clinical data
    \item \textbf{Niacin (B3):} Cheapest NAD$^+$ precursor but causes flushing; extended-release reduces flushing but has liver toxicity concerns
    \item \textbf{Nicotinamide:} No flushing; inhibits sirtuins via product inhibition at high concentrations (well-established in vitro~\cite{Bitterman2002Nicotinamide}; clinical relevance at oral supplementation doses uncertain)
    \item \textbf{NAD$^+$ sublingual:} Direct NAD$^+$ administered under the tongue (100--300~mg/day); bypasses gastric degradation but bioavailability data remain limited. Substitutable with NR or NMN---choose one precursor strategy
\end{itemize}

\rev{5}
\paragraph{Dosing.}
\begin{itemize}
    \item NR: 300--1000~mg daily typical; research doses up to 2000~mg
    \item NMN: 250--500~mg daily typical; up to 1000~mg in some regimens
    \item NAD$^+$ sublingual: 100--300~mg daily
    \item Niacin: 500--1500~mg daily (with caution)
\end{itemize}
\revend

\paragraph{Emergency Use (Speculative).} For emergency post-exertion PEM prevention, high-dose loading (1000--2000~mg NR or NMN immediately post-exertion, then 500~mg twice daily for 3--5 days) may prevent NAD$^+$ depletion hypothetically attributable to PARP activation during DNA repair---this mechanism is plausible but unconfirmed in ME/CFS specifically~\cite{Dehhaghi2022AD}. \textbf{Caution:} This protocol has not been tested in any clinical trial. High-dose NR/NMN may cause gastrointestinal side effects and has not been safety-tested at these doses in ME/CFS patients. Discuss with physician before attempting. See Chapter~\ref{ch:emerging-therapies}, \S\ref{subsec:pem-prevention} for complete emergency protocol.

\paragraph{Response Timeline.} May require 10+ weeks for noticeable benefit.

\paragraph{Cost Consideration.} NR/NMN are expensive (\$50--150/month at therapeutic doses). Niacin is cheap but has tolerability issues.

\paragraph{Cautions.} Generally well-tolerated at standard doses. Gastrointestinal discomfort may occur at high doses ($\geq$1000~mg). Theoretical concern: elevated NAD$^+$ may promote tumor growth in active cancer---rapidly proliferating cells may exploit NAD$^+$ for energy and DNA repair~\cite{Palmer2021NADcancer}. Avoid during active cancer; consult oncologist before use in patients with cancer history. Extended-release niacin carries hepatotoxicity risk (see Section~\ref{subsec:mito-combination-guide}).

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Replenish cellular NAD$^+$ pools required for hundreds of metabolic reactions including glycolysis, TCA cycle, and oxidative phosphorylation. Conversion from NR/NMN to NAD$^+$ requires minimal ATP (one phosphorylation step). Net energy balance strongly positive. Emergency high-dose protocols (see Emergency Use above) are hypothesized to provide immediate NAD$^+$ substrate for PARP-mediated DNA repair~\cite{Dehhaghi2022AD}---this rationale is mechanistically plausible but not yet confirmed in ME/CFS trials.
\revend

\subsection{D-Ribose}

\paragraph{Rationale.} D-ribose is the pentose sugar component of ATP. Supplementation may accelerate ATP resynthesis after depletion.

\paragraph{Evidence.} Low. A pilot open-label study (n=41)~\cite{Teitelbaum2006ribose} and a larger multicenter replication (n=257; $p < 0.0001$ within-group for energy, sleep, and well-being)~\cite{Teitelbaum2012ribose} in ME/CFS and fibromyalgia showed improvements with 5~g three times daily; no placebo comparison in either study, so results cannot be distinguished from placebo effect.

\paragraph{Dosing.} 5~g three times daily (15~g/day total); can be reduced to 5--10~g daily for maintenance.

\paragraph{Emergency Use.} For emergency post-exertion PEM prevention, higher acute doses (10--15~g immediately post-exertion, then 5~g every 4--6 hours for 24--48h) may be used as part of comprehensive crash prevention protocol. See Chapter~\ref{ch:emerging-therapies}, \S\ref{subsec:pem-prevention} for complete emergency protocol.

\paragraph{Practical Tips.}
\begin{itemize}
    \item Take with meals (can lower blood sugar)
    \item Sweet taste; dissolves in beverages
    \item Some patients report energy improvement within days
\end{itemize}

\paragraph{Cautions.} May lower blood sugar; diabetics should monitor carefully.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Direct ATP substrate (pentose sugar component). May reduce reliance on the energy-intensive pentose phosphate pathway by entering purine salvage directly~\cite{Dodd2004ribose}. Among the most direct energy substrates available---minimal metabolic processing required for ATP regeneration.
\revend

\subsection{Acetyl-L-Carnitine (ALCAR) and L-Carnitine}

\paragraph{Rationale.} Carnitine transports fatty acids into mitochondria for oxidation. Deficiency impairs fat-based energy production. Acetyl-L-carnitine crosses the blood-brain barrier and may support cognitive function.

\paragraph{Evidence.} Low (open-label studies only; no placebo-controlled trials). CFS patients show lower serum carnitine levels correlating with functional capacity~\cite{Plioplys1995carnitine}. L-carnitine 3~g/day improved 12 of 18 clinical parameters over 8 weeks (n=30, open-label)~\cite{Plioplys1997carnitineTreatment}. Acetyl-L-carnitine improved mental fatigue ($p = 0.015$) and propionyl-L-carnitine improved general fatigue ($p = 0.004$) in an open-label randomized study (n=90, no placebo arm)~\cite{Vermeulen2004carnitine}.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{L-Carnitine:} General mitochondrial support
    \item \textbf{Acetyl-L-Carnitine (ALCAR):} Better for cognitive symptoms; crosses BBB
    \item \textbf{Propionyl-L-Carnitine:} May be better for cardiovascular symptoms
\end{itemize}

\paragraph{Dosing.} 500--2000~mg daily; split doses.

\paragraph{Response Timeline.} 2--8 weeks.

\paragraph{Cautions.} Can increase TMAO (a proposed cardiovascular risk marker) with chronic use; the mechanism---gut microbial conversion of carnitine to trimethylamine, then hepatic oxidation to TMAO---is documented in published literature, though the clinical significance for cardiovascular risk remains debated. Some patients experience overstimulation or insomnia. L-carnitine may theoretically lower seizure threshold in patients with pre-existing epilepsy based on mechanistic considerations; consult physician before use in seizure disorders (no controlled data; clinical caution).

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Enables mitochondrial fatty acid oxidation---the most energy-dense metabolic pathway. Each fatty acid molecule transported into mitochondria by carnitine yields substantially more ATP than glucose oxidation. Processing overhead negligible relative to energy yield.
\revend

\subsection{Creatine}

\paragraph{Rationale.} Creatine buffers ATP, providing rapid energy during high-demand situations. Well-studied for muscle function; emerging evidence for cognitive benefits.

\paragraph{Evidence.} Theoretical for ME/CFS; strong for lower limb muscle strength in general populations (meta-analysis)~\cite{Lanhers2015creatine}. Emerging evidence for cognitive benefits under stress---sleep deprivation, mental fatigue---in general populations (small RCTs; no ME/CFS-specific data).

\rev{5}
\paragraph{Dosing.}
\begin{itemize}
    \item Loading (optional): 5~g four times daily for 5--7 days. Note: the loading phase is derived from sports nutrition protocols; ME/CFS patients may prefer to skip loading and start directly at maintenance doses, consistent with conservative introduction (Section~\ref{sec:protocols})
    \item Maintenance: 3--5~g daily
\end{itemize}
\revend

\paragraph{Cautions.} Requires adequate hydration. May cause water retention. Generally considered safe for kidney function in healthy individuals at normal doses (no controlled evidence of harm; serum creatinine elevation reflects creatine metabolism, not renal impairment---see Additional interactions below).

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Directly buffers ATP via the phosphocreatine system---the fastest cellular energy reserve. Creatine kinase converts phosphocreatine to ATP within milliseconds, providing immediate energy during demand spikes. Also supports brain energy metabolism. Minimal processing demands.
\revend

\subsection{PQQ (Pyrroloquinoline Quinone)}

\paragraph{Rationale.} PQQ stimulates mitochondrial biogenesis (creation of new mitochondria) and has antioxidant properties.

\paragraph{Evidence.} Preliminary. Small RCTs in healthy adults show cognitive benefits: improved attention and working memory with 20~mg/day over 12 weeks~\cite{Tamakoshi2023PQQbrain,Shiojima2022PQQcognitive}; no ME/CFS-specific trials.

\paragraph{Dosing.} 10--20~mg daily.

\rev{5}
\paragraph{Energy Profile.}
Category A (indirect; delayed onset: weeks to months). Net energy provider through mitochondrial biogenesis rather than direct substrate supply. Stimulates the creation of new mitochondria, increasing total cellular energy production capacity over weeks to months. Unlike CoQ10 or NADH, PQQ does not directly participate in electron transport---its energy benefit is mediated entirely through expanding mitochondrial mass. Do not expect acute energy improvement.
\revend

\subsection{Mitochondrial Energy Support: Combination Guide}
\label{subsec:mito-combination-guide}

\rev{5}
The supplements in this section target distinct steps in cellular energy production. Understanding which ones are complementary and which are substitutable prevents both redundant spending and missed synergies.
\revend

\subsubsection{Two Complementary Axes}

\rev{5}
Mitochondrial energy supplements can be organized into two mechanistically distinct axes. This framework is a conceptual model for understanding supplement complementarity; the empirical support comes from the Castro-Marrero trials, not from the framework itself.
\revend

\paragraph{Axis~1: Electron Transport Chain Support (NADH + CoQ10).} NADH donates electrons to Complex~I; CoQ10 shuttles electrons from Complex~I and Complex~II to Complex~III. Together they support electron supply at two distinct points in the proximal ETC. The Castro-Marrero 2021 RCT (n=207) used exactly this combination---200~mg CoQ10 + 20~mg NADH---and demonstrated significant improvements in fatigue, cognition, and sleep parameters~\cite{CastroMarrero2021fatigue}. These two supplements form a functional pair and should be taken together.

\rev{5}
\paragraph{Axis~2: NAD$^+$ Pool Replenishment (choose one precursor).} NR, NMN, and sublingual NAD$^+$ all aim to increase intracellular NAD$^+$ levels, supporting sirtuins (cellular stress response), PARP (DNA repair), and metabolic signaling. Because they converge on the same endpoint, one should be chosen based on evidence, tolerability, and cost. NR and NMN are the best-supported options and are largely interchangeable for this purpose:
\begin{itemize}
    \item \textbf{NR:} Most clinical data; well-absorbed; Tru Niagen brand used in trials
    \item \textbf{NMN:} One enzymatic step closer to NAD$^+$; popular but fewer published RCTs.\footnote{Whether NMN requires conversion to NR for cellular uptake (via CD73) or enters cells directly (via the Slc12a8 transporter) remains debated. This does not affect the practical recommendation to choose one precursor.}
    \item \textbf{NAD$^+$ sublingual:} Direct delivery; limited bioavailability data; cannot be considered fully equivalent to NR or NMN until absorption is better characterized
\end{itemize}

\rev{5}
Niacin and nicotinamide are also NAD$^+$ precursors but are not recommended for Axis~2: niacin causes flushing and carries hepatotoxicity risk at extended-release doses, while nicotinamide inhibits sirtuins at high concentrations (well-established in vitro; see Forms above)---directly counteracting one of Axis~2's intended benefits.

\rev{5}
These two axes are complementary, not redundant. After donating electrons at Complex~I, NADH is oxidized back to NAD$^+$, so it does contribute to the mitochondrial NAD$^+$ pool. However, supplemental NADH doses are modest (10--20~mg, approximately 15--30~$\mu$mol), while total body NAD$^+$ content is estimated at 1--3~g (approximately 1,500--4,500~$\mu$mol; estimate based on tissue concentration measurements~\cite{Massudi2012nad})---dwarfing the contribution from a single NADH dose. NADH's primary pharmacological action is therefore acute electron donation, not sustained NAD$^+$ pool elevation. NAD$^+$ precursors (NR, NMN) act through the salvage pathway to raise NAD$^+$ levels across all cellular compartments---cytosolic, nuclear, and mitochondrial---at much larger scale. Thus, Axis~1 provides the electrons that drive ATP synthesis now, while Axis~2 replenishes the coenzyme pool that sustains energy metabolism, DNA repair, and cellular signaling over time.
\revend

\subsubsection{Age-Stratified Dosing}

\rev{5}
\begin{warning}[Pediatric Evidence Gap]
Pediatric dosing for mitochondrial supplements is extrapolated from adult data and body-weight scaling, not from pediatric randomized controlled trials. No ME/CFS-specific supplement trials have been conducted in children or adolescents. All pediatric use should be under physician supervision, with conservative starting doses and careful monitoring.
\end{warning}

\rev{5}
\begin{table}[htbp]
\centering
\caption{Mitochondrial supplement dosing by age group}
\label{tab:mito-dosing-by-age}
\small
\begin{tabularx}{\textwidth}{l c c c X}
\toprule
\textbf{Supplement} & \textbf{Adults ($\geq$18\,y)} & \textbf{Adolescents (12--17\,y)} & \textbf{Children (6--11\,y)} & \textbf{Notes} \\
\midrule
NADH & 10--20\,mg & 5--10\,mg & Not recommended & Empty stomach, 30\,min before breakfast \\
CoQ10 (ubiquinol) & 100--300\,mg & 50--100\,mg & 25--50\,mg & With fat-containing meal; RCT dose 200\,mg \\
NR & 300--1000\,mg & 150--250\,mg & Not recommended & 300--500\,mg starting; research doses up to 2000\,mg \\
NMN & 250--500\,mg & 125--250\,mg & Not recommended & Choose one NAD$^+$ precursor \\
NAD$^+$ sublingual & 100--300\,mg & Not recommended & Not recommended & Limited safety data \\
D-Ribose & 5--15\,g & 3--5\,g & 1--3\,g & With meals; may lower blood sugar \\
ALCAR & 500--2000\,mg & 250--500\,mg & Not recommended & Split doses; may cause insomnia \\
Creatine & 3--5\,g & 2--3\,g & 1--2\,g & Adequate hydration essential \\
PQQ & 10--20\,mg & 10\,mg & Not recommended & Preliminary evidence only \\
\bottomrule
\end{tabularx}
\end{table}
\revend

\subsubsection{Recommended Combinations}

\rev{5}
Three evidence-informed combination tiers, using the two-axis framework above as an organizational heuristic; only the Minimal tier has direct RCT support. Use age-appropriate doses from Table~\ref{tab:mito-dosing-by-age}. For children under~12, where NADH is not recommended, the Minimal tier reduces to CoQ10 alone.
\revend

\paragraph{Minimal (the Castro-Marrero protocol).} NADH + CoQ10. This is the only combination tested in a large ME/CFS RCT~\cite{CastroMarrero2021fatigue}. Start here.

\paragraph{Standard.} NADH + CoQ10 + one NAD$^+$ precursor (NR or NMN). Adds Axis~2 support for the cellular NAD$^+$ pool. NR has the most clinical data; NMN is a reasonable alternative.

\paragraph{Comprehensive.} NADH + CoQ10 + one NAD$^+$ precursor + D-ribose + ALCAR. Adds ATP substrate (D-ribose) and fatty acid transport into mitochondria (carnitine). Higher cost; consider adding components one at a time to assess individual benefit.

\subsubsection{What Not to Combine}

\rev{5}
\paragraph{Redundant combinations (not dangerous, but wasteful).}
\begin{itemize}
    \item \textbf{Multiple NAD$^+$ precursors at full dose:} Taking NR 500~mg + NMN 500~mg + NAD$^+$ sublingual 300~mg simultaneously provides no proportional benefit over a single precursor---they all converge on the same intracellular NAD$^+$ pool. Because NR and NMN differ in molecular weight, a simple 50/50~mg split does not achieve equal molar doses---the simplest approach is to choose one precursor and use it at full dose
    \item \textbf{High-dose niacin alongside NR or NMN:} Niacin is also an NAD$^+$ precursor; stacking it with NR/NMN adds flushing risk without clear additional benefit
\end{itemize}

\rev{5}
\begin{warning}[Clinically Significant Drug Interactions]
\begin{itemize}
    \item \textbf{CoQ10 + warfarin:} CoQ10 may reduce warfarin effectiveness---the mechanism may involve enhanced CYP450-mediated warfarin hydroxylation~\cite{Zhou2005CoQ10Warfarin}; a structural similarity to vitamin~K has also been proposed but is less well-documented. Monitor INR closely if combining; adjust warfarin dose with physician guidance
    \item \textbf{Extended-release niacin + statins:} Extended-release niacin carries hepatotoxicity risk; combined with statins, risk of myopathy and rhabdomyolysis is increased~\cite{Anderson2014AIMHIGHNiacin}. Avoid unless under strict medical supervision with liver function monitoring
    \item \textbf{D-ribose + diabetes medications:} Ribose can lower blood glucose; monitor closely and adjust hypoglycemic medication if needed.
    \item \textbf{ALCAR + thyroid medication:} L-carnitine inhibits thyroid hormone nuclear uptake in a dose-dependent manner~\cite{Benvenga2001CarnitineHyperthyroidism}; patients on thyroid replacement should consult their physician before supplementing L-carnitine and monitor thyroid function
\end{itemize}
\end{warning}

\rev{5}
\paragraph{Additional interactions.}
\begin{itemize}
    \item \textbf{CoQ10 + antihypertensives:} CoQ10 can lower blood pressure; combining with fludrocortisone or midodrine (common in ME/CFS for orthostatic intolerance) may require dose adjustment. Monitor blood pressure when initiating
    \item \textbf{D-ribose + glucose monitoring:} Some glucose meters can misread ribose as glucose, producing falsely elevated readings. Use ribose-compatible meters if monitoring blood glucose regularly
    \item \textbf{Creatine + renal function monitoring:} Creatine supplementation raises serum creatinine (a creatine metabolite), which can confound laboratory assessment of kidney function. Inform physicians of creatine use before renal panels; cystatin~C is an alternative GFR marker unaffected by creatine intake
\end{itemize}

\rev{5}
\paragraph{Timing conflicts.}
\begin{itemize}
    \item NADH requires an empty stomach (30--60 minutes before food); CoQ10 requires fat with meals. Take NADH first thing in the morning, CoQ10 with breakfast or lunch
    \item NAD$^+$ precursors (NR, NMN) may have a mildly stimulating effect---avoid evening dosing if sleep-sensitive
    \item D-ribose should be taken with meals to prevent blood sugar dips
    \item ALCAR may cause insomnia or overstimulation; take split doses in the morning and early afternoon, not after 2--3~PM
\end{itemize}
\revend


\section{Antioxidant and Anti-inflammatory Supplements}
\label{sec:antioxidants}

Oxidative stress is documented in ME/CFS~\cite{Syed2025MitoCFS}. Antioxidants may help, though evidence for specific supplements is limited.

\subsection{N-Acetylcysteine (NAC)}

NAC is one of the most versatile and evidence-supported supplements relevant to ME/CFS.

\paragraph{Rationale.}
\begin{itemize}
    \item \textbf{Glutathione precursor:} NAC provides cysteine, the rate-limiting amino acid for glutathione synthesis
    \item \textbf{Direct antioxidant:} Scavenges free radicals
    \item \textbf{Anti-inflammatory:} Reduces NF-$\kappa$B activation
    \item \textbf{Mucolytic:} Thins mucus (originally developed for this)
    \item \textbf{Supports liver detoxification:} Used clinically for acetaminophen overdose
    \item \textbf{May reduce viral replication:} Some evidence for various viruses
\end{itemize}

\paragraph{Evidence.} Moderate for general antioxidant/anti-inflammatory effects; preliminary for ME/CFS specifically. Widely used with generally positive patient reports.

\paragraph{Dosing.}
\begin{itemize}
    \item Typical: 600--1200~mg daily
    \item Higher doses: 1800--2400~mg daily (used in psychiatric applications)
    \item Take on empty stomach for best absorption
    \item Divide doses if $>$600~mg
\end{itemize}

\paragraph{Response Timeline.} Antioxidant effects within days; systemic benefits may take 4--8 weeks.

\paragraph{Cautions.}
\begin{itemize}
    \item Can cause GI upset; start low
    \item Sulfur smell (normal)
    \item Theoretical concern about reducing beneficial ROS signaling; probably not clinically significant at normal doses
    \item May thin mucus excessively in some (actually beneficial for most)
\end{itemize}

\paragraph{Synergy.} NAC works synergistically with:
\begin{itemize}
    \item Glycine: Another glutathione precursor
    \item Selenium: Required for glutathione peroxidase function
    \item Vitamin C: Regenerates oxidized glutathione
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category A--B (provider to neutral). Glutathione precursor that reduces oxidative stress, indirectly conserving cellular energy otherwise spent on damage repair. The conversion of NAC to glutathione requires minimal ATP (two enzymatic steps). By reducing oxidative damage to mitochondrial membranes and enzymes, NAC helps preserve energy production efficiency.
\revend

\subsection{Alpha-Lipoic Acid (ALA)}

\paragraph{Rationale.} ALA is both water- and fat-soluble, allowing it to work in all cellular compartments. Regenerates other antioxidants (vitamins C and E, glutathione). Supports mitochondrial function.

\paragraph{Evidence.} Theoretical for ME/CFS. (Moderate evidence exists for diabetic neuropathy, establishing the supplement's general bioactivity, but no ME/CFS-specific trials.)

\paragraph{Dosing.} 300--600~mg daily; R-lipoic acid is the more bioactive form. Start at 100--150~mg in ME/CFS patients and titrate upward---sensitive patients may experience paradoxical worsening (increased fatigue, malaise, or gastrointestinal symptoms) at higher initial doses (clinical observation).

\paragraph{Cautions.} Can lower blood sugar. May chelate minerals (take separately from mineral supplements).

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider at supplemental doses). Essential cofactor for pyruvate dehydrogenase and alpha-ketoglutarate dehydrogenase---enables energy production rather than consuming it. At high doses, mineral chelation may transiently reduce cofactor availability (see Cautions above). See Chapter~\ref{ch:medications-mechanisms} for detailed profile.
\revend

\subsection{Omega-3 Fatty Acids (EPA/DHA)}

\paragraph{Rationale.}
\begin{itemize}
    \item Anti-inflammatory (compete with omega-6 for inflammatory mediator synthesis)
    \item Support cell membrane fluidity
    \item Neuroprotective
    \item May support endothelial function (relevant to vascular hypothesis)
\end{itemize}

\paragraph{Evidence.} Moderate for general anti-inflammatory effects; limited ME/CFS-specific data.

\paragraph{Dosing.}
\begin{itemize}
    \item General health: 1--2~g combined EPA/DHA daily
    \item Anti-inflammatory: 2--4~g daily
    \item Higher EPA ratio may be more anti-inflammatory
\end{itemize}

\paragraph{Quality Matters.} Fish oil can oxidize; look for third-party tested products (IFOS certification). Triglyceride form is better absorbed than ethyl ester.

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Anti-inflammatory effects may reduce the metabolic cost of chronic systemic inflammation. Incorporated into cell membranes through normal lipid metabolism with modest processing demands. No direct energy provision or significant energy consumption.
\revend

\subsection{Curcumin}

\paragraph{Rationale.} Potent anti-inflammatory; inhibits NF-$\kappa$B; antioxidant.

\paragraph{Evidence.} Strong for inflammation generally; no ME/CFS-specific trials.

\paragraph{Bioavailability Challenge.} Standard curcumin is poorly absorbed ($<$1\%). Enhanced formulations necessary:
\begin{itemize}
    \item Curcumin + piperine (black pepper extract): 20$\times$ absorption increase
    \item Phytosome forms (Meriva): Lipid-bound for better absorption
    \item Nano-curcumin, micellar curcumin: Various enhanced delivery systems
\end{itemize}

\paragraph{Dosing.} Depends on formulation; typically 500--2000~mg of enhanced curcumin daily.

\paragraph{Cautions.} May thin blood; caution with anticoagulants. Can cause GI upset. May interact with some medications.

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Hepatic glucuronidation and sulfation for metabolism, but enhanced bioavailability formulations reduce first-pass processing. Anti-inflammatory effects may offset processing costs by reducing chronic inflammation energy burden. Overall metabolic impact modest.
\revend

\subsection{Quercetin}

\paragraph{Rationale.}
\begin{itemize}
    \item Mast cell stabilizer (relevant if MCAS component)
    \item Antioxidant
    \item Anti-inflammatory
    \item May have antiviral properties
\end{itemize}

\paragraph{Evidence.} Theoretical for ME/CFS; moderate for mast cell conditions.

\paragraph{Dosing.} 500--1000~mg daily; enhanced absorption forms (quercetin phytosome) preferred.

\paragraph{Cautions.} Generally well-tolerated. May interact with some antibiotics.

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Flavonoid with mast cell-stabilizing properties processed through normal Phase~II hepatic conjugation. Minimal energy demands. Mast cell stabilization may reduce the significant energy cost of chronic degranulation episodes.
\revend

\section{B Vitamins}
\label{sec:vitamins}

B vitamins are essential cofactors for energy metabolism and neurological function.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). B vitamins are essential cofactors for virtually every step of cellular energy production: thiamine (pyruvate dehydrogenase), riboflavin (Complex~I/II), niacin (NAD$^+$ synthesis), pantothenic acid (CoA synthesis), pyridoxine (amino acid metabolism). Deficiency at any point creates an energy production bottleneck. Supplementation removes bottlenecks with negligible processing cost.
\revend

\subsection{Thiamine (B1)}

\paragraph{Rationale.} Essential for pyruvate dehydrogenase (PDH)---the enzyme that feeds pyruvate into the TCA cycle. PDH dysfunction is documented in ME/CFS.

\paragraph{Evidence.} Preliminary. Case reports and small studies suggest high-dose thiamine may help a subset of ME/CFS patients. One Italian study used 600--1800~mg daily with significant benefit in chronic fatigue.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Thiamine HCl:} Standard form; limited absorption
    \item \textbf{Benfotiamine:} Fat-soluble; better absorbed; doesn't cross BBB well
    \item \textbf{Thiamine TTFD (Allithiamine):} Lipid-soluble; crosses BBB; may be most relevant for ME/CFS
\end{itemize}

\paragraph{Dosing.} Standard: 50--100~mg. High-dose protocols: 300--1800~mg daily (under medical supervision).

\subsection{Riboflavin (B2)}

\paragraph{Rationale.} Precursor to FAD, essential for Complex II (succinate dehydrogenase) and fatty acid oxidation.

\paragraph{Evidence.} Theoretical; studied in migraine prevention (400~mg daily).

\paragraph{Dosing.} 25--400~mg daily. Harmless neon yellow urine at higher doses.

\subsection{Niacin/Niacinamide (B3)}

See NAD$^+$ precursors above.

\subsection{Pyridoxine/P5P (B6)}

\paragraph{Rationale.} Cofactor for neurotransmitter synthesis (serotonin, dopamine, GABA). The active form pyridoxal-5-phosphate (P5P) is the essential cofactor for aromatic L-amino acid decarboxylase, the enzyme that converts 5-HTP to serotonin. In the enterochromaffin-vagal pathway~\cite{WirthScheibenbogen2025Neurotransmitter, Barton2025}, adequate P5P is therefore required for the final step of gut serotonin synthesis and downstream vagal activation. B6 is water-soluble and may be depleted by chronic inflammation and GI malabsorption---conditions common in ME/CFS; B6 status has not been systematically studied in ME/CFS cohorts.

\paragraph{Dosing.} 25--100~mg daily. P5P (pyridoxal-5-phosphate) is the active form and may be better for those with conversion issues. When supporting gut serotonin synthesis (see Section~\ref{sec:5-htp}), pair with 5-HTP or tryptophan supplementation.

\paragraph{Cautions.} High doses ($>$200~mg/day chronically) can cause peripheral neuropathy. Assess RBC P5P levels if possible, as this is a more sensitive marker than serum B6.

\subsection{Folate (B9)}

\paragraph{Rationale.} Essential for methylation and DNA synthesis.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Folic acid:} Synthetic; requires conversion; some people have MTHFR variants impairing conversion
    \item \textbf{Methylfolate (5-MTHF):} Active form; bypasses MTHFR; often preferred
    \item \textbf{Folinic acid:} Intermediate; doesn't require MTHFR
\end{itemize}

\paragraph{Dosing.} 400--1000~mcg daily; higher doses (up to 15~mg) used for specific indications.

\paragraph{Cautions.} Must be balanced with B12; folate alone can mask B12 deficiency.

\subsection{Cobalamin (B12)}

\paragraph{Rationale.} Essential for methylation, nerve function, and energy metabolism.

\paragraph{Evidence.} Low--Moderate. Some ME/CFS patients respond dramatically to B12, especially sublingual or injectable forms; others show no benefit.

\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Cyanocobalamin:} Cheapest; requires conversion; contains cyanide moiety (trivial amount)
    \item \textbf{Methylcobalamin:} Active methylated form; supports methylation
    \item \textbf{Adenosylcobalamin:} Active form used in mitochondria
    \item \textbf{Hydroxocobalamin:} Well-retained; often used in injections
\end{itemize}

\paragraph{Dosing.} Oral: 1000--5000~mcg sublingual daily. Injections: 1000~mcg weekly to monthly (requires prescription in most countries).

\paragraph{Note on Testing.} Serum B12 is a poor marker of tissue status. Methylmalonic acid (MMA) and homocysteine are more sensitive.

\section{Vitamin D}
\label{sec:vitamin-d}

\paragraph{Rationale.} Vitamin D is actually a hormone with effects on:
\begin{itemize}
    \item Immune regulation (relevant to ME/CFS immune dysfunction)
    \item Muscle function
    \item Mood
    \item Bone health
\end{itemize}

Deficiency is common in ME/CFS patients (often housebound with limited sun exposure).

\paragraph{Evidence.} Moderate for general health; limited ME/CFS-specific data.

\paragraph{Target Levels.} Controversy exists:
\begin{itemize}
    \item Conventional: 30--50~ng/mL (75--125~nmol/L)
    \item Some ME/CFS practitioners target: 50--80~ng/mL
    \item Toxicity typically $>$150~ng/mL
\end{itemize}

\paragraph{Dosing.}
\begin{itemize}
    \item Maintenance: 1000--2000~IU daily
    \item Deficiency correction: 5000--10,000~IU daily for 8--12 weeks, then retest
    \item Take with fat-containing meal
\end{itemize}

\paragraph{Cofactors.} Vitamin D requires cofactors for proper function:
\begin{itemize}
    \item \textbf{Magnesium:} Required for vitamin D activation
    \item \textbf{Vitamin K2:} Directs calcium to bones (away from arteries)
    \item \textbf{Vitamin A:} Balances vitamin D effects
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Immune regulatory and gene expression effects operate through nuclear receptor signaling with minimal direct energy cost. Long half-life means infrequent dosing keeps metabolic processing burden low. No direct energy provision.
\revend

\section{Amino Acids}
\label{sec:amino-acids}

\subsection{Taurine}

\paragraph{Rationale.}
\begin{itemize}
    \item Mitochondrial membrane stabilization
    \item Antioxidant
    \item Supports bile acid conjugation
    \item May support cardiac and nervous system function
    \item Autonomic support
\end{itemize}

\paragraph{Evidence.} Theoretical for ME/CFS; widely used.

\paragraph{Dosing.} 500--3000~mg daily.

\rev{5}
\paragraph{Energy Profile.}
Category A--B (provider to neutral). Conditionally essential amino acid that stabilizes mitochondrial membranes and regulates calcium homeostasis---both critical for mitochondrial energy production. Also functions as an osmolyte supporting cell volume regulation with minimal ATP cost. May improve mitochondrial efficiency without direct ATP provision.
\revend

\subsection{Glycine}

\paragraph{Rationale.}
\begin{itemize}
    \item Glutathione precursor (with NAC)
    \item Inhibitory neurotransmitter (calming)
    \item Supports collagen synthesis
    \item May improve sleep quality
\end{itemize}

\paragraph{Dosing.} 1--3~g daily; 3~g before bed for sleep.

\rev{5}
\paragraph{Energy Profile.}
Category A--B (provider to neutral). Glycine is one of three glutathione precursors (alongside cysteine and glutamate; NAC is typically rate-limiting for cysteine availability), supporting mitochondrial protection from oxidative damage. Also serves as a conjugation substrate for Phase~II hepatic detoxification (glycine N-acyltransferase), potentially reducing the metabolic burden of processing other supplements. Minimal processing cost as an endogenous amino acid.
\revend

\subsection{L-Glutamine}

\paragraph{Rationale.}
\begin{itemize}
    \item Gut barrier support
    \item Immune cell fuel
    \item Glutathione precursor
\end{itemize}

\paragraph{Dosing.} 5--15~g daily for gut support.

\paragraph{Cautions.} Some patients with neurological sensitivity may not tolerate glutamine (converts to glutamate).

\rev{5}
\paragraph{Energy Profile.}
Category A--B (provider to neutral). Primary fuel for enterocytes (intestinal cells) and immune cells. Provides energy substrate for gut barrier maintenance, potentially reducing the energy cost of intestinal permeability and immune activation from microbial translocation. Also serves as gluconeogenic substrate when glucose is depleted.
\revend

\subsection{5-HTP (5-Hydroxytryptophan)}
\label{sec:5-htp}

\paragraph{Rationale.} 5-HTP is the immediate precursor to serotonin, produced from tryptophan by tryptophan hydroxylase. Supplementing with 5-HTP bypasses the IDO2 (and IDO1) diversion point~\cite{WirthScheibenbogen2025Neurotransmitter}: in ME/CFS, IDO2 is the primary enzyme shunting tryptophan into the kynurenine pathway rather than toward serotonin synthesis~\cite{phair2019ido}. Because 5-HTP is not a substrate for IDO enzymes, it directly enters the serotonin synthesis pathway regardless of IDO activity. In the gut, enterochromaffin cells convert 5-HTP to serotonin, which signals through vagal afferents to regulate autonomic tone~\cite{Barton2025}. Supplemental 5-HTP may therefore partially compensate for impaired peripheral serotonin synthesis in ME/CFS patients with gut dysbiosis.

\paragraph{Evidence.} Theoretical for ME/CFS-specific gut-vagal pathway; limited direct trial evidence. Best supported by mechanistic extrapolation from IDO2 pathway research in Long COVID and the enterochromaffin-vagal model~\cite{WirthScheibenbogen2025Neurotransmitter}.

\paragraph{Dosing.} 50--100~mg two to three times daily. Start at 50~mg once daily and titrate slowly. Requires adequate P5P (Section~\ref{sec:vitamins}) as cofactor for the decarboxylase step.

\rev{5}
\paragraph{Cautions.}
\begin{itemize}
    \item \textbf{Serotonin syndrome}: Contraindicated with SSRIs, SNRIs, MAOIs, tramadol, or other serotonergic drugs. This is a serious risk---medical supervision required if on any of these medications.
    \item Peripheral decarboxylation can cause systemic serotonin effects (GI cramping, nausea, diarrhea) before central effects appear.
    \item Avoid in pregnancy.
    \item Start very low if histamine-sensitive (serotonin can release histamine).
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category B--C (neutral to mildly demanding). Conversion to serotonin via aromatic L-amino acid decarboxylase requires pyridoxal phosphate (vitamin B6) cofactor. The serotonin synthesis pathway itself is not particularly energy-intensive, but serotonergic effects can trigger mast cell activation in susceptible patients (MCAS comorbidity), and immune activation is energy-costly. Caution: serotonin syndrome risk with SSRIs represents a potentially catastrophic energy crisis.
\revend

\section{Practical Supplement Protocols}
\label{sec:protocols}

Given the complexity, here are evidence-informed starting points organized by symptom cluster and budget.

\subsection{Minimal Cost Protocol (Under \$30/month)}

For patients with limited resources:

\begin{enumerate}
    \item \textbf{Electrolytes:} Salt + potassium salt (Nu-Salt) + DIY rehydration (\$5/month)
    \item \textbf{Magnesium glycinate:} 200--400~mg at bedtime (\$10/month)
    \item \textbf{B-complex:} Basic B-complex with methylated B12/folate (\$10/month)
    \item \textbf{Vitamin D3:} 2000--5000~IU daily (\$5/month)
\end{enumerate}

This addresses the most common deficiencies and supports autonomic function. All components are Category A (net energy providers) or B (energy-neutral), making this protocol safe for all severity levels including very severe patients. See Section~\ref{subsec:energy-categories} for the energy category framework.

\subsection{Moderate Protocol (\$50--100/month)}

Adding mitochondrial and antioxidant support. \textbf{Severity note}: For moderate-to-severe patients, establish the minimal protocol for at least 4 weeks before adding these supplements. NADH and CoQ10 may be introduced together as a pair (consistent with the Castro-Marrero protocol); introduce each subsequent supplement individually with 1--2 week observation windows. All additions are Category A or B.

\rev{5}
\begin{enumerate}
    \item Everything in minimal protocol, plus:
    \item \textbf{NADH + CoQ10 (ubiquinol):} 20~mg NADH + 200~mg CoQ10 daily---the Castro-Marrero protocol~\cite{CastroMarrero2021fatigue} (\$40--60/month combined)
    \item \textbf{NAC:} 600--1200~mg daily (\$10--15/month)
    \item \textbf{Omega-3:} 2~g EPA/DHA daily (\$15--20/month)
    \item \textbf{D-ribose:} 5--10~g daily (\$15--20/month)
\end{enumerate}
\revend

\subsection{Comprehensive Protocol (\$100--200/month)}

For mild-to-moderate patients who can afford broader support. \textbf{Severity note}: For severe and very severe patients, this protocol should only be attempted after documented tolerance of the moderate protocol over 8+ weeks. Physician supervision recommended. While all components are Category A--B, the cumulative hepatic processing load of multiple supplements warrants caution in energy-depleted patients (see Section~\ref{subsec:treatment-safety-imperative}).

\begin{enumerate}
    \item Everything above, plus:
    \item \textbf{NR or NMN:} 300--500~mg daily (\$40--60/month)
    \item \textbf{Acetyl-L-carnitine:} 1000~mg daily (\$15/month)
    \item \textbf{Alpha-lipoic acid:} 300~mg daily (\$10/month)
    \item \textbf{Curcumin (enhanced):} 500~mg daily (\$15--20/month)
\end{enumerate}

\subsection{By Symptom Cluster}

\paragraph{Predominant Orthostatic Symptoms.}
\begin{itemize}
    \item Electrolytes (priority)
    \item Magnesium
    \item Taurine
    \item Licorice root (caution: raises BP)
\end{itemize}

\rev{5}
\paragraph{Predominant Cognitive Symptoms.}
\begin{itemize}
    \item NADH + CoQ10 (cognitive fatigue improvement; Castro-Marrero 2021 RCT, n=207)
    \item Magnesium L-threonate
    \item Acetyl-L-carnitine
    \item Omega-3 (high DHA)
    \item NR/NMN
    \item Creatine
\end{itemize}
\revend

\paragraph{Predominant Immune/Inflammatory Symptoms.}
\begin{itemize}
    \item NAC
    \item Omega-3
    \item Curcumin
    \item Quercetin (especially if mast cell component)
    \item Vitamin D (optimize)
\end{itemize}

\rev{5}
\paragraph{Predominant Muscle/Fatigue Symptoms.}
\begin{itemize}
    \item NADH + CoQ10 (electron transport chain support; Section~\ref{subsec:mito-combination-guide})
    \item D-ribose (ATP backbone resynthesis)
    \item L-carnitine (fatty acid transport into mitochondria)
    \item Creatine (rapid ATP buffering via phosphocreatine)
    \item Magnesium malate (cofactor for >300 enzymatic reactions; malate feeds the TCA cycle)
\end{itemize}
\revend

\subsection{Introduction Strategy}

\rev{5}
\begin{warning}[One at a Time]
ME/CFS patients often have multiple sensitivities (clinical experience). Introducing multiple supplements simultaneously makes it impossible to identify what helps or harms. Start one new supplement at a time, at low dose, and wait 1--2 weeks before adding another. Keep a symptom diary.
\end{warning}

\rev{5}
\paragraph{Suggested Order.}
\begin{enumerate}
    \item Electrolytes and magnesium (foundational; generally well-tolerated with gradual titration---see Section~\ref{subsec:electrolyte-initiation-caution})
    \item B vitamins (essential cofactors)
    \item CoQ10 + NADH (exception to the one-at-a-time principle: the RCT evidence is for the combination, not the individual components; see Section~\ref{subsec:mito-combination-guide}). NADH requires an empty stomach 30--60~min before breakfast; CoQ10 is taken with a fat-containing meal---these timing constraints naturally separate the two supplements
    \item NAC (antioxidant; watch for sulfur sensitivity)
    \item Additional mitochondrial support (NAD$^+$ precursors, D-ribose, ALCAR) based on response
\end{enumerate}
\revend

\section{Additional Supplements}
\label{sec:additional-supplements}

\subsection{Medium-Chain Triglycerides (MCT)}

\paragraph{Rationale.} MCTs bypass normal fat digestion and are converted directly to ketones by the liver. Ketones provide alternative brain and muscle fuel, potentially bypassing impaired glucose metabolism.

\paragraph{Evidence.} Theoretical for ME/CFS; moderate for cognitive support in other conditions.

\paragraph{Dosing.} Start with 1 teaspoon and increase slowly to 1--2 tablespoons daily. Rapid introduction causes GI distress.

\rev{5}
\paragraph{Energy Profile.}
Category A (net energy provider). Converts directly to ketone bodies in the liver, bypassing the carnitine transport system and providing immediate alternative fuel for brain and muscle. Ketogenesis from MCTs is metabolically efficient compared to long-chain fatty acid oxidation. Provides energy substrate that bypasses potentially impaired glucose metabolism pathways.
\revend

\subsection{Resveratrol}

\paragraph{Rationale.} Activates sirtuins and AMPK; promotes mitochondrial biogenesis; antioxidant.

\paragraph{Evidence.} Preliminary; animal studies promising but human data limited.

\paragraph{Dosing.} 150--500~mg daily; trans-resveratrol is the active form.

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Resveratrol activates SIRT1 and AMPK signaling pathways that promote mitochondrial biogenesis over weeks to months, but does not directly provide ATP substrates. Requires hepatic CYP450 metabolism (CYP1A2, CYP3A4) for glucuronidation and sulfation, imposing modest metabolic processing cost. Net effect is neutral in the short term; potential long-term energy benefit via increased mitochondrial mass is speculative.
\revend

\subsection{Melatonin}

\paragraph{Rationale.}
\begin{itemize}
    \item Sleep initiation and circadian rhythm regulation
    \item Potent antioxidant (especially mitochondrial)
    \item Immune modulation
    \item Anti-inflammatory
\end{itemize}

\paragraph{Evidence.} Moderate for sleep; theoretical for other effects in ME/CFS.

\paragraph{Dosing.}
\begin{itemize}
    \item Sleep: 0.5--3~mg, 30--60 minutes before bed
    \item Some protocols use higher doses (5--20~mg) for antioxidant effects
    \item Extended-release forms for sleep maintenance issues
\end{itemize}

\paragraph{Note.} ``Less is more'' for sleep---higher doses can paradoxically worsen sleep quality. Start at 0.5~mg.

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Endogenous hormone with minimal hepatic processing at physiological replacement doses. Sleep improvement may indirectly enhance energy restoration during rest. No significant metabolic burden.
\revend

\subsection{Pregnenolone}
\label{subsec:pregnenolone}

Pregnenolone is a neurosteroid synthesized from cholesterol, serving as the precursor for all other steroid hormones including DHEA, progesterone, cortisol, and testosterone.

\paragraph{Rationale.}
\begin{itemize}
    \item \textbf{TRPM3 activation}: Pregnenolone sulfate (PregS) is an endogenous activator of TRPM3 ion channels (see Section~\ref{sec:trpm3-hypotheses})
    \item \textbf{Cognitive support}: Enhances memory formation and retrieval; neuroprotective
    \item \textbf{Neuronal myelination}: Supports myelin synthesis and maintenance
    \item \textbf{Fatigue relevance}: Low pregnenolone levels have been reported in ME/CFS patients
    \item \textbf{Immune modulation}: Modulates neurotransmitter-gated channels in immune cells
\end{itemize}

\paragraph{TRPM3 Connection.}
Reduced TRPM3 function in ME/CFS NK cells has been demonstrated using pregnenolone sulfate stimulation. TRPM3 channels in ME/CFS patients show significantly reduced calcium flux when stimulated with PregS. Theoretically, increasing pregnenolone availability might partially compensate for channel dysfunction by increasing agonist concentration, though this remains speculative. The established finding is that TRPM3 dysfunction can be partially restored by low-dose naltrexone (LDN), which may explain LDN's efficacy in some ME/CFS patients.

\paragraph{Evidence.}
\begin{itemize}
    \item Low pregnenolone levels documented in ME/CFS (clinical observations)
    \item No randomized controlled trials specifically in ME/CFS
    \item General evidence for cognitive and energy benefits in aging populations
    \item Theoretical basis from TRPM3 research
\end{itemize}

\paragraph{Dosing.}
\begin{itemize}
    \item Typical starting dose: 10--30~mg daily
    \item Some protocols use 50--100~mg daily
    \item Take in the morning (may be stimulating)
    \item Sublingual forms may have better absorption
\end{itemize}

\paragraph{Cautions.}
\begin{itemize}
    \item May affect hormone levels---caution in hormone-sensitive conditions
    \item Can be converted to androgens or estrogens depending on individual biochemistry
    \item Monitor for mood changes (anxiety, irritability)
    \item Avoid in pregnancy
    \item Consider baseline hormone testing before long-term use
\end{itemize}

\rev{5}
\begin{speculation}[Pregnenolone + LDN Synergy]
\label{spec:pregnenolone-ldn-synergy}
\textbf{Certainty: 0.20.} No clinical trials have evaluated this combination; rationale is purely mechanistic extrapolation from separately-characterised TRPM3 effects of each agent.
\revend

Both pregnenolone sulfate and low-dose naltrexone influence TRPM3 function in ME/CFS, potentially through different mechanisms. Pregnenolone sulfate is a direct TRPM3 agonist, while LDN appears to restore TRPM3 responsiveness (mechanism not fully characterized). Theoretically, combining adequate pregnenolone levels with LDN might optimize TRPM3 function---the channel would be more responsive (LDN effect) and have adequate agonist (pregnenolone). No clinical trials have evaluated this combination, but given the favorable safety profiles of both, it represents a rational empirical approach for patients who have partial response to LDN alone.
\end{speculation}

\rev{5}
\paragraph{Energy Profile.}
Category B--C (neutral to mildly demanding). Neurosteroid requiring hepatic CYP-mediated metabolism for conversion to downstream hormones (DHEA, progesterone, cortisol). Steroidogenesis is moderately energy-intensive. At supplemental doses, processing burden is modest but non-negligible.
\revend

\subsection{Zinc}
\label{subsec:zinc}

\paragraph{Rationale.} Zinc is a trace mineral essential for gut barrier integrity, antioxidant protection in enterocytes, and neurotransmitter function. In the context of the enterochromaffin-vagal pathway~\cite{WirthScheibenbogen2025Neurotransmitter}, zinc supports multiple nodes: (1) tight junction integrity reduces intestinal permeability and bacterial translocation; (2) antioxidant mechanisms in enterocytes that may protect enterochromaffin cells; (3) neurotransmitter synthesis and autonomic nerve function. Zinc is commonly depleted in ME/CFS due to malabsorption, chronic inflammation (acute-phase redistribution to liver), and reduced dietary diversity from fatigue limiting food preparation.

\rev{5}
\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Zinc glycinate}: Well-absorbed, gentle on GI tract
    \item \textbf{Zinc carnosine}: Combines zinc with carnosine dipeptide; additional mucosal healing properties; preferred for gut repair (see Chapter~\ref{ch:action-mild-moderate})
    \item \textbf{Zinc citrate or picolinate}: Good bioavailability alternatives
    \item \textbf{Avoid zinc oxide}: Poor bioavailability
\end{itemize}
\revend

\paragraph{Dosing.} 15--30~mg elemental zinc daily. Zinc carnosine 75~mg provides approximately 16~mg elemental zinc. Take away from or between meals for maximum absorption.

\rev{5}
\paragraph{Cautions.}
\begin{itemize}
    \item Long-term zinc supplementation above 40~mg elemental zinc/day can deplete copper---monitor copper status with extended use or consider zinc-copper combination
    \item Take separately from calcium, magnesium, and iron (compete for absorption)
    \item Check serum zinc or RBC zinc before high-dose supplementation; levels may be misleadingly normal during inflammation despite tissue deficiency
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Essential trace mineral and enzymatic cofactor absorbed through intestinal metallothionein pathways. Processing demands minimal at supplemental doses (15--30~mg). Supports immune function and wound healing without significant direct energy cost.
\revend

\section{Probiotics and Gut Health}
\label{sec:probiotics}

Given the documented gut microbiome abnormalities in ME/CFS~\cite{Giloteaux2016, ButyrateDeficiency2023}, gut-targeted interventions may help.

\subsection{Probiotics}

\paragraph{Rationale.} May restore beneficial bacteria, reduce gut permeability, modulate immune function, and reduce systemic inflammation.

\paragraph{Evidence.} Low--Moderate. Two small RCTs in ME/CFS showed modest benefit.

\paragraph{Strain Selection.}
\begin{itemize}
    \item \textbf{\textit{Lactobacillus} and \textit{Bifidobacterium} species:} Most studied; generally safe
    \item \textbf{\textit{Saccharomyces boulardii}:} Yeast-based; may help after antibiotics
    \item \textbf{Soil-based organisms (SBOs):} More controversial; some find helpful
\end{itemize}

\paragraph{Practical Notes.}
\begin{itemize}
    \item Start low; die-off reactions possible
    \item May take 4--8 weeks to assess
    \item Quality varies enormously between brands
    \item Refrigerated products generally more viable
\end{itemize}

\rev{5}
\paragraph{Energy Profile.}
Category B (energy-neutral). Live organisms that colonize the gut without requiring host metabolic processing. May improve nutrient absorption and reduce gut inflammation energy costs. No direct ATP demand from supplementation.
\revend

\subsection{Prebiotics}

\rev{5}
Prebiotics feed beneficial bacteria. Options include:
\begin{itemize}
    \item Partially hydrolyzed guar gum (PHGG)
    \item \textbf{Resistant starch} (see below for type selection)
    \item Inulin and FOS (can cause bloating)
    \item Acacia fiber (generally well-tolerated)
\end{itemize}

\rev{5}
\paragraph{Resistant Starch Types.} Not all resistant starch is equivalent. Two types differ in where they are fermented in the colon:
\begin{itemize}
    \item \textbf{RS2 (raw potato starch, green banana flour):} Fermented primarily in the proximal colon; supports butyrate-producing bacterial communities~\cite{ButyrateDeficiency2023}. Best starting choice. Dose: 1--2 tablespoons raw potato starch daily.
    \item \textbf{RS3 (cooked and cooled rice, potatoes, pasta):} More distal colonic fermentation; may complement RS2 for supporting butyrate production throughout the colon~\cite{ButyrateDeficiency2023}. Introduce after RS2 is tolerated.
\end{itemize}
\revend

\paragraph{Caution.} Prebiotics are contraindicated in active SIBO---fermentable substrate feeds small intestinal overgrowth and can severely worsen symptoms. Test for SIBO (breath test) before introducing resistant starch. Start with very small amounts (1 teaspoon) and increase slowly over 4--6 weeks.

\subsection{Butyrate and Tributyrin}
\label{subsec:butyrate}

\paragraph{Rationale.} Butyrate is the primary short-chain fatty acid (SCFA) fuel for colonocytes and a key upstream signal in the enterochromaffin-vagal pathway~\cite{Barton2025, WirthScheibenbogen2025Neurotransmitter}. Gut microbiome studies show ME/CFS patients have significantly reduced butyrate-producing bacterial capacity~\cite{ButyrateDeficiency2023}, correlating with fatigue severity. Butyrate supplementation aims to compensate for this deficit, enhancing enterochromaffin serotonin synthesis and downstream vagal afferent signaling. Butyrate also directly activates SCFA receptors (GPR41/43) on vagal neurons, providing a parallel pathway independent of serotonin.

\rev{5}
\paragraph{Forms.}
\begin{itemize}
    \item \textbf{Sodium butyrate:} Fast-absorbing in proximal colon; well-studied; strong odour (similar to parmesan). Dose: 500~mg--2~g three times daily with meals.
    \item \textbf{Tributyrin:} A triglyceride of butyric acid, more lipophilic; slower release for sustained distal colon exposure. Dose: 1--2~g three times daily. Lower odour than sodium butyrate.
    \item \textbf{Calcium/magnesium butyrate:} Combined mineral-butyrate salt; another option with good tolerability.
    \item \textbf{Dietary butyrate:} Cooked and cooled starchy foods (RS3), high-fat fermented dairy (butter, ghee contain short-chain fats), and adequate fiber intake to feed butyrate producers.
\end{itemize}
\revend

\paragraph{Evidence.} Direct ME/CFS supplementation trials are lacking. Mechanistic rationale is strong based on documented butyrate-producer deficiency in ME/CFS~\cite{ButyrateDeficiency2023} and established butyrate-enterochromaffin-vagal signaling~\cite{Barton2025}. Safety profile is excellent.

\paragraph{Combination Approach.} For targeting the enterochromaffin-vagal pathway, combining butyrate with 5-HTP and P5P (Section~\ref{sec:5-htp}) addresses both upstream substrate supply (butyrate) and downstream serotonin synthesis (5-HTP + P5P cofactor). For a phased clinical protocol incorporating these agents with sequencing rationale and dosing, see Section~\ref{subsubsec:gut-vagal-protocol} in Chapter~\ref{ch:action-mild-moderate}.

\paragraph{Cautions.} Very safe; main side effect is GI discomfort (bloating, loose stool) if dose escalated too quickly. Start low and increase gradually. The odour of sodium butyrate may be off-putting---encapsulated forms minimize this.

\rev{5}
\paragraph{Energy Profile.}
Category A--B (provider). Short-chain fatty acid that serves as primary energy source for colonocytes (intestinal cells). Direct mitochondrial fuel that supports gut barrier integrity. By providing colonocyte fuel, butyrate spares systemic energy resources for other organs. Processing demands minimal.
\revend

\section{Supplement Quality and Safety}
\label{sec:supplement-quality}

\subsection{Quality Considerations}

Supplements are minimally regulated. Quality varies enormously:
\begin{itemize}
    \item \textbf{Third-party testing:} Look for NSF, USP, ConsumerLab, or IFOS certification
    \item \textbf{GMP compliance:} Minimum standard; not sufficient alone
    \item \textbf{Bioavailability:} Cheap forms may not be absorbed
    \item \textbf{Contaminants:} Heavy metals, pesticides, adulterants possible
\end{itemize}

\subsection{Drug Interactions}

Common interactions to be aware of:
\begin{itemize}
    \item \textbf{Blood thinners (warfarin):} CoQ10, omega-3, vitamin E, curcumin can affect clotting
    \item \textbf{Blood pressure medications:} Potassium, magnesium, licorice can interact
    \item \textbf{Thyroid medications:} Take separately from calcium, magnesium, iron
    \item \textbf{Immunosuppressants:} Immune-modulating supplements may interfere
\end{itemize}

\subsection{When to Stop}

Discontinue a supplement if:
\begin{itemize}
    \item Clear worsening of symptoms
    \item No benefit after adequate trial (typically 8--12 weeks)
    \item Financial burden outweighs uncertain benefit
    \item Interactions with new medications
\end{itemize}

\begin{observation}[The Supplement Trap]
Many ME/CFS patients accumulate large, expensive supplement regimens over time without systematically evaluating benefit. Periodically reassess: stop everything non-essential for 2--4 weeks, then reintroduce one at a time. You may discover that many supplements you've been taking for years provide no discernible benefit.
\end{observation}


\section{Conclusion}

Supplements can play a supportive role in ME/CFS management, but expectations should be realistic:
\begin{itemize}
    \item \textbf{No supplement cures ME/CFS}
    \item \textbf{Effects are typically modest}---10--20\% symptom improvement is a good outcome
    \item \textbf{Response varies enormously} between individuals
    \item \textbf{Cost adds up}---prioritize based on evidence and your specific symptoms
    \item \textbf{Foundation first:} Electrolytes, magnesium, B vitamins, and vitamin D before exotic interventions
\end{itemize}

\rev{5}
The supplements most likely to help, based on current evidence and mechanistic plausibility, are:
\begin{enumerate}
    \item Electrolytes (especially if orthostatic symptoms)
    \item NADH + CoQ10 combination (strongest ME/CFS RCT evidence for mitochondrial support)
    \item NAC (antioxidant, glutathione support)
    \item Magnesium (ubiquitous cofactor, often deficient)
    \item NAD$^+$ precursors (emerging evidence for cellular NAD$^+$ pool replenishment)
\end{enumerate}
\revend

Work with a knowledgeable healthcare provider when possible, especially for higher-dose protocols or if taking multiple medications.

