% FILE: Endocrine dysfunction — HPA axis, thyroid, cortisol dysregulation, hormonal abnormalities, metabolic dysregulation
\chapter{Endocrine and Metabolic Dysfunction}
\label{ch:endocrine}

Endocrine dysfunction represents a critical but often overlooked dimension of ME/CFS pathophysiology. The endocrine system orchestrates fundamental physiological processes including stress response, energy metabolism, circadian rhythms, reproduction, and immune modulation. Disruption of these hormonal axes provides mechanistic explanations for the multi-system nature of ME/CFS symptoms and connects seemingly disparate clinical features into a coherent pathophysiological framework.

The landmark NIH deep phenotyping study by Walitt et al.\ (2024) documented central nervous system dysfunction with direct implications for neuroendocrine regulation~\cite{walitt2024deep}. Complementing this neurological evidence, recent studies have identified specific endocrine abnormalities spanning the hypothalamic-pituitary-adrenal (HPA) axis, thyroid function, sex hormones, growth factors, glucose metabolism, and circadian regulation. These findings reveal that ME/CFS involves coordinated dysfunction across multiple endocrine systems rather than isolated hormonal deficits.

This chapter examines six major endocrine systems implicated in ME/CFS pathophysiology. The HPA axis shows characteristic blunting with hypersensitive feedback, contributing to stress intolerance and immune dysregulation. Thyroid function abnormalities, particularly the ``Low T3 Syndrome,'' affect cellular metabolism despite normal TSH levels. Sex hormone dysregulation explains the striking female predominance and menstrual cycle exacerbations. Growth hormone and IGF-1 deficiencies contribute to muscle dysfunction and metabolic impairment. Insulin resistance and cerebral glucose hypometabolism connect to the energy deficit discussed in Chapter~\ref{ch:energy-metabolism}. Finally, circadian rhythm disruption integrates with the sleep abnormalities and autonomic dysfunction detailed in Chapters~\ref{ch:neurological} and~\ref{ch:cardiovascular}. Chapter~\ref{ch:integrative-models} synthesizes these endocrine connections with immune and metabolic systems into comprehensive models of ME/CFS pathophysiology.

Understanding endocrine dysfunction is essential for several reasons. First, hormonal abnormalities provide measurable biomarkers for diagnosis and disease monitoring. Second, endocrine pathways mechanistically link immune activation (Chapter~\ref{ch:immune-dysfunction}) to metabolic dysfunction (Chapter~\ref{ch:energy-metabolism}). Third, hormonal dysregulation explains symptom patterns such as post-exertional malaise, orthostatic intolerance, and cognitive impairment that define the clinical presentation. Finally, endocrine interventions represent potential therapeutic targets, though current evidence remains mixed and requires careful evaluation.

\section{Hypothalamic-Pituitary-Adrenal (HPA) Axis}
\label{sec:hpa-axis}

The hypothalamic-pituitary-adrenal axis represents one of the most extensively studied endocrine systems in ME/CFS, yet paradoxically remains among the most controversial. Through glucocorticoid signaling, the HPA axis coordinates the body's stress response and regulates immune function while maintaining glucose homeostasis and energy metabolism. It further modulates circadian rhythms, sleep-wake cycles, cognitive function, and mood regulation. Given these critical roles, HPA dysfunction provides a plausible mechanism linking the diverse symptoms of ME/CFS.

% Insert Figure: Normal HPA Axis Function
\input{figures/fig-hpa-axis-normal.tex}

% Insert Figure: ME/CFS HPA Axis Dysregulation
\input{figures/fig-hpa-axis-mecfs.tex}

Figures~\ref{fig:hpa-axis-normal} and~\ref{fig:hpa-axis-mecfs} illustrate the characteristic pattern of HPA axis dysfunction observed in ME/CFS. Unlike the robust circadian cortisol rhythm and responsive feedback regulation seen in healthy individuals, ME/CFS patients demonstrate a distinct pattern of dysregulation. This involves blunted corticotropin-releasing hormone (CRH) secretion from the hypothalamus, reduced adrenocorticotropic hormone (ACTH) response from the pituitary, flattened diurnal cortisol rhythm with loss of the normal morning peak, and paradoxically enhanced negative feedback sensitivity. This constellation of abnormalities distinguishes ME/CFS from both healthy states and primary adrenal insufficiency (Addison's disease), suggesting a unique form of central HPA axis hypofunction.

\subsection{HPA Axis Abnormalities}

\subsubsection{Cortisol Dysregulation Patterns}

\begin{observation}[Characteristic Cortisol Pattern in ME/CFS]
\label{obs:cortisol-pattern}
Multiple studies document a consistent pattern of cortisol abnormalities in ME/CFS patients that differs qualitatively from both healthy individuals and patients with primary adrenal disorders. Baseline cortisol levels tend to be lower, though typically remaining within the broad ``normal'' laboratory reference range. The diurnal cortisol rhythm shows flattening, with reduced amplitude between morning and evening values. The cortisol awakening response (CAR)---the normal sharp rise in cortisol during the first 30--60 minutes after waking---appears attenuated. Additionally, cortisol responses to physiological and psychological stressors are blunted, despite appropriate ACTH response to exogenous CRH stimulation in some studies~\cite{pipper2024steroid}.
\end{observation}

The NIH deep phenotyping study by Walitt et al.\ identified neuroendocrine abnormalities consistent with HPA axis dysfunction, including altered catecholamine metabolism that affects upstream regulation of the HPA axis~\cite{walitt2024deep}. The reduced central catecholamines documented in cerebrospinal fluid may contribute to impaired hypothalamic CRH release, providing a mechanistic link between neurological and endocrine dysfunction.

Recent sex-stratified analysis by Pipper et al.\ (2024) revealed that cortisol dysregulation patterns differ significantly between male and female ME/CFS patients and vary by disease severity~\cite{pipper2024steroid}. Female patients with severe ME/CFS demonstrated elevated 11-deoxycortisol (a cortisol precursor) and 17$\alpha$-hydroxyprogesterone, suggesting impaired final enzymatic steps in cortisol synthesis. Male patients with mild to moderate disease showed frankly reduced cortisol and corticosterone levels but paradoxically elevated progesterone. These findings indicate that HPA dysfunction may involve enzyme deficiencies in steroidogenesis rather than simple hypothalamic-pituitary signaling deficits.

\subsubsection{ACTH and CRH Abnormalities}

The central components of the HPA axis---CRH from the hypothalamus and ACTH from the pituitary---show complex abnormalities that do not fit simple models of endocrine failure. Studies employing CRH stimulation tests have yielded inconsistent results. Some report normal ACTH and cortisol responses to exogenous CRH administration, while others document blunted ACTH responses despite adequate CRH stimulation. Still others find normal ACTH responses but reduced cortisol output, suggesting adrenal hyposensitivity. These inconsistencies likely reflect the heterogeneity of ME/CFS patient populations, differences in disease duration and severity, and the limitations of single-timepoint testing to capture dynamic regulatory dysfunction.

The most consistent finding across studies is evidence of enhanced negative feedback sensitivity. Dexamethasone suppression tests demonstrate that low doses of synthetic glucocorticoid produce greater and more prolonged suppression of cortisol secretion in ME/CFS patients compared to controls. This suggests that the hypothalamus and pituitary remain exquisitely sensitive to glucocorticoid feedback signals, inappropriately dampening HPA axis output even when cortisol levels are already low-normal. This pattern resembles the neuroendocrine adaptation seen in chronic stress conditions but persists inappropriately in ME/CFS despite the clinical need for robust stress responses.

\subsubsection{Diurnal Rhythm Disruption}

\begin{observation}[Loss of Cortisol Circadian Amplitude]
\label{obs:cortisol-rhythm}
The diurnal cortisol rhythm represents one of the most robust and well-characterized circadian processes in human physiology, yet ME/CFS patients consistently demonstrate flattening of this rhythm. Healthy individuals show a sharp cortisol peak within 30--60 minutes of waking (cortisol awakening response), followed by progressive decline throughout the day, reaching a nadir around midnight, and beginning to rise again in the early morning hours (3--4 AM) in anticipation of waking. In contrast, ME/CFS patients show reduced morning cortisol peak (blunted CAR), less pronounced decline during the day (flatter slope), and reduced overall amplitude (difference between peak and nadir), resulting in a ``flattened'' 24-hour pattern~\cite{cambras2018circadian}.
\end{observation}

The mechanistic basis for circadian rhythm disruption extends beyond the HPA axis itself to involve the central circadian clock in the suprachiasmatic nucleus (SCN) of the hypothalamus. The NIH study documented abnormalities in temporal-parietal junction function and altered brain metabolism that may affect SCN regulation~\cite{walitt2024deep}. Additionally, inflammatory cytokines known to be elevated in ME/CFS (discussed in Chapter~\ref{ch:immune-dysfunction}) directly disrupt circadian clock gene expression, creating bidirectional interactions between immune activation and circadian dysregulation.

The clinical consequences of flattened cortisol rhythm are profound. The morning cortisol peak serves essential physiological functions: promoting waking and alertness, mobilizing glucose for energy availability, preparing the cardiovascular system for upright posture and activity, and modulating immune function to prevent excessive inflammation. Loss of this peak explains the characteristic morning symptom severity reported by many ME/CFS patients. These include difficulty waking, prolonged morning fatigue requiring hours to achieve minimal function, orthostatic intolerance upon standing (discussed in Chapter~\ref{ch:cardiovascular}), and cognitive dysfunction particularly severe in early morning hours.

\subsection{Mechanisms of HPA Dysfunction}

The dysregulation of the HPA axis in ME/CFS reflects multiple interconnected mechanisms operating at different levels of the neuroendocrine cascade. Understanding these mechanisms is essential for developing targeted therapeutic interventions and explaining why simple hormone replacement strategies have shown limited efficacy.

\subsubsection{Central Glucocorticoid Receptor Sensitivity}

\begin{hypothesis}[Enhanced Central Glucocorticoid Feedback]
\label{hyp:gcr-sensitivity}
The enhanced negative feedback sensitivity observed in ME/CFS may result from altered glucocorticoid receptor (GR) expression or function in hypothalamic and pituitary tissues. Several mechanisms could produce this effect. Upregulation of GR expression would increase sensitivity to existing cortisol levels. Altered GR isoform expression (GR$\alpha$ vs.\ GR$\beta$) might shift the balance toward enhanced feedback. Reduced expression of 11$\beta$-hydroxysteroid dehydrogenase type 1 (11$\beta$-HSD1), the enzyme that locally amplifies cortisol action by converting inactive cortisone to active cortisol, could diminish local glucocorticoid signaling. Finally, epigenetic modifications of the GR gene might affect transcription and receptor function.
\end{hypothesis}

This enhanced feedback creates a self-reinforcing cycle. Slightly elevated cortisol (or even normal-low cortisol) triggers disproportionate suppression of CRH and ACTH secretion, further reducing cortisol output. Under normal circumstances, this would reduce feedback inhibition and restore output, but the hypersensitive feedback prevents this compensatory response, maintaining chronically low HPA axis activity. This mechanism explains why ME/CFS patients do not develop frank adrenal insufficiency (baseline cortisol remains detectable) yet fail to mount appropriate stress responses (blunted reactivity to challenges). This self-reinforcing HPA dysfunction represents one of several vicious cycles in ME/CFS pathophysiology, as discussed in Section~\ref{sec:unifying-mechanisms} of Chapter~\ref{ch:integrative-models}.

\subsubsection{Inflammatory Cytokine Effects on HPA Axis}

The bidirectional relationship between the immune system and the HPA axis represents a critical mechanism in ME/CFS pathophysiology. Under normal circumstances, immune activation from infection or tissue damage stimulates HPA axis activity. Pro-inflammatory cytokines (IL-1, IL-6, TNF-$\alpha$) signal the hypothalamus to increase CRH secretion, resulting in elevated cortisol that dampens the immune response. This creates negative feedback that prevents excessive inflammation. The acute response adaptively contains immune activation while preventing immunopathology.

\begin{hypothesis}[Maladaptive Chronic Inflammatory Signaling]
\label{hyp:chronic-inflammation-hpa}
In ME/CFS, chronic low-grade inflammation (documented in Chapter~\ref{ch:immune-dysfunction}) may induce glucocorticoid resistance at immune cells while simultaneously increasing central negative feedback sensitivity. This paradoxical pattern produces the worst of both scenarios: insufficient cortisol secretion to control peripheral inflammation due to enhanced central feedback, yet reduced cortisol effectiveness at immune cells due to receptor downregulation or dysfunction~\cite{walitt2024deep,heng2025sexspecific}. The result is persistent inflammation despite apparent ``normal'' cortisol levels that would typically suppress such immune activation.
\end{hypothesis}

Recent evidence from Heng et al.\ (2025) documenting sex-specific immune dysregulation supports this model, showing that females with ME/CFS exhibit particularly pronounced pro-inflammatory profiles with elevated type 2 interferon signaling despite cortisol levels within the reference range~\cite{heng2025sexspecific}. This suggests functional glucocorticoid resistance at target tissues.

\subsubsection{Steroidogenic Enzyme Dysfunction}

The recent findings by Pipper et al.\ (2024) identifying elevated cortisol precursors (11-deoxycortisol, 17$\alpha$-hydroxyprogesterone) in severe ME/CFS patients suggest impaired function of steroidogenic enzymes, particularly 11$\beta$-hydroxylase (CYP11B1) which catalyzes the final step converting 11-deoxycortisol to cortisol~\cite{pipper2024steroid}. This enzyme dysfunction could result from several factors. Mitochondrial impairment may play a role, as steroidogenesis occurs in mitochondria and requires adequate ATP supply (discussed in Chapter~\ref{ch:energy-metabolism}). Cytokine-mediated suppression of enzyme expression or activity represents another possibility. Micronutrient deficiencies affecting enzyme cofactors or oxidative stress damaging enzyme proteins may also contribute.

If confirmed, this mechanism suggests that the problem is not purely regulatory (hypothalamic-pituitary signaling) but also biosynthetic (adrenal enzymatic capacity). This has important therapeutic implications, as interventions targeting upstream signaling may prove ineffective if the limiting step is enzymatic conversion within the adrenal gland.

\subsection{Clinical Consequences}

The HPA axis abnormalities documented in ME/CFS produce wide-ranging clinical effects that contribute directly to the cardinal symptoms of the disease. Understanding these consequences illuminates why seemingly minor hormonal changes cause profound functional impairment.

\subsubsection{Stress Response Abnormalities and Post-Exertional Malaise}

\begin{observation}[Blunted Physiological Stress Responses]
\label{obs:stress-response}
ME/CFS patients demonstrate inadequate cortisol responses to physiological stressors. These include exercise (both acute bouts and prolonged exertion), orthostatic challenge (standing, tilt-table testing), cognitive tasks requiring sustained mental effort, and psychological stressors. This blunted response means that stressors that healthy individuals accommodate with transient cortisol elevation produce inadequate counter-regulatory responses in ME/CFS patients, potentially explaining the delayed and prolonged symptom exacerbation characteristic of post-exertional malaise (PEM).
\end{observation}

The temporal pattern of PEM---symptom onset typically 12--48 hours after exertion rather than immediately---aligns with the kinetics of cortisol's effects on immune function and cellular metabolism. During exertion, ME/CFS patients may rely on sympathetic nervous system activation (catecholamines) to maintain function despite inadequate cortisol support. Following exertion, the delayed cortisol response fails to adequately suppress the inflammatory cascade initiated by exertion-induced cellular stress and damage. This unchecked inflammation then drives the delayed symptom exacerbation of PEM.

\subsubsection{Immune System Effects and Chronic Inflammation}

Cortisol serves as the body's primary endogenous anti-inflammatory hormone. It suppresses pro-inflammatory cytokine production (IL-1$\beta$, IL-6, TNF-$\alpha$), inhibits T cell activation and proliferation, promotes a shift from Th1 (cellular immunity) to Th2 (humoral immunity) responses, and prevents autoimmune reactions by maintaining immune tolerance. The blunted cortisol output and flattened diurnal rhythm in ME/CFS remove this tonic immunosuppressive influence, permitting chronic low-grade inflammation to persist.

\begin{hypothesis}[Loss of Diurnal Immune Regulation]
\label{hyp:diurnal-immune}
The flattened cortisol rhythm may be particularly consequential for immune regulation. Immune cells express glucocorticoid receptors and show circadian variation in their responsiveness to cortisol. The normal morning cortisol peak serves to ``reset'' immune function daily, preventing inflammatory pathways from remaining chronically activated. Loss of this peak in ME/CFS may allow inflammatory signaling to persist across day-night cycles without the normal circadian suppression~\cite{cambras2018circadian}.
\end{hypothesis}

This mechanism connects to the findings in Chapter~\ref{ch:immune-dysfunction} documenting altered cytokine profiles, NK cell dysfunction, and B cell abnormalities in ME/CFS. The sex-specific immune profiles identified by Heng et al.\ (2025) showing more pronounced inflammatory signatures in females align with the sex-specific steroid hormone abnormalities documented by Pipper et al.\ (2024), suggesting coordinated sex-dependent endocrine-immune interactions~\cite{heng2025sexspecific,pipper2024steroid}.

\subsubsection{Energy Metabolism and Glucose Homeostasis}

Cortisol plays essential roles in energy metabolism. It stimulates hepatic gluconeogenesis to maintain blood glucose availability, promotes lipolysis (fat breakdown) to provide alternative fuel sources, modulates insulin sensitivity to optimize glucose utilization, and supports mitochondrial function and cellular energy production. HPA axis dysfunction directly impairs these metabolic processes.

The cerebral glucose hypometabolism documented by PET imaging studies (Tirelli et al., 1998; Siessmeier et al., 2003) may partly reflect inadequate cortisol support for glucose uptake and utilization~\cite{tirelli1998pet,siessmeier2003pet}. Additionally, the blunted morning cortisol peak fails to provide the metabolic ``boost'' needed to transition from fasting metabolism to active daytime metabolism, contributing to severe morning fatigue and the prolonged time required to achieve minimal function after waking.

These metabolic effects connect directly to the mitochondrial dysfunction and energy metabolism deficits discussed in Chapter~\ref{ch:energy-metabolism}, suggesting that HPA axis abnormalities and cellular metabolic dysfunction represent interconnected rather than independent pathophysiological processes.

\paragraph{Gut Barrier Repair and Low Cortisol in Severe Patients.}

Beyond the well-established role of cortisol in stress response and metabolism, emerging evidence suggests that HPA axis dysfunction in severe ME/CFS may impair intestinal barrier maintenance and repair capacity, potentially contributing to chronic gut permeability and systemic inflammation.

\subparagraph{Low Morning Cortisol in Severely Ill Patients.}
A comprehensive biomarker examination of severely ill (housebound/bedbound) ME/CFS patients~\cite{Komaroff2021severe} documented significantly reduced morning salivary cortisol compared to age- and sex-matched healthy controls: median 0.20 mcg/dL in severe ME/CFS vs.\ 0.45 mcg/dL in controls (p = 0.002). This 55\% reduction in morning cortisol suggests substantial HPA axis dysregulation in the severe patient population. While the mechanisms remain debated, the functional consequence is reduced cortisol availability during periods when barrier repair processes are most active.

\subparagraph{Cortisol's Role in Intestinal Barrier Function.}
Glucocorticoids, including cortisol, play multifaceted roles in maintaining epithelial barrier integrity: (1) upregulating tight junction proteins (claudin-1, occludin, ZO-1) through glucocorticoid receptor-mediated transcription; (2) suppressing barrier-disrupting cytokines (IL-1$\beta$, IL-6, TNF-$\alpha$) via NF-$\kappa$B inhibition; (3) promoting enterocyte survival and differentiation; and (4) maintaining circadian tight junction protein expression with peak synthesis during the morning cortisol surge. The relationship is dose-dependent: physiological cortisol levels are barrier-protective, while low cortisol impairs repair capacity. ME/CFS patients appear to have insufficient cortisol for normal barrier maintenance.

\subparagraph{Impaired Barrier Repair in Severe ME/CFS: Mechanistic Hypothesis.}
Integrating low cortisol with evidence of baseline gut permeability~\cite{GutPermeability2023}, severe patients likely exhibit: (1) daily micro-damage from minimal activities (cognitive work, postural changes, meals) triggering transient splanchnic hypoperfusion; (2) insufficient nocturnal cortisol for barrier repair; (3) accumulating baseline permeability; (4) the cytokine-barrier bidirectional cycle (cytokines $\rightarrow$ tight junction disruption $\rightarrow$ LPS $\rightarrow$ cytokine amplification); and (5) nutritional deficits (low albumin in severe patients) limiting epithelial regeneration substrate. This model suggests wheat elimination response may be slower in severe patients due to impaired barrier repair capacity, but nutritional support (protein, micronutrients) and optimization of morning cortisol timing may accelerate recovery.

\section{Thyroid Function}
\label{sec:thyroid}

Thyroid dysfunction represents a critical consideration in ME/CFS for two distinct reasons: the substantial clinical overlap between hypothyroidism and ME/CFS symptoms creates diagnostic challenges requiring careful differentiation, and ME/CFS patients exhibit a specific pattern of thyroid abnormalities---the ``Low T3 Syndrome''---that occurs despite normal TSH levels and complicates interpretation of standard thyroid function tests. Understanding these thyroid-related issues is essential for appropriate diagnosis and management.

\subsection{The Low T3 Syndrome in ME/CFS}

\begin{achievement}[Low T3 Syndrome as Distinct ME/CFS Feature]
\label{ach:low-t3}
Ruiz-Núñez et al.\ (2018) conducted a rigorous case-control study comparing 98 ME/CFS patients to 99 healthy controls and documented a distinctive pattern of thyroid hormone abnormalities~\cite{ruiznunez2018thyroid}. ME/CFS patients showed 16\% prevalence of free T3 (FT3) below the reference range compared to only 7\% in controls (odds ratio 2.56). They exhibited significantly lower FT3, total T4 (TT4), and total T3 (TT3) concentrations, along with reduced T3/T4 ratio indicating impaired peripheral conversion of T4 to active T3. The percentage of reverse T3 (rT3), an inactive T3 isomer, was elevated, with increased rT3/TT3 ratio reflecting preferential conversion to the inactive form. Estimated deiodinase activity---the enzyme responsible for converting T4 to T3---showed a 14.4\% reduction.

Critically, these abnormalities occurred while thyroid-stimulating hormone (TSH) levels remained within the normal reference range. Standard thyroid screening tests would therefore classify these patients as ``euthyroid'' (normal thyroid function) despite functionally significant thyroid hormone deficits.
\end{achievement}

This pattern resembles the ``non-thyroidal illness syndrome'' (NTIS) or ``euthyroid sick syndrome'' observed in acute critical illness, starvation, and chronic diseases. However, unlike the transient thyroid changes in acute illness that normalize with recovery, the Low T3 Syndrome in ME/CFS persists chronically and may represent a maladaptive response that perpetuates rather than resolves the disease state.

\subsection{Mechanisms of Impaired T4 to T3 Conversion}

The conversion of thyroxine (T4, the major thyroid hormone secreted by the thyroid gland) to triiodothyronine (T3, the metabolically active form) occurs primarily in peripheral tissues through the action of deiodinase enzymes. Three deiodinase isoforms exist. Type 1 deiodinase (D1) in liver and kidney produces most circulating T3. Type 2 deiodinase (D2) in brain, pituitary, and brown fat produces local T3 for tissue-specific needs. Type 3 deiodinase (D3) in multiple tissues inactivates T4 and T3 by converting them to reverse T3 (rT3) and T2.

\begin{hypothesis}[Cytokine-Mediated Deiodinase Suppression]
\label{hyp:deiodinase-suppression}
The chronic low-grade inflammation documented in ME/CFS (Chapter~\ref{ch:immune-dysfunction}) likely suppresses deiodinase enzyme activity through multiple mechanisms. Pro-inflammatory cytokines, particularly IL-6 and TNF-$\alpha$, directly inhibit D1 and D2 expression and activity while upregulating D3, shifting the balance toward production of inactive reverse T3 rather than active T3. Oxidative stress, elevated in ME/CFS (Chapter~\ref{ch:energy-metabolism}), damages selenocysteine residues essential for deiodinase enzymatic function; all deiodinases are selenium-dependent enzymes. Additionally, the mitochondrial dysfunction documented in ME/CFS may impair ATP-dependent cellular uptake of T4, reducing substrate availability for conversion to T3~\cite{ruiznunez2018thyroid}.
\end{hypothesis}

This mechanism explains why simply increasing thyroid hormone replacement dose (giving more T4) often fails to improve symptoms in ME/CFS patients: the limiting factor is not T4 availability but rather the capacity to convert T4 to active T3 at the cellular level.

\subsection{Tissue-Level Thyroid Hormone Resistance}

Beyond impaired T4 to T3 conversion, emerging evidence suggests that some ME/CFS patients may exhibit functional thyroid hormone resistance at the cellular level. This could involve reduced expression or function of thyroid hormone transporters (MCT8, MCT10) that move hormones into cells, altered expression of thyroid hormone receptors (TR$\alpha$, TR$\beta$) in target tissues, or impaired receptor-coactivator interactions that reduce transcriptional responses to thyroid hormone binding.

\begin{observation}[Selenium Autoantibodies and Acquired Resistance]
\label{obs:selenium-antibodies}
A Netherlands study identified markedly elevated selenium autoantibodies in 9.6--15.6\% of ME/CFS patients compared to only 0.9--2.0\% of healthy controls. Selenium is essential for deiodinase function, selenoprotein synthesis, and thyroid hormone metabolism. Autoantibodies against selenium transport proteins could create an acquired form of thyroid hormone resistance by impairing the selenium-dependent enzymatic machinery required for thyroid hormone activation and action.
\end{observation}

This finding suggests a potential autoimmune mechanism contributing to thyroid dysfunction in a subset of ME/CFS patients and raises the possibility that interventions targeting selenium metabolism might benefit this subgroup.

\subsection{Clinical Implications and Diagnostic Challenges}

The overlap between ME/CFS symptoms and hypothyroidism creates substantial diagnostic challenges. Both conditions present with severe fatigue and exhaustion, cognitive impairment (``brain fog''), cold intolerance and temperature dysregulation, weight changes and metabolic disturbances, mood alterations including depression, muscle weakness and pain, and sleep disturbances. Undiagnosed hypothyroidism may therefore masquerade as ME/CFS, while ME/CFS-related Low T3 Syndrome may be mistaken for thyroid disease.

\begin{warning}[Limitations of Standard Thyroid Testing in ME/CFS]
\label{warn:thyroid-testing}
Standard thyroid screening using TSH alone is insufficient for evaluating thyroid function in ME/CFS patients. The Low T3 Syndrome occurs with normal TSH because the pituitary senses adequate T4 levels and reduces TSH secretion appropriately, unaware that peripheral tissues cannot effectively convert T4 to active T3. Comprehensive thyroid evaluation in ME/CFS should include TSH to exclude primary thyroid disease, free T4 (FT4) to assess thyroid hormone production, free T3 (FT3) to evaluate the active hormone level, and reverse T3 (rT3) to assess the balance between activation and inactivation. Calculation of T3/T4 and rT3/T3 ratios quantifies conversion efficiency~\cite{ruiznunez2018thyroid}.
\end{warning}

The therapeutic implications remain uncertain. While the rationale for T3 supplementation appears sound (directly providing the active hormone bypasses the impaired conversion step), clinical trial evidence remains limited and results have been mixed. Some patients report subjective improvement, others experience no benefit, and a subset develops adverse effects (palpitations, anxiety, insomnia) suggesting tissue-level hypersensitivity to thyroid hormone. Careful individualized treatment trials with close monitoring may be warranted in ME/CFS patients with documented Low T3 Syndrome, but systematic evidence for efficacy is lacking.

\section{Sex Hormones and Gender Differences}
\label{sec:sex-hormones}

Sex hormones and gender differences represent one of the most striking yet inadequately explained features of ME/CFS epidemiology and pathophysiology. The consistent 3--4:1 female-to-male prevalence ratio, menstrual cycle exacerbations reported by female patients, early menopause associations, and sex-specific patterns of immune dysfunction and steroid hormone abnormalities collectively indicate that reproductive hormones play central roles in disease susceptibility, expression, and progression.

\subsection{Epidemiology of Sex Differences}

\begin{achievement}[Female Predominance in ME/CFS]
\label{ach:female-prevalence}
A systematic review and meta-analysis by Lim et al.\ (2020) synthesized prevalence data from 45 articles representing 46 studies and 56 prevalence datasets spanning 1980--2018~\cite{lim2020prevalence}. The analysis documented overall ME/CFS prevalence of 0.89\% using CDC-1994 criteria, with female prevalence approximately 1.5--2.0 fold higher than males across all studies. In the total population, prevalence was 2.24\% ± 2.59\% in females versus 1.11\% ± 1.05\% in males. General population studies showed 2.83\% versus 1.39\%, and meta-analysis pooled estimates indicated 1.36\% versus 0.89\%. The female-to-male ratio consistently ranges from 3:1 to 4:1 across geographic regions, diagnostic criteria, and study methodologies.
\end{achievement}

More recent estimates using large-scale medical claims data and machine learning by Jason et al.\ (2018) confirmed the persistent female predominance, though highlighting that 35--40\% of ME/CFS patients are male---a substantial population whose experiences may differ from the predominantly studied female cohorts~\cite{jason2018prevalence}. The consistency of the sex ratio across diverse populations and diagnostic approaches argues strongly for biological sex differences in disease susceptibility or expression rather than artifacts of health-seeking behavior or diagnostic bias.

\subsection{Sex-Specific Pathophysiology: NIH Deep Phenotyping Findings}

\begin{achievement}[Distinct Male and Female Pathophysiological Patterns]
\label{ach:sex-specific-pathophys}
The 2024 NIH deep phenotyping study by Walitt et al.\ employed multi-omics analysis to directly compare male and female ME/CFS patients~\cite{walitt2024deep}. This rigorous investigation revealed fundamentally different pathophysiological signatures. Males showed altered T cell activation patterns and abnormal innate immunity markers, while females demonstrated abnormal B cell function and altered white blood cell growth patterns. The sexes exhibited distinct inflammatory marker profiles, divergent gene expression patterns in immune cells, different immune cell population distributions, and sex-specific metabolic marker abnormalities.

This finding challenges the implicit assumption in ME/CFS research that male and female patients share a common pathophysiology differing only in prevalence. Instead, it suggests that ME/CFS may represent partially distinct disease processes in males and females requiring sex-stratified approaches to diagnosis, biomarker development, and treatment.
\end{achievement}

Complementing the NIH findings, Heng et al.\ (2025) documented sex-specific immune dysregulation in long COVID patients with ME/CFS~\cite{heng2025sexspecific}. Females exhibited decreased lymphocyte counts with increased neutrophils and monocytes (a myelopoiesis shift), elevated pro-inflammatory cytokines, and upregulated type 2 interferon signaling (IP-10, IFN-$\gamma$). Males showed fewer inflammatory alterations overall, with more balanced profiles, elevated anti-inflammatory IL-10, and IL-1 signaling dominance rather than interferon predominance. These immune differences likely reflect underlying hormonal influences on immune cell development, activation, and cytokine production.

\subsection{Steroid Hormone Abnormalities}

\begin{achievement}[Sex and Severity-Stratified Steroid Hormone Profiles]
\label{ach:steroid-profiles}
Pipper et al.\ (2024) conducted the first comprehensive sex-stratified analysis of steroid hormones in ME/CFS using high-precision UHPLC-MS/MS (ultra-high performance liquid chromatography tandem mass spectrometry)~\cite{pipper2024steroid}. This study of 97 total participants revealed striking sex-specific and severity-dependent patterns. Female patients with severe ME/CFS demonstrated elevated 11-deoxycortisol (a cortisol precursor) and 17$\alpha$-hydroxyprogesterone, suggesting impaired final enzymatic steps in cortisol biosynthesis. Females with mild-to-moderate disease showed increased progesterone levels. Male patients with mild-to-moderate ME/CFS exhibited frankly reduced cortisol and corticosterone but paradoxically elevated progesterone---an unexpected finding suggesting complex dysregulation of steroidogenic pathways.

The machine learning classifier achieved 71.2\% accuracy for discriminating female ME/CFS patients from controls and 84.6\% accuracy for males based solely on steroid hormone profiles, supporting the potential development of sex-specific hormonal biomarkers for diagnosis and disease monitoring.
\end{achievement}

These findings indicate that sex hormone abnormalities in ME/CFS extend beyond simple deficiency or excess of individual hormones to involve coordinated dysregulation of steroidogenic pathways, enzyme activities, and metabolic ratios. The sex-specific patterns suggest that estrogen and progesterone in females versus testosterone and its metabolites in males exert distinct effects on disease expression.

\subsection{Reproductive Health and Gynecological Risk Factors}

\subsubsection{Menstrual Cycle Effects and Hormonal Fluctuations}

The majority of premenopausal women with ME/CFS report significant symptom exacerbations related to menstrual cycle phases, particularly during the premenstrual week (late luteal phase when progesterone declines rapidly) and during menstruation itself. Common cyclical exacerbations include increased fatigue and post-exertional malaise severity, worsened cognitive impairment and brain fog, intensified pain and sensory sensitivity, heightened orthostatic intolerance symptoms, and mood disturbances such as irritability, anxiety, and depression.

\begin{observation}[Hormone-Symptom Correlations Across Menstrual Cycle]
\label{obs:menstrual-symptoms}
Preliminary findings from chronobiology-based studies mapping hormonal fluctuations across the menstrual cycle in ME/CFS reveal systematic symptom-hormone relationships. Fatigue and pain peak premenstrually when estradiol and progesterone levels fall. Cognitive impairment shows lowest severity at ovulation when estradiol peaks. Low estradiol and progesterone concentrations correlate with higher fatigue and pain ratings. Additionally, luteinizing hormone (LH) and follicle-stimulating hormone (FSH) levels positively correlate with fatigue severity and orthostatic symptoms, suggesting pituitary dysregulation may contribute to symptom variability.
\end{observation}

These patterns suggest that the absolute levels of sex hormones may be less important than the dynamic fluctuations and ratios between estrogen, progesterone, and pituitary gonadotropins. The rapid hormonal changes during certain cycle phases may destabilize physiological systems already compromised by ME/CFS, triggering symptom exacerbations.

\subsubsection{Gynecological Comorbidities and Early Menopause}

\begin{achievement}[Gynecological Risk Factors for ME/CFS]
\label{ach:gynecological-risk}
Population-based case-control studies by Boneva et al.\ have identified specific gynecological risk factors associated with ME/CFS development~\cite{boneva2011gynecological,boneva2015menopause}. The 2011 study documented that ME/CFS cases reported significantly higher rates of pelvic pain unrelated to menstruation (22.2\% versus 1.7\% in controls), endometriosis diagnosis (36.1\% versus 16.7\%), prolonged periods of amenorrhea (absence of menstruation), and history of gynecological surgery including hysterectomy and oophorectomy. Premenopausal women with ME/CFS tended toward lower luteal phase progesterone with higher FSH.

Most strikingly, the 2015 study revealed that women with ME/CFS experienced menopause approximately 11 years earlier than controls (mean age 37.6 years versus 48.6 years)~\cite{boneva2015menopause}. Early menopause (<40 years) represents premature ovarian insufficiency, indicating fundamental dysfunction of the hypothalamic-pituitary-gonadal (HPG) axis. This finding suggests that HPG dysfunction may precede or contribute to ME/CFS development rather than merely resulting from the disease.
\end{achievement}

The mechanisms connecting gynecological abnormalities to ME/CFS remain incompletely understood but likely involve bidirectional interactions. Chronic inflammation (documented in Chapter~\ref{ch:immune-dysfunction}) affects ovarian function, endometrial health, and hormonal regulation. Conversely, sex hormone abnormalities modulate immune function, with estrogen generally enhancing immune responses (potentially contributing to female predominance of autoimmune diseases) and progesterone providing immunosuppressive and anti-inflammatory effects. Disruption of these hormonal immunomodulatory signals could perpetuate the immune dysfunction characteristic of ME/CFS.

\subsection{Testosterone and Androgens}

While research has focused predominantly on female ME/CFS patients due to the higher prevalence, emerging evidence indicates that male patients exhibit distinct hormonal abnormalities, particularly involving androgens (testosterone, DHEA, and their metabolites). A pilot study of 23 women ages 35--55 with ME/CFS found that 89\% had suboptimal DHEA-S (dehydroepiandrosterone sulfate) levels. Supplementation with DHEA led to statistically significant improvements: 18\% reduction in pain, 21\% reduction in fatigue, 35\% reduction in anxiety, 26\% improvement in thinking ability, 17\% improvement in memory, and 22\% improvement in sexual function.

\begin{hypothesis}[Androgen Deficiency Contribution to Symptoms]
\label{hyp:androgen-deficiency}
Androgens, particularly DHEA and testosterone, serve multiple physiological functions beyond reproductive roles. They support mitochondrial function and cellular energy production, promote muscle mass maintenance and physical strength, modulate immune responses with generally anti-inflammatory effects, influence mood, motivation, and cognitive function, and affect pain perception and nociceptive processing. Deficiency of these hormones could mechanistically contribute to core ME/CFS symptoms including fatigue, cognitive impairment, reduced physical capacity, and pain amplification.
\end{hypothesis}

The sex-specific steroid profiles identified by Pipper et al.\ showing reduced cortisol but elevated progesterone in male ME/CFS patients suggest complex interactions between the HPA axis and gonadal steroid production~\cite{pipper2024steroid}. Further research specifically examining male ME/CFS patients is critically needed to elucidate androgen metabolism and its therapeutic potential.

\subsection{Mechanisms of Sex Hormone Influence on ME/CFS}

Sex hormones exert pervasive effects on immune function, energy metabolism, neurotransmitter systems, and autonomic regulation---all domains disrupted in ME/CFS. Understanding these mechanisms illuminates how hormonal dysregulation contributes to pathophysiology.

\subsubsection{Immune Modulation by Sex Hormones}

Estrogen generally enhances immune responses. It promotes B cell maturation and antibody production, enhances T helper 2 (Th2) responses, increases pro-inflammatory cytokine production in certain contexts, and potentially contributes to higher autoimmune disease prevalence in females. Progesterone exerts immunosuppressive effects. It promotes T regulatory cell (Treg) function essential for immune tolerance, suppresses Th1 inflammatory responses, reduces pro-inflammatory cytokine production, and normally balances estrogen's immune-enhancing effects during the menstrual cycle.

Testosterone generally suppresses immune activation. It reduces B cell activity and antibody production, suppresses pro-inflammatory cytokine secretion, and potentially explains lower autoimmune disease rates in males. The loss of normal hormonal modulation of immune function---whether through estrogen-progesterone imbalance in females, DHEA/testosterone deficiency in both sexes, or altered receptor sensitivity---could permit the chronic immune activation documented in Chapter~\ref{ch:immune-dysfunction} to persist unchecked.

\subsubsection{Neuroendocrine Integration}

The sex-specific pathophysiology documented by Walitt et al.\ and Heng et al.\ likely reflects coordinated neuroendocrine-immune interactions rather than isolated hormonal effects~\cite{walitt2024deep,heng2025sexspecific}. The hypothalamus and pituitary integrate signals from immune cytokines, metabolic hormones, and gonadal steroids to coordinate systemic responses. In ME/CFS, this integration appears fundamentally disrupted, with females showing stronger interferon signatures (potentially reflecting estrogen's enhancement of interferon responses) and males showing more balanced but still abnormal patterns (potentially reflecting testosterone's dampening effects on certain immune pathways).

The early menopause and gynecological abnormalities documented by Boneva et al.\ suggest that the HPG axis dysfunction may represent a form of neuroendocrine exhaustion analogous to the HPA axis hypofunction discussed earlier in this chapter~\cite{boneva2015menopause}. Chronic immune activation and cytokine exposure may dysregulate both axes, creating a state of multi-system endocrine insufficiency despite the absence of structural gland failure.

\section{Growth Hormone and IGF-1}
\label{sec:growth-hormone}

Growth hormone (GH) and its primary mediator, insulin-like growth factor 1 (IGF-1), coordinate critical aspects of metabolism, body composition, and cellular function that directly relate to ME/CFS symptoms. The growth hormone axis regulates protein synthesis and muscle mass maintenance, lipolysis (fat breakdown) and glucose metabolism, bone density and connective tissue integrity, immune function and wound healing, and cognitive function and mood. GH/IGF-1 deficiency could therefore contribute to the muscle weakness, metabolic dysfunction, and cognitive impairment characteristic of ME/CFS.

\subsection{Evidence for GH/IGF-1 Axis Dysfunction}

Research on the GH/IGF-1 axis in ME/CFS has produced conflicting findings, likely reflecting the heterogeneity of patient populations and the complexity of growth hormone regulation. Some studies document clear abnormalities, while others find normal GH dynamics, suggesting that GH dysfunction characterizes a subset of ME/CFS patients rather than representing a universal feature.

\begin{observation}[Reduced IGF-1 in ME/CFS Subset]
\label{obs:igf1-low}
Bennett et al.\ (1997) documented significantly lower serum IGF-1 levels in ME/CFS patients compared to healthy controls, accompanied by reduced nocturnal secretion of growth hormone. IGF-1 serves as the primary mediator of GH's anabolic effects, produced predominantly in the liver in response to GH stimulation and acting on peripheral tissues to promote protein synthesis, muscle growth, and metabolic regulation. Low IGF-1 despite normal or near-normal GH secretion suggests either hepatic resistance to GH or impaired liver function affecting IGF-1 synthesis.
\end{observation}

However, contradicting these findings, other rigorous studies found no differences in basal IGF-1 or IGF-binding protein (IGFBP) levels between ME/CFS patients and controls, normal urinary growth hormone excretion, and similar GH responses to provocative testing. These inconsistencies highlight the challenge of identifying reliable biomarkers in a heterogeneous disease and suggest the need for subgroup stratification based on clinical phenotypes or other biomarkers.

\subsection{Growth Hormone Treatment Trial}

\begin{achievement}[Physiological but Limited Clinical Benefits of GH Treatment]
\label{ach:gh-treatment}
Moorkens et al.\ (2000) conducted a randomized, double-blind, placebo-controlled trial of growth hormone treatment in ME/CFS patients with documented low IGF-1 levels~\cite{moorkens2000gh}. The study involved 20 patients receiving 12 weeks of active treatment followed by a 9-month open-label phase. Results demonstrated clear physiological effects: mean serum IGF-1 increased from 173 $\pm$ 46 $\mu$g/L to 296 $\pm$ 89 $\mu$g/L (p$<$0.001), fat-free mass significantly increased, and total body water increased, confirming the anabolic effects of GH. However, clinical outcomes were mixed: quality of life measures did not show significant improvement overall, though notably, 4 patients resumed work after prolonged illness, suggesting substantial benefit in a subset.
\end{achievement}

The dissociation between clear physiological effects (increased IGF-1, improved body composition) and limited symptom improvement suggests that while GH deficiency may contribute to certain ME/CFS features (particularly muscle weakness and poor exercise tolerance), it is not the primary driver of fatigue, post-exertional malaise, or cognitive symptoms. This pattern aligns with the multi-system nature of ME/CFS pathophysiology, where correcting a single hormonal deficit proves insufficient to restore overall function.

\subsection{Mechanisms and Clinical Implications}

Several mechanisms could explain GH/IGF-1 axis dysfunction in ME/CFS, each with distinct therapeutic implications:

\begin{hypothesis}[Mechanisms of GH/IGF-1 Dysfunction]
\label{hyp:gh-mechanisms}
Multiple non-exclusive mechanisms may contribute to growth hormone axis dysfunction. Hypothalamic dysfunction may reduce GH-releasing hormone (GHRH) secretion, paralleling the HPA axis hypofunction discussed earlier. The NIH study documented temporal-parietal junction and broader brain abnormalities that could affect hypothalamic regulation~\cite{walitt2024deep}. Hepatic resistance to GH action may impair IGF-1 synthesis despite adequate GH secretion, potentially reflecting the mitochondrial dysfunction and oxidative stress documented in Chapter~\ref{ch:energy-metabolism}. Cytokine-mediated suppression may inhibit the GH axis, as chronic inflammation suppresses both GH secretion and IGF-1 synthesis while inducing IGF-1 resistance at target tissues. Finally, sleep disruption may reduce nocturnal GH pulses; the majority of daily GH secretion occurs during deep sleep, particularly slow-wave sleep, which is disrupted in ME/CFS.
\end{hypothesis}

The clinical implications of these findings remain uncertain. GH treatment showed physiological effects but limited symptom benefit in the controlled trial, suggesting it may help selected patients but is not a universal solution. The expense, need for daily injections, and potential adverse effects (fluid retention, carpal tunnel syndrome, glucose intolerance) argue for restricting GH treatment to patients with documented IGF-1 deficiency and careful monitoring of both physiological parameters and functional outcomes. Alternative approaches targeting upstream causes (sleep improvement, inflammation reduction, mitochondrial support) might prove more effective than hormone replacement alone.

\section{Insulin and Glucose Metabolism}
\label{sec:insulin-glucose}

Glucose metabolism abnormalities in ME/CFS connect endocrine dysfunction directly to the cellular energy deficits discussed in Chapter~\ref{ch:energy-metabolism}. Insulin regulates glucose uptake into cells, coordinates switching between glucose and fat oxidation, affects mitochondrial function and ATP production, and modulates inflammatory responses and immune function. Dysregulation of insulin signaling and glucose metabolism therefore has cascading effects on multiple systems compromised in ME/CFS.

\subsection{Metabolic Syndrome and Insulin Resistance}

\begin{achievement}[ME/CFS Association With Metabolic Syndrome]
\label{ach:metabolic-syndrome}
A population-based case-control study by Maloney et al.\ (2010) examined the relationship between ME/CFS and metabolic syndrome in Georgia~\cite{maloney2010metabolic}. The analysis revealed that ME/CFS patients were approximately 2-fold more likely to have metabolic syndrome compared to controls (odds ratio 2.12, 95\% confidence interval 1.06--4.23). The key discriminating factors were increased waist circumference, elevated triglycerides, and higher fasting glucose. Notably, each additional metabolic syndrome component present associated with a 37\% increase in the likelihood of having ME/CFS, demonstrating a dose-response relationship.
\end{achievement}

Metabolic syndrome comprises a cluster of abnormalities: abdominal obesity (increased waist circumference), elevated triglycerides, reduced HDL cholesterol, elevated blood pressure, and elevated fasting glucose or insulin resistance. The syndrome reflects underlying insulin resistance (reduced cellular responsiveness to insulin signaling) and predicts increased risk for type 2 diabetes, cardiovascular disease, and inflammatory conditions.

The association between ME/CFS and metabolic syndrome raises important mechanistic questions about causality. Does insulin resistance contribute to ME/CFS pathophysiology, or does ME/CFS-related inflammation, inactivity, and mitochondrial dysfunction lead to insulin resistance? Likely, bidirectional relationships exist, with each condition exacerbating the other in a self-reinforcing cycle.

\subsection{Metabolic Phenotypes and Insulin Dynamics}

Armstrong et al.\ (2021) employed comprehensive metabolomics to identify distinct ME/CFS subtypes with different metabolic signatures. One subtype (ME-M2) demonstrated elevated triglyceride and insulin levels despite normal glucose, reflecting low-grade lipid-induced insulin resistance. ME/CFS patients overall showed slightly elevated insulin and leptin (an adipose tissue hormone signaling energy status) and lower high molecular weight adiponectin (an anti-inflammatory adipokine that enhances insulin sensitivity).

\begin{hypothesis}[Peripheral Insulin Resistance With Central Deficits]
\label{hyp:insulin-resistance}
ME/CFS may involve a paradoxical state of peripheral insulin resistance (reduced glucose uptake in muscle and adipose tissue) combined with inadequate glucose delivery or utilization in the central nervous system. This would explain the constellation of elevated peripheral insulin levels reflecting compensatory hyperinsulinemia, cerebral glucose hypometabolism documented by PET imaging, symptoms resembling hypoglycemia without documented low blood glucose, and the energy deficit despite apparently adequate systemic glucose availability~\cite{tirelli1998pet,siessmeier2003pet}.
\end{hypothesis}

\subsection{Cerebral Glucose Hypometabolism}

\begin{achievement}[Objective Evidence of Brain Energy Deficit]
\label{ach:brain-hypometabolism}
Positron emission tomography (PET) studies using fluorodeoxyglucose (FDG) tracer have documented reduced cerebral glucose metabolism in ME/CFS patients. Tirelli et al.\ (1998) identified significant glucose hypometabolism in the right mediofrontal cortex (p=0.010) and brainstem (p=0.013) in ME/CFS patients compared to controls, with moderate hypometabolism in the pons~\cite{tirelli1998pet}. Siessmeier et al.\ (2003) employed observer-independent analysis in 26 ME/CFS patients and found that 12/26 (46\%) showed hypometabolism bilaterally in the cingulate gyrus and adjacent mesial cortical areas, with 5 also demonstrating decreased orbitofrontal metabolism~\cite{siessmeier2003pet}. Importantly, hypometabolism correlated with anxiety and depression measures but not directly with fatigue severity, suggesting complex relationships between metabolic deficits and symptom expression.
\end{achievement}

The consistent documentation of reduced cerebral glucose uptake across independent studies using rigorous quantitative methods provides objective evidence of brain energy deficit in ME/CFS. The affected regions---brainstem nuclei, cingulate cortex, prefrontal areas---overlap substantially with regions showing functional abnormalities in the NIH study and with brain networks subserving autonomic control, attention, emotion regulation, and effort-based decision-making~\cite{walitt2024deep}.

Several mechanisms could produce cerebral hypometabolism beyond simple insulin resistance:

\begin{hypothesis}[Mechanisms of Cerebral Hypometabolism]
\label{hyp:brain-hypometabolism}
Multiple factors likely contribute to reduced brain glucose utilization. Cerebral hypoperfusion documented by SPECT imaging reduces glucose delivery to neurons. Glucose transporter dysfunction (GLUT1 at blood-brain barrier, GLUT3 in neurons) may impair glucose uptake even when delivery is adequate. Mitochondrial dysfunction in neurons reduces the capacity to metabolize glucose to ATP, causing glucose accumulation rather than utilization. Neuroinflammation with activated microglia alters brain energetics, as activated immune cells in the brain preferentially utilize glucose via glycolysis rather than oxidative phosphorylation. Finally, reduced neuronal activity secondary to other ME/CFS-related dysfunction may reduce metabolic demand, causing secondary hypometabolism as a consequence rather than cause of neurological symptoms.
\end{hypothesis}

\subsection{Hypoglycemia Symptoms Versus Orthostatic Intolerance}

\begin{warning}[Misattribution of Orthostatic Symptoms to Hypoglycemia]
\label{warn:hypoglycemia-misattribution}
Many ME/CFS patients report symptoms they attribute to ``hypoglycemia'': nausea, lightheadedness and faintness, sweating and tremor, weakness and malaise, and cognitive impairment. However, studies examining ME/CFS patients during symptomatic episodes have shown that these symptoms frequently occur without documented hypoglycemia (blood glucose <70 mg/dL). Instead, the symptoms typically reflect orthostatic intolerance (inadequate blood pressure and cerebral perfusion upon standing or with prolonged upright posture, discussed extensively in Chapter~\ref{ch:cardiovascular}). Critically, when orthostatic intolerance is treated effectively, the ``hypoglycemia'' symptoms often improve despite no specific glucose intervention.
\end{warning}

This misattribution has important clinical implications. ME/CFS patients may consume frequent snacks or high-carbohydrate meals attempting to prevent ``hypoglycemia,'' potentially exacerbating insulin resistance and weight gain. Additionally, focusing on blood sugar management diverts attention from the actual orthostatic problem requiring different interventions (increased salt and fluid intake, compression garments, medications affecting blood pressure or blood volume).

\subsection{Integration With Energy Metabolism Dysfunction}

The glucose metabolism abnormalities documented in this section connect directly to the mitochondrial dysfunction and cellular energy deficits discussed in Chapter~\ref{ch:energy-metabolism}. Insulin resistance reduces glucose uptake into cells, limiting substrate availability for ATP production. Impaired mitochondrial function reduces the capacity to oxidize glucose efficiently, causing metabolic bottlenecks. The resulting cellular energy deficit triggers compensatory responses including increased reliance on glycolysis (less efficient ATP production), activation of AMP-activated protein kinase (AMPK) signaling cellular energy stress, and metabolic shifts toward fat oxidation when glucose utilization fails.

These metabolic derangements help explain post-exertional malaise, as exertion depletes limited ATP stores that cannot be rapidly replenished due to impaired glucose metabolism and mitochondrial dysfunction. The delayed recovery characteristic of PEM reflects the slow restoration of cellular energy status when metabolic pathways remain compromised.

\section{Melatonin and Circadian Rhythms}
\label{sec:circadian}

Circadian rhythm disruption represents a pervasive feature of ME/CFS that intersects with nearly every other aspect of pathophysiology discussed in this chapter. The circadian system coordinates temporal organization of physiological processes including the HPA axis cortisol rhythm, immune function cycling between pro- and anti-inflammatory states, metabolic switching between anabolic and catabolic metabolism, body temperature regulation and thermoregulation, and sleep-wake cycles and alertness patterns. Disruption of circadian timing thus has cascading effects across multiple systems already compromised in ME/CFS.

\subsection{Objective Documentation of Circadian Disruption}

\begin{observation}[Actigraphy-Documented Circadian Abnormalities]
\label{obs:circadian-disruption}
Cambras et al.\ (2018) employed rigorous actigraphy monitoring to objectively document circadian rhythm abnormalities in ME/CFS patients~\cite{cambras2018circadian}. This case-control study of 10 women with ME/CFS and 10 matched controls revealed that daily activity levels were significantly lower in ME/CFS patients, relative amplitude of the activity rhythm (difference between peak and nadir) was reduced, indicating flattened circadian variation, and stability of the activity rhythm across days was decreased, showing less consistent day-to-day patterns. Additionally, distal skin temperature showed lower nocturnal values in winter, suggesting impaired circadian regulation of peripheral blood flow and thermoregulation.
\end{observation}

These findings demonstrate that circadian disruption in ME/CFS is not merely subjective patient reports of ``feeling tired at the wrong times'' but rather represents measurable alterations in the fundamental 24-hour organization of physiological functions. The reduced amplitude of activity rhythms parallels the flattened cortisol rhythm discussed earlier, suggesting a coordinated loss of circadian regulation across multiple output systems.

\subsection{Sleep Architecture Versus Circadian Timing}

An important conceptual distinction must be maintained between sleep architecture abnormalities (changes in sleep stage distribution, fragmentation, sleep efficiency) and circadian timing disruption (shifts in the phase or amplitude of 24-hour rhythms). ME/CFS patients exhibit both types of abnormalities, but they reflect different underlying mechanisms and require different therapeutic approaches.

Sleep architecture studies consistently document that ME/CFS patients experience longer sleep latency (time to fall asleep), more frequent awakenings during the night, more time in bed relative to total sleep time (reduced sleep efficiency), later and more variable wake times, irregular sleep patterns across days, and the paradox of unrefreshing sleep despite adequate or even prolonged total sleep duration. Children with ME/CFS often show continuous sleep exceeding 10 hours yet wake unrefreshed, indicating profound sleep dysfunction.

The circadian timing component involves altered phase relationships between sleep-wake cycles and other circadian outputs (body temperature, hormone secretion, immune function). Some ME/CFS patients show delayed sleep phase (natural sleep-wake times shifted later, resembling ``night owl'' patterns), while others exhibit irregular rhythms without clear 24-hour periodicity, and some maintain normal phase relationships but with reduced amplitude of rhythms.

\subsection{Molecular Clock Dysfunction}

\begin{hypothesis}[Clock Gene Dysregulation in ME/CFS]
\label{hyp:clock-genes}
Emerging evidence suggests disruption at the molecular level of circadian clock gene expression. Genome-wide association studies (GWAS) have reported nominally significant associations with NPAS2 (neuronal PAS domain protein 2), a core clock gene. Transcriptomic analysis of ME/CFS patient samples showed 10-fold higher NPAS2 expression compared to controls and elevated expression of other circadian rhythm genes in peripheral blood mononuclear cells (PBMCs). Enrichment of CLOCK gene variants in ME/CFS patients with comorbid fibromyalgia and epigenetic changes in ``circadian entrainment'' pathways suggest heritable and acquired alterations in clock gene function.
\end{hypothesis}

The molecular clock operates as a transcriptional-translational feedback loop involving core clock genes (CLOCK, BMAL1, PER1/2/3, CRY1/2) that regulate their own expression with approximately 24-hour periodicity. These clock genes also control thousands of downstream genes involved in metabolism, immune function, and cellular processes, creating temporal coordination across organ systems. Disruption of clock gene function could therefore produce pleiotropic effects consistent with the multi-system nature of ME/CFS.

Inflammatory cytokines, particularly IL-1$\beta$ and TNF-$\alpha$ elevated in ME/CFS (Chapter~\ref{ch:immune-dysfunction}), directly disrupt clock gene expression and alter circadian rhythms. This creates bidirectional interactions where immune dysfunction disturbs circadian regulation, while circadian disruption impairs proper immune function, perpetuating a self-reinforcing cycle.

\subsection{Melatonin and Its Therapeutic Potential}

Melatonin serves as both a marker and mediator of circadian rhythms, secreted by the pineal gland predominantly at night in response to darkness signals from the suprachiasmatic nucleus (SCN). Melatonin synchronizes peripheral clocks throughout the body, exerts direct antioxidant and anti-inflammatory effects, modulates immune function and cytokine production, and facilitates sleep initiation though it is not primarily a sedative.

Limited evidence suggests altered melatonin production in ME/CFS, though findings have been inconsistent, likely reflecting the heterogeneity of circadian dysfunction patterns. Some patients show reduced melatonin amplitude, others exhibit phase shifts (melatonin rise at inappropriate times), while some maintain apparently normal melatonin profiles despite subjective circadian symptoms.

\begin{achievement}[Melatonin Treatment Benefits]
\label{ach:melatonin-treatment}
Castro-Marrero et al.\ (2021) conducted a 16-week randomized, double-blind, placebo-controlled trial of melatonin (1 mg) plus zinc (10 mg) daily in 50 ME/CFS patients~\cite{castromarrero2021melatonin}. The intervention significantly reduced physical fatigue perception (p<0.05) and improved the physical component summary score compared to placebo. Urinary 6-sulfatoxymelatonin (the primary melatonin metabolite) increased significantly in the treatment group (p<0.0001), confirming adequate absorption and metabolism. Importantly, the intervention was safe and well-tolerated with no significant adverse effects.
\end{achievement}

This represents the first rigorous randomized controlled trial evidence that melatonin supplementation may provide symptomatic benefit in ME/CFS. However, several important caveats apply: the effect size was modest (improvement but not remission), the mechanism of benefit remains unclear (improved sleep, circadian resynchronization, anti-inflammatory effects, or antioxidant actions), and individual responses varied substantially (some patients benefited greatly, others not at all), and long-term efficacy and optimal dosing require further study.

\begin{warning}[Limitations of Melatonin Supplementation]
\label{warn:melatonin-limits}
While melatonin supplementation showed benefits in the Castro-Marrero trial, clinicians and patients should recognize important limitations. Melatonin primarily aids sleep initiation but does not address sleep maintenance (frequent awakenings), may temporarily improve symptoms without addressing underlying circadian dysfunction, risks masking underlying sleep disorders requiring different treatments (sleep apnea, restless legs syndrome), and exhibits substantial individual variation in absorption, metabolism, and response. Additionally, optimal timing of melatonin administration depends on the specific circadian phase abnormality (delayed, advanced, irregular), which typically requires formal assessment.
\end{warning}

\subsection{Circadian Disruption as an Integrative Mechanism}

The circadian rhythm abnormalities documented in ME/CFS should not be viewed as isolated sleep problems but rather as disruption of a master regulatory system that normally coordinates multi-system physiology. Loss of circadian organization contributes to HPA axis dysfunction (flattened cortisol rhythm discussed earlier in this chapter), immune dysfunction (loss of circadian immune regulation), metabolic dysfunction (disrupted glucose homeostasis and lipid metabolism), autonomic dysfunction (altered cardiovascular circadian patterns discussed in Chapter~\ref{ch:cardiovascular}), and thermoregulatory dysfunction (impaired circadian temperature variation).

This integrative perspective suggests that interventions targeting circadian resynchronization---whether through melatonin, light therapy, behavioral scheduling, or other chronotherapeutic approaches---might provide broader benefits than expected from improving sleep alone. By restoring temporal coordination across multiple systems, circadian interventions could theoretically address multiple aspects of ME/CFS pathophysiology simultaneously. However, this hypothesis requires rigorous testing in well-designed clinical trials.

\section{Integrated Endocrine-Metabolic Model}
\label{sec:endocrine-integration}

The endocrine abnormalities documented in this chapter do not represent independent, isolated dysfunctions but rather form an integrated network of disrupted hormonal regulation that mechanistically connects to the immune, neurological, metabolic, and cardiovascular dysfunction discussed in preceding chapters. Understanding these connections is essential for developing a coherent model of ME/CFS pathophysiology and identifying potential therapeutic targets.

\subsection{Neuroendocrine-Immune Integration}

The most critical integration involves bidirectional relationships between endocrine and immune systems. The HPA axis normally restrains immune activation through cortisol's anti-inflammatory effects, preventing excessive or prolonged inflammatory responses. In ME/CFS, blunted cortisol output and flattened circadian rhythm remove this restraint, permitting chronic low-grade inflammation to persist (Chapter~\ref{ch:immune-dysfunction}). Conversely, chronic immune activation through pro-inflammatory cytokines (IL-1$\beta$, IL-6, TNF-$\alpha$) suppresses HPA axis function by increasing central glucocorticoid feedback sensitivity and impairing adrenal steroidogenic enzyme function.

The sex-specific patterns documented by Walitt et al.\ (2024) and Heng et al.\ (2025)---with females showing stronger interferon-driven inflammation and males exhibiting more balanced immune profiles---directly reflect sex hormone influences on immune function~\cite{walitt2024deep,heng2025sexspecific}. Estrogen enhances type 2 interferon responses, while testosterone dampens multiple inflammatory pathways. The steroid hormone abnormalities documented by Pipper et al.\ (2024) thus contribute directly to the sex-specific immune dysregulation patterns~\cite{pipper2024steroid}.

These neuroendocrine-immune interactions create self-reinforcing pathological cycles. Inflammation disrupts HPA and HPG axis function, hormonal dysregulation permits unchecked inflammation, impaired cortisol rhythm disrupts circadian immune regulation, and circadian disruption further dysregulates neuroendocrine function. Breaking these cycles likely requires multi-targeted interventions addressing both endocrine and immune dysfunction simultaneously.

\subsection{Endocrine-Metabolic Connections}

The endocrine abnormalities documented in this chapter directly contribute to the cellular energy deficits discussed in Chapter~\ref{ch:energy-metabolism}. Cortisol supports hepatic gluconeogenesis and glucose availability; HPA dysfunction impairs this metabolic support. Thyroid hormone (T3) regulates mitochondrial biogenesis and oxidative phosphorylation efficiency; Low T3 Syndrome reduces cellular metabolic capacity. Insulin resistance and impaired glucose utilization limit substrate availability for ATP production; cerebral glucose hypometabolism reflects this at the brain level. Growth hormone supports protein synthesis and muscle metabolism; GH/IGF-1 deficiency contributes to muscle weakness and poor exercise tolerance.

The metabolic syndrome association documented by Maloney et al.\ (2010) indicates that ME/CFS involves not just cellular energy deficits but also systemic metabolic dysregulation affecting glucose homeostasis, lipid metabolism, and body composition~\cite{maloney2010metabolic}. This suggests that ME/CFS represents a form of ``metabolic failure'' spanning from mitochondrial dysfunction at the cellular level to whole-body insulin resistance and dysregulated energy partitioning.

\subsection{Circadian Disruption as Central Organizing Principle}

Circadian rhythm disruption may represent a central organizing principle connecting multiple aspects of ME/CFS pathophysiology. The suprachiasmatic nucleus (SCN) in the hypothalamus serves as the master circadian pacemaker, coordinating peripheral clocks throughout the body. Loss of this temporal coordination produces the constellation of abnormalities documented: flattened HPA axis cortisol rhythm, disrupted circadian immune function, altered metabolic switching between fed and fasted states, impaired cardiovascular circadian patterns (blood pressure, heart rate), and dysregulated body temperature variation.

The NIH study's documentation of temporal-parietal junction dysfunction and broader brain abnormalities suggests that the neurological impairments in ME/CFS may affect hypothalamic function, disrupting the SCN's ability to maintain circadian organization~\cite{walitt2024deep}. Inflammatory cytokines directly disrupt clock gene expression, creating a mechanistic link between immune activation and circadian dysfunction. The successful melatonin treatment trial by Castro-Marrero et al.\ (2021) supports the potential for circadian-targeted interventions~\cite{castromarrero2021melatonin}.

\subsection{Sex as a Critical Biological Variable}

The consistent 3--4:1 female-to-male prevalence ratio and the sex-specific pathophysiological patterns documented by multiple studies establish that biological sex is not merely a demographic variable but rather a critical determinant of disease susceptibility and expression. The mechanisms involve sex hormone modulation of immune responses (estrogen enhancing, testosterone dampening), sex-specific steroidogenic enzyme activity affecting stress hormone production, differential HPA and HPG axis regulation between sexes, and sex chromosome effects on immune gene expression (X chromosome contains numerous immune-related genes).

This recognition has profound implications for research and clinical care. Studies must stratify by sex to avoid obscuring sex-specific patterns, biomarker development should pursue sex-specific panels rather than assuming universal markers, and treatment trials should evaluate efficacy separately in male and female patients, as interventions effective in one sex may prove ineffective or even harmful in the other.

The early menopause finding by Boneva et al.\ (2015)---approximately 11 years earlier than controls---suggests that endocrine dysfunction may precede or contribute to ME/CFS onset rather than solely resulting from the disease~\cite{boneva2015menopause}. This raises the possibility that hormonal interventions (hormone replacement therapy in appropriate contexts, androgen supplementation for documented deficiency) might prevent or mitigate disease progression in susceptible individuals, though this hypothesis requires prospective testing.

\subsection{Clinical Implications and Therapeutic Considerations}

The integrated endocrine-metabolic dysfunction documented in this chapter provides both biomarker opportunities and therapeutic targets. However, several principles should guide clinical application:

First, single-system hormonal interventions have shown limited efficacy. Growth hormone treatment produced physiological effects but modest symptom improvement, thyroid hormone supplementation helps some patients but not others, and simple hormone replacement does not address underlying regulatory dysfunction. This pattern suggests that ME/CFS involves coordinated multi-system dysregulation rather than simple deficiency states amenable to replacement therapy.

Second, addressing upstream drivers (inflammation, oxidative stress, mitochondrial dysfunction) may prove more effective than downstream hormone replacement. If chronic inflammation suppresses multiple endocrine axes simultaneously, anti-inflammatory interventions might restore coordinated hormonal function more effectively than replacing individual hormones. The partial success of melatonin supplementation, which has anti-inflammatory and antioxidant effects beyond its chronobiotic actions, supports this multi-targeted approach.

Third, interventions must be individualized based on specific endocrine phenotypes. The heterogeneity documented across studies indicates that not all ME/CFS patients exhibit the same endocrine abnormalities. Some show pronounced HPA dysfunction, others demonstrate primarily thyroid or sex hormone abnormalities, and metabolic syndrome patterns characterize a distinct subset. Precision medicine approaches matching interventions to individual endocrine profiles may prove superior to universal treatment protocols.

Fourth, sex-specific treatment strategies warrant investigation. The sex-specific steroid hormone profiles and immune patterns suggest that males and females may require different therapeutic approaches. Interventions effective in predominantly female cohorts may not generalize to male patients, and vice versa.

\subsection{Future Research Directions}

Critical gaps in understanding endocrine dysfunction in ME/CFS include longitudinal studies tracking hormonal changes from disease onset through chronic phases, mechanistic studies elucidating causal relationships between immune activation and endocrine dysfunction, biomarker validation studies assessing whether hormonal measurements can predict disease severity or treatment response, intervention trials testing multi-targeted approaches addressing both endocrine and immune dysfunction, and sex-stratified research examining whether male and female ME/CFS represent partially distinct diseases requiring different treatments.

The endocrine system provides an attractive therapeutic target because hormones are measurable, hormone replacement therapies already exist for many deficiency states, and endocrine interventions have established safety profiles when properly monitored. However, the limited success of single-hormone interventions to date indicates that simplistic approaches will not suffice. Future therapeutic development must embrace the complexity of multi-system dysregulation, targeting coordinated restoration of neuroendocrine-immune-metabolic integration rather than isolated hormone replacement.
